---
title: "Study 03: Single-trait parameter estimation"
description: "Current five-replicate posterior recovery benchmark."
---

```{r}
#| echo: false
capsule <- "results/reference/03_parameter_estimation/current"
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"),
  simplifyVector = TRUE)
recovery <- read.csv(file.path(capsule, "parameter_recovery_summary.csv"),
  check.names = FALSE)
paired <- read.csv(file.path(capsule, "paired_comparison_summary.csv"),
  check.names = FALSE)
status <- read.csv(file.path(capsule, "replicate_status.csv"),
  check.names = FALSE)
```

::: {.callout-warning}
Five simulations per architecture make interval coverage coarse: each miss
changes observed coverage by 20 percentage points.
:::

## Design and controls

The current benchmark contains two architectures, five deterministic replicates,
and four methods for `r nrow(status)` total fits. It evaluates matched and
misspecified BayesC/SBayesC and BayesR/SBayesR priors using paired truth and
inputs. Unlike the earlier development snapshot, this fresh run deliberately
adopts the current Study 04 recommendations: 250 burn-in plus 250 retained draws
for BayesC/SBayesC and 250 burn-in plus 2,000 retained draws for BayesR/SBayesR,
with four chains and no thinning.

The exact package source is `r manifest$sblr_commit`. Truth, estimand
availability, interval definitions, and matched/misspecified classifications are
frozen in the capsule.

## Recovery

```{r}
#| echo: false
shown <- recovery[recovery$estimand_id %in%
  c("heritability", "genetic_variance", "residual_variance",
    "effect_variance") & recovery$successful_replicates > 0,
  c("architecture", "method", "model_specification", "estimand_id",
    "successful_replicates", "mean_truth", "mean_posterior_mean",
    "mean_bias", "rmse", "observed_coverage_proportion")]
knitr::kable(shown, digits = 4)
```

## Paired contrasts

```{r}
#| echo: false
knitr::kable(utils::head(paired, 24), digits = 4)
```

Unavailable estimands remain explicit in the full tables rather than being
dropped or assigned artificial truth values.

## Reproducibility

- [Current capsule](../../results/reference/03_parameter_estimation/current/README.md)
- [Manifest](../../results/reference/03_parameter_estimation/current/benchmark_manifest.json)
- [Parameter estimates](../../results/reference/03_parameter_estimation/current/parameter_estimates.csv)
- [Recovery summary](../../results/reference/03_parameter_estimation/current/parameter_recovery_summary.csv)
- [Checksums](../../results/reference/03_parameter_estimation/current/checksums.csv)

Earlier development snapshots remain available through Git history.
