---
title: "Study 03: Single-trait parameter estimation"
description: "A five-replicate benchmark of posterior parameter summaries under matched and misspecified prior classes."
---

::: {.callout-warning .callout-development}
## Development benchmark—not a final scientific comparison.
This report validates the parameter-estimation workflow and posterior-summary
contract. Five replicates give coarse observed recovery and coverage evidence,
not precise calibration or universal method performance.
:::

```{r}
historical_capsule <- file.path("results", "reference", "03_parameter_estimation",
  "st-parameter-estimation-one-replicate-development-v1")
five_capsule <- file.path("results", "reference", "03_parameter_estimation",
  "st-parameter-estimation-five-replicate-development-v1")
source("studies/reporting_helpers.R")
if (dir.exists(five_capsule)) {
  source(file.path("studies", "five_replicate_promotion.R"), local = TRUE)
  .five_validate_study03_capsule(five_capsule); capsule <- five_capsule
} else {
  source(file.path("studies", "03_parameter_estimation", "promotion.R"))
  .study03_validate_capsule(historical_capsule); capsule <- historical_capsule
}
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
estimates <- read.csv(file.path(capsule, "parameter_estimates.csv"))
registry <- read.csv(file.path(capsule, "estimand_registry.csv"))
paired <- read.csv(file.path(capsule, "paired_parameter_differences.csv"))
recovery_summary <- read.csv(file.path(capsule, "parameter_recovery_summary.csv"))
paired_summary <- read.csv(file.path(capsule, "paired_comparison_summary.csv"))
runtime <- read.csv(file.path(capsule, "computational_summary.csv"))
example_data <- read.csv(file.path(capsule,"example_data_manifest.csv"))
```

## Benchmark status

This is a **Complete five-replicate development benchmark**. Completion means all
expected fits and summaries exist. Observed interval coverage remains coarse:
each missed interval changes a five-replicate proportion by 20 percentage points.

The grid contains `r manifest$expected_fit_count` expected fits and
`r manifest$successful_fit_count` successful fits across two architectures and
four single-trait methods.

```{r}
#| echo: false
knitr::kable(recovery_summary[, intersect(c("architecture", "method", "model_specification",
  "estimand_id", "replicate_count", "successful_replicates", "mean_bias", "rmse", "mae",
  "observed_coverage_count", "observed_coverage_proportion", "mean_interval_width_95"),
  names(recovery_summary))], digits = 4,
  caption = "Aggregate recovery and coarse observed interval coverage.")
```

```{r}
#| echo: false
knitr::kable(paired_summary[, intersect(c("architecture", "comparison_id", "estimand_id",
  "replicate_count", "successful_replicates", "mean_estimate_difference",
  "sd_estimate_difference"), names(paired_summary))], digits = 4,
  caption = "Replicate-paired BED/CSR and prior-class differences.")
```

## Scientific question

Study 03 asks whether posterior summaries recover known finite-sample
quantities under approximately BayesC-like (`sparse_homogeneous`) and
BayesR-like (`sparse_mixture`) architectures. It is parameter estimation—not
held-out prediction ([Study 02](../02_prediction/prediction-development-pilot.qmd))
or multichain convergence ([Study 04](../04_convergence/convergence-development-pilot.qmd)).

## Methods compared

The benchmark uses ST-BED BayesC, ST-BED BayesR, ST-CSR SBayesC and ST-CSR
SBayesR in the canonical site-wide order.

## Data and simulation design

All 5,000 individuals and 37,991 post-QC markers are analyzed. Each simulation
has 50 causal markers and realized heritability 0.30. BED and CSR fits receive
the same phenotype and individuals; CSR summary statistics and sparse LD use
that same analysis sample. The four methods are ST-BED BayesC, ST-BED BayesR,
ST-CSR SBayesC, and ST-CSR SBayesR.

The homogeneous/BayesC and mixture/BayesR combinations are labelled `matched`;
cross-prior combinations are `misspecified` robustness checks. Whole-vector
heritability rescaling means the simulations are only approximately prior-like.

## Configuration

The primary five-replicate fits use the frozen Study 04 operational settings:
250 burn-in iterations, thinning 1, four identifiable chains and four cores.
BayesC and SBayesC retain 250 post-burn-in draws per chain; BayesR and SBayesR
retain 1,000. The original one-replicate capsule remains unchanged at 500
retained iterations, 250 burn-in iterations, one chain and one core; Study 04
recommendations were not adopted retroactively into that historical capsule.

```{r}
#| echo: false
knitr::kable(manifest$mcmc[, c(
  "method", "recommended_nburn", "recommended_post_burnin_draws",
  "recommended_nthin", "recommended_nchains", "recommended_ncores"
)], caption = "MCMC controls recorded in the primary five-replicate capsule.")
```

## Validation checks

All 40 architecture–replicate–method fits completed; simulation oracles passed;
BED and CSR analyses use the same 5,000 samples, phenotype and canonical markers.

## Results

### Generating and realized truths

```{r}
knitr::kable(registry[, c("estimand_id", "description", "truth_type",
  "truth_source", "posterior_source", "primary")])
```

The primary common estimands are effect variance, genetic variance, residual
variance, and heritability. Inclusion proportion and total marker-effect
variance are reported only where an aligned `pi_trace` exists; marker-level
BayesR component summaries are not treated as draw-level traces. Nonlinear
quantities are transformed draw by draw before posterior summarization.

### Parameter recovery

```{r}
main <- estimates[estimates$estimand_id %in% registry$estimand_id[registry$primary],
  c("architecture", "method", "model_specification", "estimand_id", "truth",
    "posterior_mean", "lower_95", "upper_95", "bias", "covered_95")]
main$method_label <- sblrbench_method_factor(main$method)
main$architecture_label <- sblrbench_architecture_factor(main$architecture)
main_display <- main[c("architecture_label","method_label","model_specification","estimand_id","truth",
  "posterior_mean","lower_95","upper_95","bias","covered_95")]
names(main_display) <- c("Architecture","Method","Model specification","Estimand","Truth",
  "Posterior mean","Lower 95%","Upper 95%","Bias","Covered")
knitr::kable(main_display, digits = 4)
```

```{r}
ggplot2::ggplot(main, ggplot2::aes(truth, posterior_mean, colour = method,shape=method)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2) +
  ggplot2::geom_point(size = 2) +
  ggplot2::facet_grid(estimand_id ~ architecture_label, scales = "free") +
  sblrbench_method_scales()+theme_sblrbench()+
  ggplot2::labs(x = "Simulation truth", y = "Posterior mean")
```

```{r}
ggplot2::ggplot(main, ggplot2::aes(method, bias, colour = method,shape=method)) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2) +
  ggplot2::geom_point() + ggplot2::coord_flip() +
  ggplot2::facet_grid(estimand_id ~ architecture_label, scales = "free_x")+
  sblrbench_method_scales()+theme_sblrbench()+ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())
```

```{r}
ggplot2::ggplot(main, ggplot2::aes(posterior_mean, method, colour = method,shape=method)) +
  ggplot2::geom_errorbarh(ggplot2::aes(xmin = lower_95, xmax = upper_95), height = .2) +
  ggplot2::geom_point() +
  ggplot2::facet_grid(estimand_id ~ architecture_label, scales = "free_x")+
  sblrbench_method_scales()+theme_sblrbench()+ggplot2::scale_y_discrete(limits=rev(sblrbench_method_levels()),labels=sblrbench_method_labels())
```

Each interval is an equal-tailed 95% posterior interval. `covered_95` is one
binary observation, not empirical coverage calibration.

### Matched versus misspecified models

Matched combinations are the homogeneous architecture with BayesC/SBayesC and
the variance-mixture architecture with BayesR/SBayesR. Cross-prior rows are
robustness observations, not direct recovery of the generating distribution.

### CSR versus BED paired comparisons

```{r}
knitr::kable(paired[paired$comparison_type == "BED-versus-CSR", ], digits = 4)
```

Differences use focal minus comparison with exact architecture, replicate and
estimand keys. Prior-class comparisons are retained separately and are not
presented as equivalent implementations.

## Computational performance

```{r}
knitr::kable(runtime[, c("architecture", "method", "runtime", "status",
  "nit", "nburn", "nchains", "posterior_draw_count")], digits = 2)
```

Runtimes are descriptive across five simulations. The primary capsule records
the Study 04 operational settings shown above; these remain development
recommendations rather than universal convergence guarantees.

## Interpretation

All 40 fits produced finite, aligned posterior summaries and the report contract
joins them to explicit generating or realized truths. Five-replicate recovery
differences and coarse coverage are descriptive only; they do not establish
unbiasedness or interval calibration. The five-replicate extension uses Study 04
operational recommendations, while the original one-replicate capsule retains
its historical settings unchanged.

## Limitations

Five replicates cannot establish unbiasedness or precise interval calibration.
Four-chain operational settings do not establish universal convergence. The
public example genotype resource has limited LD, and results do not support
general method rankings.
The separate [Study 04 convergence benchmark](../04_convergence/convergence-development-pilot.qmd)
provides the operational settings used by the primary five-replicate extension;
they were not retroactively applied to the historical one-replicate capsule.

## Code and reproduction

### Worked example

This frozen example demonstrates truth construction, one real fit, draw-wise
transformations, posterior summaries and recovery metrics. It is not executed
during rendering.

```{r, echo=FALSE, results='asis'}
display_capsule_script(file.path(historical_capsule,"worked_parameter_estimation_example.R"))
```

[Download the worked example](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/worked_parameter_estimation_example.R).

### Public example data

```{r, echo=FALSE}
knitr::kable(example_data[c("filename","role","size_bytes","md5")],
  col.names=c("File","Role","Bytes","MD5"))
```

[Download the example-data manifest](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/example_data_manifest.csv).

### Developer contract smoke test

The sampler-free contract test checks posterior extraction, draw-wise derived
quantities, truth construction and recovery metrics.

```{r, echo=FALSE, results='asis'}
display_capsule_script_collapsed(file.path(historical_capsule,"parameter_estimation_contract_smoke_test.R"))
```

[Download the smoke test](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/parameter_estimation_contract_smoke_test.R).

### Complete benchmark reproduction

The frozen targets entry point reproduces the benchmark and is displayed without execution.

```{r, echo=FALSE, results='asis'}
display_capsule_script(file.path(historical_capsule,"run_parameter_estimation_benchmark.R"))
```

[Download the reproduction script](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/run_parameter_estimation_benchmark.R).

## Downloads and provenance

| Resource | Download |
|---|---|
| Primary capsule | [README](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/README.md); [manifest](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/benchmark_manifest.json) |
| Primary results | [Parameter estimates](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/parameter_estimates.csv); [recovery summary](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/parameter_recovery_summary.csv); [paired differences](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/paired_parameter_differences.csv) |
| Historical reproduction material | [Configuration](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/config.R); [worked example](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/worked_parameter_estimation_example.R); [smoke test](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/parameter_estimation_contract_smoke_test.R); [run script](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/run_parameter_estimation_benchmark.R) |
| Primary provenance | [Example data](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/example_data_manifest.csv); [source inventory](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/source_files.csv); [session information](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/session_info.txt); [checksums](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/checksums.csv) |

The [five-replicate capsule](../../results/reference/03_parameter_estimation/st-parameter-estimation-five-replicate-development-v1/README.md)
is the primary result source. The [original one-replicate capsule](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/README.md)
remains an immutable historical reference.

source_tree_status <- if (
  is.null(manifest$source_tree_clean) ||
  length(manifest$source_tree_clean) == 0L
) {
  "Not recorded"
} else if (isTRUE(manifest$source_tree_clean)) {
  "Clean"
} else {
  "Dirty"
}

knitr::kable(data.frame(
  Item = c(
    "sblrbench commit",
    "sblr commit",
    "qgdata commit",
    "Source tree",
    "Profile"
  ),
  Value = c(
    manifest$repository_commit,
    manifest$sblr_commit,
    manifest$qgdata_commit,
    source_tree_status,
    "Development"
  )
))
