---
title: "Single-trait parameter-estimation development benchmark"
---

```{r}
capsule <- file.path("results", "reference", "03_parameter_estimation",
  "st-parameter-estimation-one-replicate-development-v1")
source(file.path("studies", "03_parameter_estimation", "promotion.R"))
.study03_validate_capsule(capsule)
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
estimates <- read.csv(file.path(capsule, "parameter_estimates.csv"))
registry <- read.csv(file.path(capsule, "estimand_registry.csv"))
paired <- read.csv(file.path(capsule, "paired_parameter_differences.csv"))
runtime <- read.csv(file.path(capsule, "computational_summary.csv"))
```

## Benchmark status

> **Complete one-replicate development benchmark.** This report validates the
> parameter-estimation workflow and posterior-summary contract. One replicate
> and short development MCMC do not estimate replicate-to-replicate variability,
> interval calibration, convergence, or final method performance.

The grid contains `r manifest$expected_fit_count` expected fits and
`r manifest$successful_fit_count` successful fits across two architectures and
four single-trait methods.

## Objective and scope

Study 03 asks whether posterior summaries recover known finite-sample
quantities under approximately BayesC-like (`sparse_homogeneous`) and
BayesR-like (`sparse_mixture`) architectures. It is parameter estimation—not
held-out prediction (Study 02) or multichain convergence (future Study 04).

## Data, simulation, and methods

All 5,000 individuals and 37,991 post-QC markers are analyzed. Each simulation
has 50 causal markers and realized heritability 0.30. BED and CSR fits receive
the same phenotype and individuals; CSR summary statistics and sparse LD use
that same analysis sample. The four methods are ST-BED BayesC, ST-BED BayesR,
ST-CSR SBayesC, and ST-CSR SBayesR.

The homogeneous/BayesC and mixture/BayesR combinations are labelled `matched`;
cross-prior combinations are `misspecified` robustness checks. Whole-vector
heritability rescaling means the simulations are only approximately prior-like.

## Generating and realized truths

```{r}
knitr::kable(registry[, c("estimand_id", "description", "truth_type",
  "truth_source", "posterior_source", "primary")])
```

The primary common estimands are effect variance, genetic variance, residual
variance, and heritability. Inclusion proportion and total marker-effect
variance are reported only where an aligned `pi_trace` exists; marker-level
BayesR component summaries are not treated as draw-level traces. Nonlinear
quantities are transformed draw by draw before posterior summarization.

## Parameter recovery

```{r}
main <- estimates[estimates$estimand_id %in% registry$estimand_id[registry$primary],
  c("architecture", "method", "model_specification", "estimand_id", "truth",
    "posterior_mean", "lower_95", "upper_95", "bias", "covered_95")]
knitr::kable(main, digits = 4)
```

```{r}
ggplot2::ggplot(main, ggplot2::aes(truth, posterior_mean, colour = method)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2) +
  ggplot2::geom_point(size = 2) +
  ggplot2::facet_grid(estimand_id ~ architecture, scales = "free") +
  ggplot2::labs(x = "Simulation truth", y = "Posterior mean")
```

```{r}
ggplot2::ggplot(main, ggplot2::aes(method, bias, colour = model_specification)) +
  ggplot2::geom_hline(yintercept = 0, linetype = 2) +
  ggplot2::geom_point() + ggplot2::coord_flip() +
  ggplot2::facet_grid(estimand_id ~ architecture, scales = "free_x")
```

```{r}
ggplot2::ggplot(main, ggplot2::aes(posterior_mean, method, colour = model_specification)) +
  ggplot2::geom_errorbarh(ggplot2::aes(xmin = lower_95, xmax = upper_95), height = .2) +
  ggplot2::geom_point() +
  ggplot2::facet_grid(estimand_id ~ architecture, scales = "free_x")
```

Each interval is an equal-tailed 95% posterior interval. `covered_95` is one
binary observation, not empirical coverage calibration.

## BED-versus-CSR paired comparisons

```{r}
knitr::kable(paired[paired$comparison_type == "BED-versus-CSR", ], digits = 4)
```

Differences use focal minus comparison with exact architecture, replicate and
estimand keys. Prior-class comparisons are retained separately and are not
presented as equivalent implementations.

## Computational summary

```{r}
knitr::kable(runtime[, c("architecture", "method", "runtime", "status",
  "nit", "nburn", "nchains", "posterior_draw_count")], digits = 2)
```

Runtimes are descriptive for this single run. The development settings use 500
iterations after 250 burn-in iterations, one chain and one core; they are not
convergence recommendations.

## Limitations

One replicate cannot establish unbiasedness or interval calibration. Short
single-chain MCMC does not establish convergence. The public example genotype
resource has limited LD, and results do not support general method rankings.
Study 04 will address multichain diagnostics and MCMC-length selection.

## Reproduction and downloads

The [capsule README](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/README.md),
[configuration](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/config.R),
[worked example](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/worked_parameter_estimation_example.R),
[run script](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/run_parameter_estimation_benchmark.R),
[parameter estimates](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/parameter_estimates.csv),
[source inventory](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/source_files.csv), and
[checksums](../../results/reference/03_parameter_estimation/st-parameter-estimation-one-replicate-development-v1/checksums.csv)
form the tracked reproducibility capsule. Genotypes, native fits, individual
phenotypes, and posterior draw tables are not published.
