---
title: "Study 05: Annotation-model convergence decision"
description: "Current maximum-history BayesRC and SBayesRC convergence stop."
---

```{r}
#| echo: false
capsule <- "results/reference/05_annotation_models/current-stop"
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"),
  simplifyVector = TRUE)
recommendations <- read.csv(file.path(capsule, "method_recommendations.csv"),
  check.names = FALSE)
status <- read.csv(file.path(capsule, "fit_status.csv"), check.names = FALSE)
design <- read.csv(file.path(capsule, "annotation_design_summary.csv"),
  check.names = FALSE)
```

::: {.callout-important}
Study 05 is stopped at its prespecified convergence gate. This is not a
completed annotation-model scientific benchmark, and no five-replicate result
capsule exists.
:::

## Completed work

The public BayesRC and SBayesRC interfaces, annotation orientation, probit
stick-breaking probabilities, marker alignment, and chain-output contracts were
validated. A single deterministic annotation matrix with intercept,
`enriched_binary`, `continuous_signal`, and `null_annotation` was used.

Both maximum-history annotation fits completed with four identifiable chains
under `sblr` 0.2.0 at `r manifest$installed_sblr_commit`. The full benchmark flag is
`r manifest$full_benchmark_started`; no simulation grid beyond the pilot was
started.

## Prespecified stop decision

```{r}
#| echo: false
knitr::kable(recommendations, digits = 4)
```

Neither annotation-informed method passed every unchanged Study 04 threshold at
any candidate window. The limiting quantity for both was the R-hat of the
`continuous_signal` coefficient for the component-0 stick. The largest assessed
window used 1,000 burn-in and 2,000 retained draws. This failure is preserved as
scientific evidence; thresholds were not weakened and seeds were not replaced.

## Reproducibility

- [Current stop capsule](../../results/reference/05_annotation_models/current-stop/README.md)
- [Manifest](../../results/reference/05_annotation_models/current-stop/benchmark_manifest.json)
- [Convergence diagnostics](../../results/reference/05_annotation_models/current-stop/convergence_diagnostics.csv)
- [Candidate settings](../../results/reference/05_annotation_models/current-stop/candidate_settings.csv)
- [Development run record](../../docs/dev/study05_annotation_models_run.md)
- [Checksums](../../results/reference/05_annotation_models/current-stop/checksums.csv)

A later annotation benchmark requires a new, supported convergence design.
