---
title: "Study 05: Single-trait annotation-informed models"
description: "A five-replicate development benchmark isolating annotation-informed mixture probabilities."
---

::: {.callout-warning .callout-development}
## Development benchmark—not a final method ranking

This package-specific benchmark reports five paired simulations under
informative and uninformative annotations. The evidence is descriptive, is not
universal convergence validation, and describes the current installed `sblr`
behaviour.
:::

```{r}
convergence_capsule <- file.path("results", "reference",
  "05_annotation_models",
  "st-annotation-models-convergence-development-v1")
benchmark_capsule <- file.path("results", "reference",
  "05_annotation_models",
  "st-annotation-models-five-replicate-development-v1")
source("studies/reporting_helpers.R")
source(file.path("studies", "05_annotation_models", "promotion.R"))
.study05_validate_convergence_capsule(convergence_capsule)
.study05_validate_benchmark_capsule(benchmark_capsule)
manifest <- jsonlite::read_json(file.path(benchmark_capsule,
  "benchmark_manifest.json"), simplifyVector = TRUE)
pilot_manifest <- jsonlite::read_json(file.path(convergence_capsule,
  "benchmark_manifest.json"), simplifyVector = TRUE)
prediction <- read.csv(file.path(benchmark_capsule, "prediction_metrics.csv"))
prediction_summary <- read.csv(file.path(benchmark_capsule, "prediction_summary.csv"))
parameters <- read.csv(file.path(benchmark_capsule, "parameter_estimates.csv"))
parameter_summary <- read.csv(file.path(benchmark_capsule, "parameter_recovery_summary.csv"))
alpha <- read.csv(file.path(benchmark_capsule, "annotation_coefficient_estimates.csv"))
alpha_summary <- read.csv(file.path(benchmark_capsule, "annotation_recovery_summary.csv"))
probability <- read.csv(file.path(benchmark_capsule, "probability_recovery_metrics.csv"))
probability_summary <- read.csv(file.path(benchmark_capsule, "probability_recovery_summary.csv"))
enrichment <- read.csv(file.path(benchmark_capsule, "enrichment_metrics.csv"))
enrichment_summary <- read.csv(file.path(benchmark_capsule, "enrichment_summary.csv"))
paired <- read.csv(file.path(benchmark_capsule, "paired_replicate_differences.csv"))
paired_summary <- read.csv(file.path(benchmark_capsule, "paired_comparison_summary.csv"))
interactions <- read.csv(file.path(benchmark_capsule, "annotation_value_interactions.csv"))
interaction_summary <- read.csv(file.path(benchmark_capsule,
  "annotation_value_interaction_summary.csv"))
diagnostics <- read.csv(file.path(benchmark_capsule, "convergence_diagnostics.csv"))
validation <- read.csv(file.path(benchmark_capsule, "convergence_validation_summary.csv"))
runtime <- read.csv(file.path(benchmark_capsule, "computational_summary.csv"))
simulation <- read.csv(file.path(benchmark_capsule, "simulation_summary.csv"))
annotation_summary <- read.csv(file.path(benchmark_capsule,
  "annotation_design_summary.csv"))
true_alpha <- read.csv(file.path(benchmark_capsule, "true_alpha.csv"),
  check.names = FALSE)
recommendations <- read.csv(file.path(convergence_capsule,
  "method_recommendations.csv"))
availability <- read.csv(file.path(benchmark_capsule,
  "method_output_availability.csv"))
method_labels <- c(
  st_bed_bayesr = "ST-BED BayesR",
  st_bed_bayesrc = "ST-BED BayesRC",
  st_csr_sbayesr = "ST-CSR SBayesR",
  st_csr_sbayesrc = "ST-CSR SBayesRC")
scenario_labels <- c(
  informative_annotations = "Informative annotations",
  uninformative_annotations = "Uninformative annotations")
for (x in c("prediction", "prediction_summary", "parameters",
            "parameter_summary", "alpha", "alpha_summary", "probability",
            "probability_summary", "enrichment", "enrichment_summary",
            "diagnostics", "validation", "runtime")) {
  object <- get(x)
  if ("method" %in% names(object))
    object$method_label <- unname(method_labels[object$method])
  if ("scenario" %in% names(object))
    object$scenario_label <- unname(scenario_labels[object$scenario])
  assign(x, object)
}
```

## Scientific question

Study 05 asks whether marker annotations improve mixture allocation and
held-out prediction when annotations truly identify likely non-null or larger
effects, while avoiding an apparent advantage when the same annotations are
uninformative. The comparison isolates annotation-informed mixture
probabilities by pairing BayesRC with BayesR in both BED and CSR interfaces.
Block-eigen models are reserved for Study 06.

## Methods and benchmark specification

The four methods are ST-BED BayesR, ST-BED BayesRC, ST-CSR SBayesR and
ST-CSR SBayesRC. Every method uses the Study 05 explicit four-component
variance grid:

```{r}
manifest$mixture_var
```

This grid is a benchmark specification, not a model-specific default.

## Annotation design and exact generating model

One aligned marker-level matrix is reused for both scenarios and all five
replicates. It contains an explicit intercept, a deterministic 10% binary
enrichment class, a standardized continuous signal, and an independent
standardized null annotation.

```{r}
knitr::kable(annotation_summary)
```

The verified installed probit stick-breaking orientation is documented in the
[development run record](../../docs/dev/study05_annotation_models_run.md).
The frozen coefficients are:

```{r}
knitr::kable(true_alpha)
```

The informative scenario uses nonzero enriched and continuous coefficients and
a zero null-annotation coefficient. The uninformative scenario sets all
non-intercept coefficients to zero and matches the informative
marker-averaged component occupancy using intercepts calibrated once before
simulation.

## Data and held-out design

The benchmark uses 5,000 cached qgdata samples and the canonical post-QC
chromosome-1 marker set. The deterministic 70/30 split is shared by every
method in a scenario-replicate pair. Scaling, allele frequencies, summary
statistics and CSR LD use training individuals only. Test phenotypes remain
held out.

```{r}
knitr::kable(simulation)
```

## Convergence-pilot recommendations

BayesR and SBayesR settings come from Study 04's frozen recommendation capsule.
BayesRC and SBayesRC settings were selected from four-chain, 3,000-draw maximum
histories on the first informative simulation. Candidate burn-in was removed
exactly once.

```{r}
knitr::kable(recommendations)
```

## Five-replicate completion status

All `r manifest$expected_fit_count` planned multichain fits completed only if
this frozen capsule passed its grid validator: two scenarios by five paired
replicates by four methods, with four identifiable chains per fit.

## Prediction results

```{r}
pred_plot <- prediction[prediction$metric %in% c(
  "phenotype_prediction_correlation", "phenotype_prediction_rmse",
  "genetic_value_correlation"), ]
ggplot2::ggplot(pred_plot, ggplot2::aes(method_label, value,
  colour = method_label)) +
  ggplot2::geom_point(position = ggplot2::position_jitter(width = .08),
    alpha = .8) +
  ggplot2::facet_grid(metric ~ scenario_label, scales = "free_y") +
  ggplot2::labs(x = NULL, y = NULL, colour = "Method") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1)) +
  theme_sblrbench()
```

```{r}
knitr::kable(prediction_summary)
```

Each point is one paired simulation. No uncertainty has been fabricated.

## Heritability and variance recovery

```{r}
knitr::kable(parameter_summary)
```

Observed interval coverage is deliberately labelled coarse: with five
replicates, each miss changes the proportion by 20 percentage points. The
known SBayesR heritability behaviour is benchmarked as currently installed and
is not diagnosed or corrected here. Unavailable estimands remain in the
downloadable tables with `available = FALSE` and a reason.

## Annotation coefficient and variance behaviour

```{r}
alpha_nonintercept <- alpha[alpha$annotation_name != "Intercept", ]
ggplot2::ggplot(alpha_nonintercept,
  ggplot2::aes(true_alpha, posterior_mean, colour = method_label)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2) +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = lower_95, ymax = upper_95),
    width = 0) +
  ggplot2::geom_point() +
  ggplot2::facet_grid(stick_name ~ scenario_label) +
  ggplot2::labs(x = "True alpha", y = "Posterior alpha", colour = "Method") +
  theme_sblrbench()
```

`sigmaSqAlpha` has no comparable generating truth in this simulation and is
therefore reported as posterior behaviour, not parameter recovery.

## Marker probability and enrichment recovery

```{r}
knitr::kable(probability_summary)
```

Annotation-implied prior probabilities, posterior component-allocation
probabilities, global component proportions, and posterior marker non-null
probabilities are separate quantities throughout the outputs.

```{r}
enrich_plot <- enrichment[enrichment$metric %in% c(
  "enrichment_difference", "enrichment_ratio",
  "continuous_annotation_direction"), ]
ggplot2::ggplot(enrich_plot, ggplot2::aes(method_label, value,
  colour = method_label)) +
  ggplot2::geom_point(position = ggplot2::position_jitter(width = .08)) +
  ggplot2::facet_grid(metric ~ scenario_label, scales = "free_y") +
  ggplot2::labs(x = NULL, y = NULL, colour = "Method") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1)) +
  theme_sblrbench()
```

## Paired RC-minus-R comparisons

```{r}
knitr::kable(paired_summary[paired_summary$comparison_id %in%
  c("bed_rc_minus_r", "csr_rc_minus_r"), ])
```

Positive differences always mean focal RC minus its paired R baseline; metric
orientation remains explicit, so lower-is-better RMSE differences are
interpreted accordingly.

## Informative-minus-uninformative interaction

```{r}
knitr::kable(interaction_summary)
```

This is the descriptive paired interaction
`(RC - R | informative) - (RC - R | uninformative)`. Five pairs do not provide
definitive inferential evidence.

## BED-versus-CSR comparisons

```{r}
knitr::kable(paired_summary[paired_summary$comparison_id %in%
  c("csr_rc_minus_bed_rc", "csr_r_minus_bed_r"), ])
```

## Main-run convergence validation

```{r}
knitr::kable(validation)
```

The status summarizes the same applicable rank-normalized split R-hat, bulk
ESS, tail ESS, MCSE and relative-MCSE thresholds used for recommendation
selection. A failed estimand is not hidden inside the method-level aggregate.
Support across five simulations is not universal convergence validation.

## Runtime

```{r}
runtime_summary <- aggregate(cbind(model_fitting_time, total_time) ~
  scenario_label + method_label, runtime, function(x)
    c(mean = mean(x), sd = stats::sd(x), min = min(x), max = max(x)))
knitr::kable(runtime_summary)
```

Preprocessing, summary-statistic construction, LD, fitting, prediction and
metric generation remain separate in the downloadable computational table.

## Limitations

This is a development-scale, package-specific benchmark using five paired
simulations and one qgdata chromosome. It is not a definitive ranking, a claim
about other software implementations, or universal convergence validation.
The annotation design is deliberately simple, and global quantities absent
from the current returned chain contract are not reconstructed from proxies.

## Downloads and provenance

- [Convergence capsule README](../../results/reference/05_annotation_models/st-annotation-models-convergence-development-v1/README.md)
- [Five-replicate capsule README](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/README.md)
- [Benchmark manifest](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/benchmark_manifest.json)
- [Prediction metrics](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/prediction_metrics.csv)
- [Parameter estimates](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/parameter_estimates.csv)
- [Annotation coefficient estimates](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/annotation_coefficient_estimates.csv)
- [Probability recovery](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/probability_recovery_metrics.csv)
- [Paired differences](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/paired_replicate_differences.csv)
- [Checksums](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/checksums.csv)
- [Reproduction runner](../../results/reference/05_annotation_models/st-annotation-models-five-replicate-development-v1/reproduce.R)

The report reads only these frozen capsules; rendering never constructs LD,
simulates data, or calls a sampler.
