---
title: "Study 04: Multichain convergence"
description: "Current recommendation selection and five-replicate validation."
---

```{r}
#| echo: false
selection <- "results/reference/04_convergence/current-selection"
validation <- "results/reference/04_convergence/current-validation"
selection_manifest <- jsonlite::read_json(file.path(selection,
  "benchmark_manifest.json"), simplifyVector = TRUE)
recommendations <- read.csv(file.path(selection, "method_recommendations.csv"),
  check.names = FALSE)
validation_summary <- read.csv(file.path(validation,
  "method_validation_summary.csv"), check.names = FALSE)
selection_status <- read.csv(file.path(selection, "chain_status.csv"),
  check.names = FALSE)
validation_status <- read.csv(file.path(validation, "chain_status.csv"),
  check.names = FALSE)
```

::: {.callout-warning}
Operational support in five simulations is not universal convergence
validation. Recommendations apply to the tested package, data, priors, and
diagnostic registry.
:::

## Selection and validation stages

The recommendation-generating stage used four maximum-history fits—one matched
method for each architecture/representation combination—and 16 identifiable
chains. Candidate burn-in and retained-draw windows were assessed with the
unchanged R-hat, bulk ESS, tail ESS, relative-MCSE, and burn-in-stability rules.

The separate fixed-setting stage then used 20 fits and 80 chains: five paired
replicates for each recommendation. All four recommendations were supported in
all five validation replicates. Both stages used `sblr` 0.2.0 at
`r selection_manifest$sblr_commit`.

## Current recommendations

```{r}
#| echo: false
knitr::kable(recommendations[, c("method", "architecture",
  "recommendation_status", "recommended_nchains", "recommended_nburn",
  "recommended_post_burnin_draws", "recommended_nthin",
  "limiting_estimand", "limiting_diagnostic")], digits = 4)
```

## Five-replicate support

```{r}
#| echo: false
knitr::kable(validation_summary[, c("architecture", "method",
  "replicate_count", "replicates_passing_all_core_estimands",
  "maximum_rhat", "minimum_bulk_ess", "minimum_tail_ess",
  "maximum_relative_mcse", "recommendation_validation_status")], digits = 4)
```

## Reproducibility

- [Selection capsule](../../results/reference/04_convergence/current-selection/README.md)
- [Validation capsule](../../results/reference/04_convergence/current-validation/README.md)
- [Recommendations](../../results/reference/04_convergence/current-selection/method_recommendations.csv)
- [Validation summary](../../results/reference/04_convergence/current-validation/method_validation_summary.csv)
- [Selection checksums](../../results/reference/04_convergence/current-selection/checksums.csv)
- [Validation checksums](../../results/reference/04_convergence/current-validation/checksums.csv)

Earlier recommendation snapshots remain available through Git history.
