---
title: "Study 04: Multichain convergence"
description: "A frozen matched-architecture benchmark of four identifiable chains and operational scalar convergence diagnostics."
---

::: {.callout-warning .callout-development}
## Development benchmark—not a final scientific comparison.
Operational multichain diagnostics are evaluated on one simulated dataset per
matched architecture. Passing thresholds supports provisional settings for
subsequent development work, not universal convergence.
:::

```{r}
capsule <- file.path("results", "reference", "04_convergence",
  "st-multichain-convergence-development-v1")
source("studies/reporting_helpers.R")
source(file.path("studies", "04_convergence", "promotion.R"))
.study04_validate_capsule(capsule)
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
draws <- read.csv(file.path(capsule, "scalar_chain_draws.csv"))
diagnostics <- read.csv(file.path(capsule, "convergence_diagnostics.csv"))
stability <- read.csv(file.path(capsule, "burnin_stability.csv"))
recommendations <- read.csv(file.path(capsule, "method_recommendations.csv"))
runtime <- read.csv(file.path(capsule, "computational_summary.csv"))
agreement <- read.csv(file.path(capsule, "chain_agreement.csv"))
thresholds <- read.csv(file.path(capsule, "diagnostic_thresholds.csv"))
example_data <- read.csv(file.path(capsule,"example_data_manifest.csv"))
method_labels <- sblrbench_method_labels()
draws$method_label <- unname(method_labels[draws$method])
diagnostics$method_label <- unname(method_labels[diagnostics$method])
```

## Benchmark status

This is a complete one-replicate matched-architecture development benchmark.
Benchmark completion means all sixteen planned chains and diagnostic products
exist; it does not mean every diagnostic must pass.

All `r manifest$expected_chain_count` planned chains completed: four identifiable
chains for each of four matched method–architecture combinations. Diagnostic
failure means the evaluated length is insufficient or inconclusive—not that the
statistical model is invalid.

## Scientific question

Study 04 asks whether principal scalar chains appear to sample a common stable
posterior and how much warm-up and retained history met prespecified development
criteria. Simulation truth is not used in pass/fail rules. The study makes no
parameter-recovery, predictive-performance, convergence-proof, or method-ranking
claim.

## Methods compared

The matched grid contains ST-BED BayesC, ST-BED BayesR, ST-CSR SBayesC and
ST-CSR SBayesR. Method display order and visual identities follow the common
site contract even though only matched architecture–prior combinations run.

## Data and simulation design

The verified native `sblr` interface ran four independent chains with explicit
seeds, `nchains = 4`, `ncores = 4`, `keep_chains = TRUE`, and retained convergence
traces. `nit` is post-warm-up draws and `nburn` is additional warm-up. The maximum
run used `nit = 3000`, `nburn = 0`, and `nthin = 1`, preserving all 3,000 draws;
candidate burn-in was applied downstream exactly once.

The homogeneous architecture is paired with BED/CSR BayesC, and the sparse
mixture architecture with BED/CSR BayesR. Each simulation uses 5,000 samples,
37,991 canonical markers, 50 causal markers, and realized heritability 0.30.

## Configuration

The maximum run uses four chains, four native chain workers, 3,000 raw draws,
zero sampler burn-in and no thinning. Candidate burn-in is applied downstream.

## Validation checks

All sixteen planned chains completed with unique recorded seeds, finite scalar
draws, valid domains and unambiguous chain identities.

## Results

### Diagnostics and operational thresholds

Core quantities are effect variance, genetic variance, residual variance, and
draw-wise heritability. Aligned inclusion-proportion traces are unavailable in
the retained common contract, so no proxy was substituted.

```{r}
knitr::kable(thresholds)
```

Diagnostics are rank-normalized split R-hat, bulk ESS, tail ESS, MCSE of the
mean, and relative MCSE. Burn-in candidates are 250, 500, and 1,000 draws;
retained checkpoints are 250, 500, 1,000, and 2,000 draws per chain.

### Trace and running-mean views

```{r}
trace_data <- draws[draws$raw_iteration %% 10L == 0L, ]
trace_long <- reshape(trace_data,
  varying=c("effect_variance","genetic_variance","residual_variance","heritability"),
  v.names="value",timevar="estimand",times=c("effect_variance","genetic_variance","residual_variance","heritability"),
  direction="long")
ggplot2::ggplot(trace_long, ggplot2::aes(raw_iteration,value,colour=factor(chain))) +
  ggplot2::geom_line(alpha=.65,linewidth=.25) +
  ggplot2::facet_grid(estimand ~ method_label, scales="free_y") +
  ggplot2::labs(colour="Chain",x="Raw iteration",y=NULL)+theme_sblrbench()
```

```{r}
running <- do.call(rbind,lapply(split(draws,interaction(draws$architecture,draws$method,draws$chain,drop=TRUE)),function(x){
  x$running_mean <- cumsum(x$heritability)/seq_len(nrow(x)); x[x$raw_iteration%%10L==0L,]
}))
ggplot2::ggplot(running,ggplot2::aes(raw_iteration,running_mean,colour=factor(chain)))+
  ggplot2::geom_line(linewidth=.3)+ggplot2::facet_wrap(~method_label,scales="free_y")+
  ggplot2::labs(colour="Chain",x="Raw iteration",y="Running mean heritability")+theme_sblrbench()
```

### Diagnostic trajectories

```{r}
ggplot2::ggplot(diagnostics,ggplot2::aes(retained_draw_candidate,rhat,colour=estimand))+
  ggplot2::geom_hline(yintercept=1.01,linetype=2)+ggplot2::geom_line()+ggplot2::geom_point()+
  ggplot2::facet_grid(burnin_candidate~method_label)+ggplot2::labs(x="Retained draws/chain",y="Rank-normalized split R-hat")+theme_sblrbench()
```

```{r}
ess_long <- reshape(diagnostics,varying=c("ess_bulk","ess_tail"),v.names="ess",timevar="type",
  times=c("bulk","tail"),direction="long")
ggplot2::ggplot(ess_long,ggplot2::aes(retained_draw_candidate,ess,colour=estimand))+
  ggplot2::geom_hline(yintercept=400,linetype=2)+ggplot2::geom_line()+ggplot2::geom_point()+
  ggplot2::facet_grid(type~method_label)+ggplot2::labs(x="Retained draws/chain",y="Effective sample size")+theme_sblrbench()
```

```{r}
ggplot2::ggplot(diagnostics,ggplot2::aes(retained_draw_candidate,relative_mcse,colour=estimand))+
  ggplot2::geom_hline(yintercept=.05,linetype=2)+ggplot2::geom_line()+ggplot2::geom_point()+
  ggplot2::facet_grid(burnin_candidate~method_label)+ggplot2::labs(x="Retained draws/chain",y="Relative MCSE")+theme_sblrbench()
```

### Burn-in stability and recommendations

```{r}
stability_display <- stability[,c("architecture","method","burnin_candidate","estimand","standardized_mean_shift","stable")]
stability_display$architecture <- sblrbench_architecture_factor(stability_display$architecture)
stability_display$method <- sblrbench_method_factor(stability_display$method)
names(stability_display) <- c("Architecture","Method","Burn-in","Estimand","Standardized mean shift","Stable")
knitr::kable(stability_display,digits=4)
```

```{r}
recommendations_display <- recommendations
recommendations_display$architecture <- sblrbench_architecture_factor(recommendations_display$architecture)
recommendations_display$method <- sblrbench_method_factor(recommendations_display$method)
knitr::kable(recommendations_display,digits=4)
```

All recommendations use four chains and no thinning. A checkpoint qualifies
only if every core estimand passes there and at every later checkpoint. The
reported limiting quantity is descriptive for this one simulated dataset.

### CSR versus BED and marker agreement

BED and CSR are shown within their matched prior class only. Scalar convergence
diagnostics—not high-dimensional agreement—determine recommendations.
Chain-specific marker summaries were retained, so pairwise effect/PIP agreement
is available as secondary context.

```{r}
agreement_display <- agreement
agreement_display$architecture <- sblrbench_architecture_factor(agreement_display$architecture)
agreement_display$method <- sblrbench_method_factor(agreement_display$method)
knitr::kable(agreement_display,digits=4)
```

## Computational performance

```{r}
method_runtime <- unique(runtime[c("architecture","method","method_total_runtime")])
method_runtime$architecture <- sblrbench_architecture_factor(method_runtime$architecture)
method_runtime$method <- sblrbench_method_factor(method_runtime$method)
names(method_runtime) <- c("Architecture","Method","Total runtime (s)")
knitr::kable(method_runtime,digits=1)
```

Runtime is descriptive. Chain-level runtime was not exposed by native
orchestration, so only method totals are reported; per-chain runtime remains NA.

## Interpretation

All four matched workflows retained four identifiable chains and produced
complete operational diagnostics. The provisional recommendations distinguish
the shorter BayesC checkpoints from the longer BayesR checkpoints observed in
these simulations. They are not universal settings and do not assess model
accuracy, prediction, or scientific superiority.

## Limitations

There is one simulated dataset per architecture, one prior configuration, and a
public example genotype resource with limited LD. Operational thresholds are not
proof of convergence. No thinning was used to hide autocorrelation. Initial-state
controls were unavailable, so chains used identical priors and independent seeds.
The study does not update Studies 02 or 03 automatically.

## Code and reproduction

## Worked example

The complete worked example is displayed below and is deliberately not executed
during site rendering.

```{r, echo=FALSE, results='asis'}
display_capsule_script(file.path(capsule,"worked_convergence_example.R"))
```

[Download the worked example](../../results/reference/04_convergence/st-multichain-convergence-development-v1/worked_convergence_example.R).

## Public example data

```{r, echo=FALSE}
knitr::kable(example_data[c("filename","role","size_bytes","md5")],
  col.names=c("File","Role","Bytes","MD5"))
```

[Download the example-data manifest](../../results/reference/04_convergence/st-multichain-convergence-development-v1/example_data_manifest.csv).

## Developer contract smoke test

The tiny native probe verifies chain retention, seed identity and trace shape;
it is not benchmark evidence.

```{r, echo=FALSE, results='asis'}
display_capsule_script_collapsed(file.path(capsule,"convergence_contract_smoke_test.R"))
```

[Download the smoke test](../../results/reference/04_convergence/st-multichain-convergence-development-v1/convergence_contract_smoke_test.R).

## Complete benchmark reproduction

The frozen targets entry point reproduces the full matched benchmark and is not executed during rendering.

```{r, echo=FALSE, results='asis'}
display_capsule_script(file.path(capsule,"run_convergence_benchmark.R"))
```

[Download the reproduction script](../../results/reference/04_convergence/st-multichain-convergence-development-v1/run_convergence_benchmark.R).

## Downloads and provenance

| Resource | Download |
|---|---|
| Capsule | [README](../../results/reference/04_convergence/st-multichain-convergence-development-v1/README.md); [manifest](../../results/reference/04_convergence/st-multichain-convergence-development-v1/benchmark_manifest.json) |
| Primary results | [Raw scalar chains](../../results/reference/04_convergence/st-multichain-convergence-development-v1/scalar_chain_draws.csv); [diagnostics](../../results/reference/04_convergence/st-multichain-convergence-development-v1/convergence_diagnostics.csv); [recommendations](../../results/reference/04_convergence/st-multichain-convergence-development-v1/method_recommendations.csv) |
| Reproduction | [Configuration](../../results/reference/04_convergence/st-multichain-convergence-development-v1/config.R); [worked example](../../results/reference/04_convergence/st-multichain-convergence-development-v1/worked_convergence_example.R); [smoke test](../../results/reference/04_convergence/st-multichain-convergence-development-v1/convergence_contract_smoke_test.R); [run script](../../results/reference/04_convergence/st-multichain-convergence-development-v1/run_convergence_benchmark.R) |
| Provenance | [Example data](../../results/reference/04_convergence/st-multichain-convergence-development-v1/example_data_manifest.csv); [source inventory](../../results/reference/04_convergence/st-multichain-convergence-development-v1/source_files.csv); [session information](../../results/reference/04_convergence/st-multichain-convergence-development-v1/session_info.txt); [checksums](../../results/reference/04_convergence/st-multichain-convergence-development-v1/checksums.csv) |

```{r, echo=FALSE}
knitr::kable(data.frame(Item=c("sblrbench commit","sblr commit","qgdata commit","Source tree","Profile"),
  Value=c(manifest$repository_commit,manifest$sblr_commit,manifest$qgdata_commit,
    if(manifest$source_tree_clean) "Clean" else "Dirty","Development")))
```
