---
title: "sblrbench"
---

```{r}
#| echo: false
capsule <- "results/reference/01_finemapping/separated-development-v1.1"
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
status <- read.csv(file.path(capsule, "replicate_status.csv"))
summary <- read.csv(file.path(capsule, "benchmark_summary.csv"))
```

`sblrbench` provides reusable contracts, validation, metrics, provenance and reproducible benchmark studies for scalable Bayesian linear regression methods implemented in `sblr`.

## Latest reference benchmark

### [Separated-locus fine-mapping development pilot](studies/01_finemapping/separated-development-pilot.qmd)

```{r}
#| echo: false
facts <- data.frame(
  Item = c("Simulation replicates", "Successful fits", "Methods", "Canonical markers", "Causal markers per replicate", "Status"),
  Value = c(manifest$replicate_count, paste0(sum(status$status == "ok"), "/", nrow(status)),
    length(manifest$methods), format(manifest$canonical_marker_count, big.mark = ","),
    manifest$simulation$n_causal, if (manifest$development_settings) "Development settings" else "Scientific settings")
)
knitr::kable(facts)
```

> Results currently demonstrate pipeline correctness and software consistency. They are not final method-performance claims.

Browse the [benchmark catalogue](studies/index.qmd) or download the [v1.1 reproducibility capsule](results/reference/01_finemapping/separated-development-v1.1/README.md).
