---
title: "Study 01: Separated-locus fine-mapping"
description: "Current ten-replicate fine-mapping and credible-set benchmark."
---

```{r}
#| echo: false
capsule <- "results/reference/01_finemapping/current"
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"),
  simplifyVector = TRUE)
summary <- read.csv(file.path(capsule, "benchmark_summary.csv"),
  check.names = FALSE)
status <- read.csv(file.path(capsule, "replicate_status.csv"),
  check.names = FALSE)
warnings <- read.csv(file.path(capsule, "target_warnings.csv"),
  check.names = FALSE)
```

::: {.callout-warning}
This is a focused development benchmark. Ten simulations support structural
validation and descriptive comparison, not a universal method ranking.
:::

## Design and completion

The current benchmark uses one separated-locus architecture, ten deterministic
replicates, ten causal markers per replicate, and four scalar methods: ST-BED
BayesC/BayesR and ST-CSR SBayesC/SBayesR. All `r nrow(status)` fits completed.
Each fit used 250 burn-in and 500 retained iterations in one chain, with no
thinning. This explicit Study 01 policy is independent of the multichain
recommendations selected in Study 04.

The run used `sblr` 0.2.0 at
`r manifest$sblr_source_commit`. Data and effects are paired across methods within
each replicate. Causal selection preserves the MAF and distance constraints,
marker identities, and canonical chromosome order.

## Marker and credible-set outcomes

```{r}
#| echo: false
keep <- c("average_precision", "causal_top_10_recall", "effect_rmse",
  "pip_brier", "exact_coverage_fraction", "ld_proxy_coverage_fraction",
  "mean_set_size")
knitr::kable(summary[summary$metric %in% keep,
  c("method", "metric", "n", "mean", "sd", "min", "max")],
  digits = 4)
```

Credible sets are cumulative marginal-PIP sets, not per-effect causal
configurations. The sparse-LD warning audit found `r nrow(warnings)` recorded
target warnings. Direct genotype correlations confirmed that every zero-edge
warning was consistent with the frozen LD threshold rather than marker-order or
indexing failure.

## Reproducibility

- [Current capsule](../../results/reference/01_finemapping/current/README.md)
- [Manifest](../../results/reference/01_finemapping/current/benchmark_manifest.json)
- [Benchmark summary](../../results/reference/01_finemapping/current/benchmark_summary.csv)
- [LD warning validation](../../results/reference/01_finemapping/current/ld_warning_validation.csv)
- [Checksums](../../results/reference/01_finemapping/current/checksums.csv)

Earlier development versions remain available through Git history.
