---
title: "Study 01: Separated-locus fine-mapping"
description: "Ten-replicate marker-recovery and credible-set benchmark."
---

```{r}
#| echo: false
source("R/benchmark-reporting.R")
capsule <- "results/reference/01_finemapping/current"
read_capsule_csv <- function(name) read.csv(file.path(capsule, name), check.names = FALSE)
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
benchmark <- read_capsule_csv("benchmark_summary.csv")
marker <- read_capsule_csv("marker_metrics.csv")
credible <- read_capsule_csv("credible_set_summary.csv")
credible_loci <- read_capsule_csv("credible_set_metrics.csv")
computational <- read_capsule_csv("computational_summary.csv")
fit_status <- read_capsule_csv("replicate_status.csv")
simulations <- read_capsule_csv("simulation_summary.csv")
ld_audit <- read_capsule_csv("ld_warning_validation.csv")
target_warnings <- read_capsule_csv("target_warnings.csv")
marker$method <- factor(marker$method, levels = sblrbench_method_levels())
credible$method <- factor(credible$method, levels = sblrbench_method_levels())
computational$method <- factor(computational$method, levels = sblrbench_method_levels())

plot_metric <- function(data, metric, label, lower_better = FALSE) {
  z <- data[data$metric == metric, ]
  p <- prepare_sblrbench_replicates(z, "method")
  ggplot2::ggplot(p$observations, ggplot2::aes(method, value, colour = method, shape = method)) +
    ggplot2::geom_point(position = ggplot2::position_jitter(width = .08), alpha = .65) +
    ggplot2::geom_pointrange(data = p$summary,
      ggplot2::aes(method, mean, ymin = minimum, ymax = maximum), inherit.aes = FALSE,
      shape = 18, linewidth = .45) +
    sblrbench_method_scales() + theme_sblrbench() +
    ggplot2::scale_x_discrete(labels = sblrbench_method_labels()) +
    ggplot2::labs(x = NULL, y = label,
      caption = paste("Points are ten deterministic replicates; diamonds show mean and observed range.",
        if (lower_better) "Lower is better." else "Higher is better.")) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 25, hjust = 1), legend.position = "none")
}
```

::: {.callout-warning}
This is a complete *development* benchmark. Ten deterministic simulations can expose implementation failures and show replicate variation, but they do not support universal method rankings or formal uncertainty statements.
:::

## Benchmark status and scientific question

All `r manifest$successful_fit_count` of `r manifest$expected_fit_count` expected fits completed and the capsule is marked `r manifest$validation_status`. The question is deliberately narrow: under separated causal loci, how do individual-level BED BayesC/BayesR and summary-data CSR SBayesC/SBayesR recover marker effects, rank causal markers, and construct marginal-PIP credible sets?

## Methods compared

The four methods cross two prior classes with two input representations. BayesC/SBayesC use a common non-zero variance class; BayesR/SBayesR use a variance mixture. BED methods fit individual-level packed genotypes, whereas CSR methods fit training summary statistics and sparse LD. Comparisons therefore describe the complete tested workflows, not just interchangeable storage formats.

## Data and simulation design

The pinned public qgdata genotype panel contains `r format(manifest$sample_count, big.mark=",")` samples and `r format(manifest$canonical_marker_count, big.mark=",")` canonical markers. Each of `r manifest$replicate_count` deterministic replicates simulates a trait at target heritability `r manifest$simulation$h2`.

Ten causal markers are selected separately in every replicate. Eligible markers have MAF in [`r manifest$causal_selection$min_maf`, `r manifest$causal_selection$max_maf`] and selected markers must be at least `r format(manifest$causal_selection$min_distance_bp, big.mark=",")` bp apart. Frozen causal-selection, simulation, method and chain seeds pair the same truth and inputs across methods. The capsule’s oracle comparison checks the generated phenotype against the intended linear predictor.

```{r}
#| echo: false
design <- data.frame(
  Item = c("Replicates", "Samples", "Canonical markers", "Causal markers per replicate", "Target heritability", "Minimum causal-marker distance"),
  Value = c(manifest$replicate_count, manifest$sample_count, manifest$canonical_marker_count,
    manifest$simulation$n_causal, manifest$simulation$h2,
    paste(format(manifest$causal_selection$min_distance_bp, big.mark=","), "bp")))
knitr::kable(design)
```

## MCMC controls and provenance

Study 01 retains its explicit one-chain policy: `r manifest$mcmc$nburn` burn-in iterations, `r manifest$mcmc$nit` retained iterations, no thinning (`nthin = `r manifest$mcmc$nthin`), one chain and one core. This policy is intentionally separate from Study 04’s later multichain recommendations.

The fits used `sblr` `r manifest$package_versions$sblr` at exact source commit `r manifest$sblr_source_commit` (`r manifest$sblr_source_status`). The benchmark source commit recorded by the numerical capsule is `r manifest$sblrbench_source_commit`.

## Validation checks

```{r}
#| echo: false
checks <- data.frame(
  Check = c("Expected fits", "Successful fits", "Failed fits", "Oracle-valid replicates",
    "Causal-count checks", "Marker/sample alignment", "LD warning targets",
    "Threshold-consistent zero-edge regions", "LD audit mismatches"),
  Result = c(manifest$expected_fit_count, manifest$successful_fit_count, manifest$failed_fit_count,
    sum(simulations$oracle_ok), sum(simulations$causal_count == manifest$simulation$n_causal),
    paste(manifest$canonical_marker_count, "markers /", manifest$sample_count, "samples"),
    length(unique(target_warnings$target)), sum(ld_audit$threshold_consistent),
    sum(!ld_audit$threshold_consistent)))
knitr::kable(checks)
```

The sparse-LD audit is substantive rather than cosmetic: direct genotype correlations were recomputed for every recorded zero-edge warning. All `r nrow(ld_audit)` audited regions were consistent with the frozen $r^2$ threshold; none indicated marker reordering, sample mismatch, or sparse-index corruption. The full fit-status table remains visible so failed branches could not disappear from a summary.

```{r}
#| echo: false
knitr::kable(fit_status)
```

## Main marker-level results

```{r}
#| echo: false
marker_table <- benchmark[benchmark$metric %in% c("effect_rmse", "pip_brier", "average_precision", "causal_top_10_recall"),
  c("method", "metric", "n", "mean", "sd", "median", "min", "max")]
knitr::kable(marker_table, digits = 4)
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
plot_metric(marker, "effect_rmse", "Marker-effect RMSE", TRUE)
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
plot_metric(marker, "pip_brier", "PIP Brier score", TRUE)
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
plot_metric(marker, "average_precision", "Average precision")
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
plot_metric(marker, "causal_top_10_recall", "Causal top-10 recall")
```

## Credible-set results

Credible sets accumulate marginal PIP to `r manifest$credible_sets$coverage` within detected loci. They are not per-effect posterior configurations. Exact coverage requires the causal marker itself; LD-proxy coverage additionally accepts a qualifying proxy at the frozen $r^2$ threshold. Coverage is conditional on constructed sets, while `causal_loci_detected` answers the separate detection question.

```{r}
#| echo: false
cs_table <- benchmark[benchmark$metric %in% c("exact_coverage_fraction", "ld_proxy_coverage_fraction", "mean_set_size", "causal_loci_detected"),
  c("method", "metric", "n", "mean", "sd", "median", "min", "max")]
knitr::kable(cs_table, digits = 4)
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
plot_metric(credible, "exact_coverage_fraction", "Exact credible-set coverage")
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
plot_metric(credible, "ld_proxy_coverage_fraction", "LD-proxy credible-set coverage")
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
plot_metric(credible, "mean_set_size", "Mean credible-set size", TRUE)
```

## Runtime comparison

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
p <- prepare_sblrbench_replicates(transform(computational, value = runtime), "method")
ggplot2::ggplot(p$observations, ggplot2::aes(method, value, colour=method, shape=method)) +
  ggplot2::geom_point(position=ggplot2::position_jitter(width=.08), alpha=.65) +
  ggplot2::geom_pointrange(data=p$summary, ggplot2::aes(method,mean,ymin=minimum,ymax=maximum), inherit.aes=FALSE, shape=18) +
  sblrbench_method_scales() + theme_sblrbench() +
  ggplot2::scale_x_discrete(labels=sblrbench_method_labels()) +
  ggplot2::labs(x=NULL, y="Fit runtime (seconds)", caption="Points are replicates; diamonds show mean and observed range.") +
  ggplot2::theme(axis.text.x=ggplot2::element_text(angle=25,hjust=1), legend.position="none")
```

## Interpretation

The tables and replicate plots should be read together: mean performance can conceal replicate-specific failures, and credible-set coverage can look favorable simply because sets are large. Marker recovery, causal ranking, locus detection, exact/proxy coverage, set size and runtime therefore form a joint descriptive profile. No single metric licenses an overall winner.

## Limitations

Only one separated-locus architecture, one public genotype panel, ten replicates and short one-chain fits were studied. Marginal-PIP credible sets do not represent full multi-effect posterior configurations. LD-proxy coverage depends on the chosen threshold, and observed min/max ranges are descriptive—not confidence intervals. Convergence cannot be established from the one-chain Study 01 policy.

## Reproducibility and downloads

Rendering reads the compact capsule only; it does not run simulation, LD construction or a sampler. The complete [analysis workflow](analysis.R), [scientific specification](spec.R), and reusable [fine-mapping template](../../inst/templates/finemapping-analysis.R) show how to run the study separately.

- [Current capsule](../../results/reference/01_finemapping/current/README.md)
- [Manifest](../../results/reference/01_finemapping/current/benchmark_manifest.json)
- [Marker metrics](../../results/reference/01_finemapping/current/marker_metrics.csv)
- [Credible-set metrics](../../results/reference/01_finemapping/current/credible_set_metrics.csv)
- [Credible-set summary](../../results/reference/01_finemapping/current/credible_set_summary.csv)
- [Computational summary](../../results/reference/01_finemapping/current/computational_summary.csv)
- [LD warning audit](../../results/reference/01_finemapping/current/ld_warning_validation.csv)
- [Checksums](../../results/reference/01_finemapping/current/checksums.csv)
