---
title: "Separated-locus fine-mapping development pilot"
description: "A frozen 10-replicate structural benchmark of four single-trait sblr workflows."
---

> **Development benchmark—not a final scientific comparison.**  
> Markers in the example Glist have limited LD, and models were run with 500 iterations, 250 burn-in iterations and one chain. Results primarily validate orchestration, alignment, fitting, extraction and reporting.

```{r}
#| echo: false
capsule <- "results/reference/01_finemapping/separated-development-v1"
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
benchmark <- read.csv(file.path(capsule, "benchmark_summary.csv"), check.names = FALSE)
marker <- read.csv(file.path(capsule, "marker_metrics.csv"), check.names = FALSE)
cs <- read.csv(file.path(capsule, "credible_set_summary.csv"), check.names = FALSE)
computational <- read.csv(file.path(capsule, "computational_summary.csv"), check.names = FALSE)
fit_status <- read.csv(file.path(capsule, "replicate_status.csv"), check.names = FALSE)
sources <- read.csv(file.path(capsule, "source_files.csv"), check.names = FALSE)
labels <- setNames(benchmark$method_label, benchmark$method)
fmt_pm <- function(x, digits = 3) sprintf(paste0("%.", digits, "f ± %.", digits, "f"), x[1], x[2])
metric_wide <- function(metrics) marker[marker$metric %in% metrics, c("replicate", "method", "metric", "value")]
plot_metric <- function(metric, ylab, lower_better = FALSE) {
  z <- marker[marker$metric == metric, ]
  groups <- split(z$value, factor(z$method, levels = benchmark$method))
  boxplot(groups, names = benchmark$method_label, ylab = ylab, las = 2,
          col = "#dbeafe", border = "#2563eb", outline = FALSE)
  for (i in seq_along(groups)) points(jitter(rep(i, length(groups[[i]])), amount = .08),
                                      groups[[i]], pch = 19, col = "#0f172acc")
}
```

## Benchmark status

The frozen capsule records `r manifest$replicate_count` deterministic replicates and `r sum(fit_status$status == "ok")`/`r nrow(fit_status)` successful fits. It is a development infrastructure benchmark.

## Scientific question

Can one complete separated-causal-locus path—simulation, oracle validation, BED and CSR fitting, standardized extraction, marker metrics, credible sets and reporting—run reproducibly across four single-trait methods?

## Methods compared

`r paste(benchmark$method_label, collapse = ", ")`.

## Data and simulation design

```{r}
#| echo: false
design <- data.frame(
  Setting = c("Architecture", "Replicates", "Canonical markers", "Causal variants per replicate", "Heritability",
    "Minimum causal distance", "MAF range", "Methods", "MCMC", "Software"),
  Value = c(manifest$architecture, manifest$replicate_count, format(manifest$canonical_marker_count, big.mark = ","),
    manifest$simulation$n_causal, manifest$simulation$h2,
    paste(format(manifest$causal_selection$min_distance_bp, big.mark = ","), "bp"),
    paste0("[", manifest$causal_selection$min_maf, ", ", manifest$causal_selection$max_maf, "]"),
    paste(manifest$methods, collapse = ", "),
    paste0("nit=", manifest$mcmc$nit, ", nburn=", manifest$mcmc$nburn, ", nthin=", manifest$mcmc$nthin,
      ", chains=", manifest$mcmc$nchains, ", cores=", manifest$mcmc$ncores),
    paste0("sblr ", manifest$package_versions$sblr, "; sblrbench ", manifest$package_versions$sblrbench,
      "; qgg ", manifest$package_versions$qgg)), stringsAsFactors = FALSE)
knitr::kable(design)
```

## Configuration

The exact immutable settings are in [`config.R`](/results/reference/01_finemapping/separated-development-v1/config.R). Changed settings require a new versioned snapshot.

## Validation checks

```{r}
#| echo: false
v <- manifest$validation
checks <- data.frame(Check = c("Distinct deterministic causal sets", "Exact causals per replicate",
  "Minimum observed causal separation", "Oracle checks", "Successful fits", "Canonical markers",
  "Marker alignment", "Sample alignment"),
  Result = c(v$distinct_causal_sets, manifest$simulation$n_causal,
    paste(format(v$minimum_observed_causal_distance_bp, big.mark = ","), "bp"),
    paste0(v$oracle_checks_passed, "/", manifest$replicate_count, " passed"),
    paste0(v$successful_fits, "/", v$expected_fits), format(manifest$canonical_marker_count, big.mark = ","),
    if (v$marker_alignment_validated) "Validated" else "Not recorded",
    if (v$sample_alignment_validated) "Validated" else "Not recorded"))
knitr::kable(checks)
```

## Results

```{r}
#| echo: false
main_table <- data.frame(
  Method = benchmark$method_label,
  `Average precision` = mapply(function(a,b) fmt_pm(c(a,b)), benchmark$average_precision_mean, benchmark$average_precision_sd),
  `PIP Brier` = mapply(function(a,b) fmt_pm(c(a,b), 5), benchmark$pip_brier_mean, benchmark$pip_brier_sd),
  `Effect RMSE` = mapply(function(a,b) fmt_pm(c(a,b), 5), benchmark$effect_rmse_mean, benchmark$effect_rmse_sd),
  `Top-10 recall` = mapply(function(a,b) fmt_pm(c(a,b)), benchmark$causal_top_10_recall_mean, benchmark$causal_top_10_recall_sd),
  `Causal loci detected` = sprintf("%.1f / %d", benchmark$causal_loci_detected_mean, manifest$simulation$n_causal),
  `Detection fraction` = sprintf("%.3f", benchmark$causal_locus_detection_fraction),
  `Conditional exact coverage` = sprintf("%.3f", benchmark$conditional_exact_coverage_mean),
  `Mean set size` = sprintf("%.2f", benchmark$mean_set_size),
  `Runtime (s)` = mapply(function(a,b) fmt_pm(c(a,b), 1), benchmark$runtime_mean_seconds, benchmark$runtime_sd_seconds),
  check.names = FALSE)
knitr::kable(main_table)
```

### Replicate-level plots

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Average precision across the ten replicates. Points are replicate-level observations."
plot_metric("average_precision", "Average precision")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "PIP Brier score across replicates; lower is better."
plot_metric("pip_brier", "PIP Brier score")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Effect RMSE across replicates; lower is better."
plot_metric("effect_rmse", "Effect RMSE")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Top-10 causal recall across replicates."
plot_metric("causal_top_10_recall", "Top-10 causal recall")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Number of true causal loci with a constructed credible set."
z <- cs[cs$metric == "causal_loci_detected", ]
groups <- split(z$value, factor(z$method, levels = benchmark$method))
boxplot(groups, names = benchmark$method_label, ylab = "Causal loci detected", las = 2,
  col = "#dcfce7", border = "#15803d", outline = FALSE)
for (i in seq_along(groups)) points(jitter(rep(i, length(groups[[i]])), amount=.08), groups[[i]], pch=19, col="#14532dcc")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Elapsed fitting time by method."
groups <- split(computational$runtime, factor(computational$method, levels = benchmark$method))
boxplot(groups, names = benchmark$method_label, ylab = "Runtime (seconds)", las = 2,
  col = "#fef3c7", border = "#b45309", outline = FALSE)
for (i in seq_along(groups)) points(jitter(rep(i, length(groups[[i]])), amount=.08), groups[[i]], pch=19, col="#78350fcc")
```

### Causal-marker ranking

```{r}
#| echo: false
rank_metrics <- c("causal_rank_best", "causal_rank_median", "causal_rank_mean", "causal_rank_worst",
                  "causal_top_10_recall", "causal_top_20_recall", "causal_top_50_recall")
rank_table <- do.call(rbind, lapply(benchmark$method, function(m) {
  vals <- sapply(rank_metrics, function(q) mean(marker$value[marker$method == m & marker$metric == q]))
  data.frame(Method = labels[[m]], t(vals), check.names = FALSE)
}))
names(rank_table)[-1] <- c("Best rank", "Median rank", "Mean rank", "Worst rank", "Top-10 recall", "Top-20 recall", "Top-50 recall")
knitr::kable(rank_table, digits = 3)
```

A low median rank paired with a large mean or worst rank means most causal variants are recovered well while a small subset receives little support.

### Credible sets

```{r}
#| echo: false
cs_table <- benchmark[c("method_label", "conditional_exact_coverage_mean", "conditional_ld_proxy_coverage_mean",
  "causal_locus_detection_fraction", "credible_sets_mean", "mean_set_size")]
names(cs_table) <- c("Method", "Conditional exact coverage", "Conditional LD-proxy coverage",
  "Causal-locus detection fraction", "Number of sets", "Mean set size")
knitr::kable(cs_table, digits = 3)
```

Markers have limited LD, so credible sets are mostly or entirely singleton sets and exact and proxy coverage may be identical. These results validate structure rather than realistic fine-mapping uncertainty.

Some BayesR branches reported regional sparse-LD matrices with zero off-diagonal entries. This is consistent with the low-LD example data but must be revisited with realistic LD before scientific interpretation.

### Runtime

BED and CSR runtimes differ in this small development pilot, but the observed timing is specific to these inputs, controls and environment. It does not establish general computational superiority.

## Interpretation

All four workflows complete; extracted outputs and truth-aware metrics are aligned and finite; BED and CSR show broadly similar pilot behaviour; and the reporting pipeline is ready for realistic-LD and scientific-length studies.

## Limitations

- Limited LD and no realistic competition among correlated markers.
- Only 500 MCMC iterations, 250 burn-in iterations and one chain.
- Ten replicates, one heritability and separated causal loci only.
- No final convergence assessment.
- No scientific method-ranking claims.
- Cumulative marginal-PIP sets are not guaranteed per-effect causal configurations.

## Code and reproduction

### Minimal runnable example

This fast synthetic example validates the public simulation, oracle, result and metric contracts. It deliberately avoids a full sampler because public samplers require BED/Glist or CSR/LD inputs.

```{r}
#| echo: false
#| results: asis
cat("```r\n", paste(readLines(file.path(capsule, "minimal_example.R")), collapse="\n"), "\n```\n", sep="")
```

[Download `minimal_example.R`](/results/reference/01_finemapping/separated-development-v1/minimal_example.R).

### Exact benchmark invocation

```{r}
#| echo: false
#| results: asis
cat("```r\n", paste(readLines(file.path(capsule, "run_benchmark.R")), collapse="\n"), "\n```\n", sep="")
```

[Download `run_benchmark.R`](/results/reference/01_finemapping/separated-development-v1/run_benchmark.R). The genotype input is not redistributed; invoking this workflow requires the original or a compatible Glist.

### Source code

```{r}
#| echo: false
base_url <- paste0("https://github.com/psoerensen/sblrbench/blob/", manifest$source_snapshot_commit, "/")
source_links <- data.frame(File = sources$path, Role = sources$role,
  Source = paste0("[view](", base_url, sources$path, ")"), check.names = FALSE)
knitr::kable(source_links, escape = FALSE)
```

### Reproducibility level

> The public capsule supports complete inspection of the configuration, reporting code, metric definitions and study pipeline. Exact numerical reproduction additionally requires the original genotype data, which are not redistributed.

## Downloads

**Benchmark reproducibility capsule:** [capsule README](/results/reference/01_finemapping/separated-development-v1/README.md)

- [Benchmark summary](/results/reference/01_finemapping/separated-development-v1/benchmark_summary.csv)
- [Marker metrics](/results/reference/01_finemapping/separated-development-v1/marker_metrics.csv)
- [Credible-set summary](/results/reference/01_finemapping/separated-development-v1/credible_set_summary.csv)
- [Credible-set metrics](/results/reference/01_finemapping/separated-development-v1/credible_set_metrics.csv)
- [Computational summary](/results/reference/01_finemapping/separated-development-v1/computational_summary.csv)
- [Replicate status](/results/reference/01_finemapping/separated-development-v1/replicate_status.csv)
- [Benchmark manifest](/results/reference/01_finemapping/separated-development-v1/benchmark_manifest.json)
- [Frozen configuration](/results/reference/01_finemapping/separated-development-v1/config.R)
- [Exact run script](/results/reference/01_finemapping/separated-development-v1/run_benchmark.R)
- [Minimal example](/results/reference/01_finemapping/separated-development-v1/minimal_example.R)
- [Source inventory](/results/reference/01_finemapping/separated-development-v1/source_files.csv)
- [Session information](/results/reference/01_finemapping/separated-development-v1/session_info.txt)
- [Checksums](/results/reference/01_finemapping/separated-development-v1/checksums.csv)

## Software versions and commits

```{r}
#| echo: false
versions <- data.frame(Item = c("sblrbench source snapshot", "Fit provenance commit", "sblr Git commit",
  "sblr", "sblrbench", "qgg"), Value = c(manifest$source_snapshot_commit, manifest$fit_provenance_commit,
  manifest$sblr_git_commit, manifest$package_versions$sblr, manifest$package_versions$sblrbench,
  manifest$package_versions$qgg))
knitr::kable(versions)
```
