---
title: "Study 01: Separated-locus fine-mapping"
description: "A frozen 10-replicate structural benchmark of four single-trait sblr workflows."
---

::: {.callout-warning .callout-development}
## Development benchmark—not a final scientific comparison.
Markers in the example Glist have limited LD, and models were run with 500 iterations, 250 burn-in iterations and one chain. Results primarily validate orchestration, alignment, fitting, extraction and reporting.
:::

```{r}
#| echo: false
capsule <- "results/reference/01_finemapping/separated-development-v1.1"
source("studies/reporting_helpers.R")
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
benchmark <- read.csv(file.path(capsule, "benchmark_summary.csv"), check.names = FALSE)
marker <- read.csv(file.path(capsule, "marker_metrics.csv"), check.names = FALSE)
cs <- read.csv(file.path(capsule, "credible_set_summary.csv"), check.names = FALSE)
computational <- read.csv(file.path(capsule, "computational_summary.csv"), check.names = FALSE)
fit_status <- read.csv(file.path(capsule, "replicate_status.csv"), check.names = FALSE)
sources <- read.csv(file.path(capsule, "source_files.csv"), check.names = FALSE)
example_data <- read.csv(file.path(capsule, "example_data_manifest.csv"), check.names = FALSE)
labels <- setNames(benchmark$method_label, benchmark$method)
fmt_pm <- function(x, digits = 3) sprintf(paste0("%.", digits, "f ± %.", digits, "f"), x[1], x[2])
metric_wide <- function(metrics) marker[marker$metric %in% metrics, c("replicate", "method", "metric", "value")]
plot_metric <- function(metric, ylab, lower_better = FALSE) {
  z <- marker[marker$metric == metric, ]
  prepared <- prepare_sblrbench_replicates(z, "method")
  ggplot2::ggplot(prepared$observations, ggplot2::aes(method, value, colour=method, shape=method)) +
    ggplot2::geom_point(position=ggplot2::position_jitter(width=.08,height=0),alpha=.65) +
    ggplot2::geom_pointrange(data=prepared$summary,
      ggplot2::aes(method,mean,ymin=minimum,ymax=maximum),inherit.aes=FALSE,shape=18,size=.5) +
    sblrbench_method_scales() + theme_sblrbench() +
    ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels()) +
    ggplot2::labs(x=NULL,y=ylab,caption=if(lower_better) "Lower values are better." else NULL) +
    ggplot2::theme(axis.text.x=ggplot2::element_text(angle=25,hjust=1),legend.position="none")
}
```

## Benchmark status

The frozen capsule records `r manifest$replicate_count` deterministic replicates and `r sum(fit_status$status == "ok")`/`r nrow(fit_status)` successful fits. It is a development infrastructure benchmark.

## Scientific question

Can one complete separated-causal-locus path—simulation, oracle validation, BED and CSR fitting, standardized extraction, marker metrics, credible sets and reporting—run reproducibly across four single-trait methods?

## Methods compared

`r paste(benchmark$method_label, collapse = ", ")`.

## Data and simulation design

```{r}
#| echo: false
design <- data.frame(
  Setting = c("Architecture", "Replicates", "Canonical markers", "Causal variants per replicate", "Heritability",
    "Minimum causal distance", "MAF range", "Methods", "MCMC", "Software"),
  Value = c(manifest$architecture, manifest$replicate_count, format(manifest$canonical_marker_count, big.mark = ","),
    manifest$simulation$n_causal, manifest$simulation$h2,
    paste(format(manifest$causal_selection$min_distance_bp, big.mark = ","), "bp"),
    paste0("[", manifest$causal_selection$min_maf, ", ", manifest$causal_selection$max_maf, "]"),
    paste(manifest$methods, collapse = ", "),
    paste0("nit=", manifest$mcmc$nit, ", nburn=", manifest$mcmc$nburn, ", nthin=", manifest$mcmc$nthin,
      ", chains=", manifest$mcmc$nchains, ", cores=", manifest$mcmc$ncores),
    paste0("sblr ", manifest$package_versions$sblr, "; sblrbench ", manifest$package_versions$sblrbench,
      "; qgg ", manifest$package_versions$qgg)), stringsAsFactors = FALSE)
knitr::kable(design)
```

## Configuration

The exact immutable settings are in [`config.R`](../../results/reference/01_finemapping/separated-development-v1.1/config.R). Changed settings require a new versioned snapshot.

## Validation checks

```{r}
#| echo: false
v <- manifest$validation
checks <- data.frame(Check = c("Distinct deterministic causal sets", "Exact causals per replicate",
  "Minimum observed causal separation", "Oracle checks", "Successful fits", "Canonical markers",
  "Marker alignment", "Sample alignment"),
  Result = c(v$distinct_causal_sets, manifest$simulation$n_causal,
    paste(format(v$minimum_observed_causal_distance_bp, big.mark = ","), "bp"),
    paste0(v$oracle_checks_passed, "/", manifest$replicate_count, " passed"),
    paste0(v$successful_fits, "/", v$expected_fits), format(manifest$canonical_marker_count, big.mark = ","),
    if (v$marker_alignment_validated) "Validated" else "Not recorded",
    if (v$sample_alignment_validated) "Validated" else "Not recorded"))
knitr::kable(checks)
```

## Results

```{r}
#| echo: false
main_table <- data.frame(
  Method = benchmark$method_label,
  `Average precision` = mapply(function(a,b) fmt_pm(c(a,b)), benchmark$average_precision_mean, benchmark$average_precision_sd),
  `PIP Brier` = mapply(function(a,b) fmt_pm(c(a,b), 5), benchmark$pip_brier_mean, benchmark$pip_brier_sd),
  `Effect RMSE` = mapply(function(a,b) fmt_pm(c(a,b), 5), benchmark$effect_rmse_mean, benchmark$effect_rmse_sd),
  `Top-10 recall` = mapply(function(a,b) fmt_pm(c(a,b)), benchmark$causal_top_10_recall_mean, benchmark$causal_top_10_recall_sd),
  `Causal loci detected` = sprintf("%.1f / %d", benchmark$causal_loci_detected_mean, manifest$simulation$n_causal),
  `Detection fraction` = sprintf("%.3f", benchmark$causal_locus_detection_fraction),
  `Conditional exact coverage` = sprintf("%.3f", benchmark$conditional_exact_coverage_mean),
  `Mean set size` = sprintf("%.2f", benchmark$mean_set_size),
  `Runtime (s)` = mapply(function(a,b) fmt_pm(c(a,b), 1), benchmark$runtime_mean_seconds, benchmark$runtime_sd_seconds),
  check.names = FALSE)
knitr::kable(main_table)
```

### Replicate-level plots

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Average precision across the ten replicates. Points are replicate-level observations."
plot_metric("average_precision", "Average precision")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "PIP Brier score across replicates; lower is better."
plot_metric("pip_brier", "PIP Brier score")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Effect RMSE across replicates; lower is better."
plot_metric("effect_rmse", "Effect RMSE")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Top-10 causal recall across replicates."
plot_metric("causal_top_10_recall", "Top-10 causal recall")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Number of true causal loci with a constructed credible set."
z <- cs[cs$metric == "causal_loci_detected", ]
p <- prepare_sblrbench_replicates(z,"method")
ggplot2::ggplot(p$observations,ggplot2::aes(method,value,colour=method,shape=method))+
  ggplot2::geom_point(position=ggplot2::position_jitter(width=.08,height=0),alpha=.65)+
  ggplot2::geom_pointrange(data=p$summary,ggplot2::aes(method,mean,ymin=minimum,ymax=maximum),inherit.aes=FALSE,shape=18)+
  sblrbench_method_scales()+theme_sblrbench()+ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())+
  ggplot2::labs(x=NULL,y="Causal loci detected")+ggplot2::theme(axis.text.x=ggplot2::element_text(angle=25,hjust=1),legend.position="none")
```

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Elapsed fitting time by method."
p <- prepare_sblrbench_replicates(transform(computational,value=runtime),"method")
ggplot2::ggplot(p$observations,ggplot2::aes(method,value,colour=method,shape=method))+
  ggplot2::geom_point(position=ggplot2::position_jitter(width=.08,height=0),alpha=.65)+
  ggplot2::geom_pointrange(data=p$summary,ggplot2::aes(method,mean,ymin=minimum,ymax=maximum),inherit.aes=FALSE,shape=18)+
  sblrbench_method_scales()+theme_sblrbench()+ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())+
  ggplot2::labs(x=NULL,y="Runtime (seconds)")+ggplot2::theme(axis.text.x=ggplot2::element_text(angle=25,hjust=1),legend.position="none")
```

### Causal-marker ranking

```{r}
#| echo: false
rank_metrics <- c("causal_rank_best", "causal_rank_median", "causal_rank_mean", "causal_rank_worst",
                  "causal_top_10_recall", "causal_top_20_recall", "causal_top_50_recall")
rank_table <- do.call(rbind, lapply(benchmark$method, function(m) {
  vals <- sapply(rank_metrics, function(q) mean(marker$value[marker$method == m & marker$metric == q]))
  data.frame(Method = labels[[m]], t(vals), check.names = FALSE)
}))
names(rank_table)[-1] <- c("Best rank", "Median rank", "Mean rank", "Worst rank", "Top-10 recall", "Top-20 recall", "Top-50 recall")
knitr::kable(rank_table, digits = 3)
```

A low median rank paired with a large mean or worst rank means most causal variants are recovered well while a small subset receives little support.

### Credible sets

```{r}
#| echo: false
cs_table <- benchmark[c("method_label", "conditional_exact_coverage_mean", "conditional_ld_proxy_coverage_mean",
  "causal_locus_detection_fraction", "credible_sets_mean", "mean_set_size")]
names(cs_table) <- c("Method", "Conditional exact coverage", "Conditional LD-proxy coverage",
  "Causal-locus detection fraction", "Number of sets", "Mean set size")
knitr::kable(cs_table, digits = 3)
```

Markers have limited LD, so credible sets are mostly or entirely singleton sets and exact and proxy coverage may be identical. These results validate structure rather than realistic fine-mapping uncertainty.

Some BayesR branches reported regional sparse-LD matrices with zero off-diagonal entries. This is consistent with the low-LD example data but must be revisited with realistic LD before scientific interpretation.

## Computational performance

BED and CSR runtimes differ in this small development pilot, but the observed timing is specific to these inputs, controls and environment. It does not establish general computational superiority.

## Interpretation

All four workflows complete; extracted outputs and truth-aware metrics are aligned and finite; BED and CSR show broadly similar pilot behaviour; and the reporting pipeline is ready for realistic-LD and scientific-length studies.

## Limitations

- Limited LD and no realistic competition among correlated markers.
- Only 500 MCMC iterations, 250 burn-in iterations and one chain.
- Ten replicates, one heritability and separated causal loci only.
- No final convergence assessment.
- No scientific method-ranking claims.
- Cumulative marginal-PIP sets are not guaranteed per-effect causal configurations.

## Code and reproduction

### Worked example

This example downloads the public simulated PLINK data used by the study, creates a Glist, simulates one phenotype, fits one actual `sblr` model, and evaluates recovery of the known causal effects. It uses 400 samples, 500 markers, and short demonstration-only ST-BED BayesC settings.

```{r}
#| echo: false
#| results: asis
display_capsule_script(file.path(capsule, "worked_finemapping_example.R"))
```

[Download `worked_finemapping_example.R`](../../results/reference/01_finemapping/separated-development-v1.1/worked_finemapping_example.R).

### Public example data

The five files are simulated and publicly accessible from [`psoerensen/qgdata`](https://github.com/psoerensen/qgdata) at pinned commit `r manifest$public_data_source$commit`. The local benchmark inputs match these sizes and checksums exactly.

```{r}
#| echo: false
data_table <- example_data[c("filename", "role", "size_bytes", "md5", "download_url")]
data_table$download_url <- paste0("[download](", data_table$download_url, ")")
names(data_table) <- c("File", "Role", "Bytes", "MD5", "Pinned download")
knitr::kable(data_table, escape = FALSE)
```

No explicit data licence was found in the qgdata repository root at the pinned revision. The files are publicly accessible, but reuse terms should be clarified with the repository owner.

[Download `example_data_manifest.csv`](../../results/reference/01_finemapping/separated-development-v1.1/example_data_manifest.csv).

### Developer contract smoke test

This secondary test checks simulation, oracle, result-object, and metric contracts without fitting a sampler.

```{r}
#| echo: false
#| results: asis
display_capsule_script_collapsed(file.path(capsule, "contract_smoke_test.R"))
```

[Download `contract_smoke_test.R`](../../results/reference/01_finemapping/separated-development-v1.1/contract_smoke_test.R).

### Complete benchmark reproduction

The targets entry point downloads the same pinned public data when needed, validates checksums, reuses the local data and `_targets/` caches, and runs only missing or outdated work.

```{r}
#| echo: false
#| results: asis
display_capsule_script(file.path(capsule, "run_benchmark.R"))
```

[Download `run_benchmark.R`](../../results/reference/01_finemapping/separated-development-v1.1/run_benchmark.R).

### Source code

```{r}
#| echo: false
base_url <- paste0("https://github.com/psoerensen/sblrbench/blob/", manifest$source_snapshot_commit, "/")
published_paths <- c("_targets.R", "studies/01_finemapping/config.R",
  "studies/01_finemapping/targets.R", "studies/01_finemapping/setup_example_data.R",
  "studies/01_finemapping/pilot.R", "R/metrics.R", "R/alignment.R",
  "R/simulation.R", "R/adapter-sblr.R")
published_sources <- sources[sources$path %in% published_paths, ]
source_links <- data.frame(File = published_sources$path, Role = published_sources$role,
  Source = paste0("[view](", base_url, published_sources$path, ")"), check.names = FALSE)
knitr::kable(source_links, escape = FALSE)
```

The v1.1 capsule additionally records content hashes for the new downloader and example scripts in `source_files.csv`; those files are directly downloadable below pending the next reviewed source commit.

### Reproducibility level

> The simulated genotype files are publicly available from the pinned qgdata revision and are downloaded automatically. The reference capsule records file checksums, software versions, seeds, and study code. Small numerical differences may still arise across platforms and numerical libraries.

## Downloads and provenance

**Benchmark reproducibility capsule:** [v1.1 capsule README](../../results/reference/01_finemapping/separated-development-v1.1/README.md). The [original v1 capsule](../../results/reference/01_finemapping/separated-development-v1/README.md) remains available unchanged.

| Resource | Download |
|---|---|
| Primary results | [Benchmark summary](../../results/reference/01_finemapping/separated-development-v1.1/benchmark_summary.csv); [marker metrics](../../results/reference/01_finemapping/separated-development-v1.1/marker_metrics.csv); [credible-set summary](../../results/reference/01_finemapping/separated-development-v1.1/credible_set_summary.csv) |
| Computational evidence | [Computational summary](../../results/reference/01_finemapping/separated-development-v1.1/computational_summary.csv); [replicate status](../../results/reference/01_finemapping/separated-development-v1.1/replicate_status.csv) |
| Reproduction | [Configuration](../../results/reference/01_finemapping/separated-development-v1.1/config.R); [worked example](../../results/reference/01_finemapping/separated-development-v1.1/worked_finemapping_example.R); [smoke test](../../results/reference/01_finemapping/separated-development-v1.1/contract_smoke_test.R); [run script](../../results/reference/01_finemapping/separated-development-v1.1/run_benchmark.R) |
| Provenance | [Manifest](../../results/reference/01_finemapping/separated-development-v1.1/benchmark_manifest.json); [example data](../../results/reference/01_finemapping/separated-development-v1.1/example_data_manifest.csv); [source inventory](../../results/reference/01_finemapping/separated-development-v1.1/source_files.csv); [session information](../../results/reference/01_finemapping/separated-development-v1.1/session_info.txt); [checksums](../../results/reference/01_finemapping/separated-development-v1.1/checksums.csv) |

## Software versions and commits

```{r}
#| echo: false
versions <- data.frame(Item = c("sblrbench source snapshot", "Fit provenance commit", "sblr Git commit",
  "sblr", "sblrbench", "qgg"), Value = c(manifest$source_snapshot_commit, manifest$fit_provenance_commit,
  manifest$sblr_git_commit, manifest$package_versions$sblr, manifest$package_versions$sblrbench,
  manifest$package_versions$qgg))
knitr::kable(versions)
```
