---
title: "Study 06: Annotation-informed models"
description: "In-development maximum-history BayesRC/SBayesRC evidence and prespecified stop."
---

```{r}
#| echo: false
source("studies/reporting_helpers.R")
capsule <- "results/reference/06_annotation_models/current-stop"
rcsv <- function(x) read.csv(file.path(capsule,x),check.names=FALSE)
manifest <- jsonlite::read_json(file.path(capsule,"benchmark_manifest.json"),simplifyVector=TRUE)
design <- rcsv("annotation_design_summary.csv")
true_alpha <- rcsv("true_alpha.csv")
draws <- rcsv("scalar_chain_draws.csv")
diagnostics <- rcsv("convergence_diagnostics.csv")
candidates <- rcsv("candidate_settings.csv")
recommendations <- rcsv("method_recommendations.csv")
status <- rcsv("fit_status.csv")
computation <- rcsv("computational_summary.csv")
method_labels <- c(st_bed_bayesrc="ST-BED BayesRC",st_csr_sbayesrc="ST-CSR SBayesRC")
draws$method_label <- factor(draws$method,levels=names(method_labels),labels=method_labels)
diagnostics$method_label <- factor(diagnostics$method,levels=names(method_labels),labels=method_labels)
```

::: {.callout-important}
**Status: In development.** Study 06 stopped at its prespecified convergence
gate. It is not a completed annotation-model performance benchmark, and no
five-replicate benchmark was started.
:::

## Scope and status

The completed work is an interface, annotation-design and convergence-selection study for public BayesRC and SBayesRC interfaces. Both maximum-history fits completed, but no candidate window passed every unchanged diagnostic threshold. The capsule therefore records `r manifest$validation_status`, with `full_benchmark_started = `r manifest$full_benchmark_started`.

## Public interfaces evaluated

ST-BED BayesRC consumes individual-level packed genotypes; ST-CSR SBayesRC consumes summary statistics and sparse LD. The audit fixed marker orientation, annotation row order, intercept handling, mixture-variance ordering and chain-specific output extraction before any convergence decision. The installed package was `sblr` `r manifest$installed_sblr_version` at `r manifest$installed_sblr_commit`; sibling source access was read-only.

## Annotation design

The single deterministic pilot includes an intercept, a binary enriched annotation, a continuous signal and a null annotation. The design table records prevalence, moments, rank and the annotation-implied expected non-null allocation.

```{r}
#| echo: false
knitr::kable(design,digits=4)
knitr::kable(true_alpha,digits=4,caption="Generating probit stick coefficients by annotation and stick.")
```

## Probit stick-breaking interpretation

For each marker, linear annotation predictors enter sequential probit sticks that define marker-specific prior component probabilities. These are prior probabilities implied by annotations—not posterior results. Four distinct objects must remain separate:

1. **Annotation-implied prior probabilities** are deterministic functions of the annotation row and alpha coefficients before observing the trait.
2. **Posterior component-allocation probabilities** summarize a marker’s sampled allocation after observing data.
3. **Global mixture proportions** average allocation behavior across markers and are not an annotation coefficient.
4. **Posterior marker non-null probabilities** collapse non-null components for marker-level selection and are not identical to either prior enrichment or a global mixture proportion.

Conflating these quantities would turn a convergence failure in an alpha stick into an unsupported biological conclusion.

## MCMC and chain-output contract

Each method used four identifiable chains and one core per chain under a maximum stored history of `r manifest$maximum_history$nit` iterations, no initial burn-in and no thinning. Candidate views use burn-in values `r paste(manifest$maximum_history$burnin_candidates,collapse=", ")` and retained lengths `r paste(manifest$maximum_history$retained_candidates,collapse=", ")`; they slice the same maximum-history fits rather than refitting.

The extraction contract preserves `quantity`, parameter/annotation/stick labels, chain, iteration and value. Finite non-constant values and exactly four chains are required before R-hat, ESS or relative MCSE can pass.

## Maximum-history fits

```{r}
#| echo: false
knitr::kable(status)
knitr::kable(computation,digits=3)
```

## Diagnostic thresholds and candidate windows

R-hat ≤ `r manifest$maximum_history$thresholds$rhat`, bulk ESS ≥ `r manifest$maximum_history$thresholds$ess_bulk`, tail ESS ≥ `r manifest$maximum_history$thresholds$ess_tail`, relative MCSE ≤ `r manifest$maximum_history$thresholds$relative_mcse`, and four valid chains were required simultaneously for every registered quantity.

```{r}
#| echo: false
knitr::kable(candidates,digits=4)
```

## Diagnostic evidence for the limiting quantity

The recurring limiter is `alpha:continuous_signal:component_0_stick`. Its trace retains visible chain separation even where many global scalar summaries have adequate effective sample size.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
limiter <- "alpha:continuous_signal:component_0_stick"
z <- draws[draws$quantity==limiter & draws$iteration %% 10==0,]
ggplot2::ggplot(z,ggplot2::aes(iteration,value,colour=factor(chain),group=chain))+
 ggplot2::geom_line(linewidth=.4,alpha=.8)+ggplot2::facet_wrap(~method_label,scales="free_y")+
 theme_sblrbench()+ggplot2::labs(x="Iteration",y="Alpha coefficient",colour="Chain",
 caption="Every tenth stored draw is displayed; diagnostics use complete frozen histories.")
```

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
z <- diagnostics[diagnostics$quantity==limiter,]
dl <- rbind(transform(z,diagnostic="R-hat",value=rhat,threshold=1.01),
 transform(z,diagnostic="Bulk ESS",value=ess_bulk,threshold=400),
 transform(z,diagnostic="Tail ESS",value=ess_tail,threshold=400),
 transform(z,diagnostic="Relative MCSE",value=relative_mcse,threshold=.05))
ggplot2::ggplot(dl,ggplot2::aes(retained_draws,value,colour=factor(burnin),group=burnin))+
 ggplot2::geom_line()+ggplot2::geom_point()+ggplot2::facet_grid(diagnostic~method_label,scales="free_y")+
 theme_sblrbench()+ggplot2::labs(x="Retained draws",y=NULL,colour="Burn-in",
 caption="Panels use their registered thresholds: maximum for R-hat/relative MCSE, minimum for ESS.")
```

## Broader annotation-quantity evidence

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6
key_quantities <- c(limiter,"prior_nonnull_mean_enriched","prior_nonnull_mean_unannotated","global_nonnull_proportion")
z <- diagnostics[diagnostics$quantity %in% key_quantities & diagnostics$burnin==1000,]
ggplot2::ggplot(z,ggplot2::aes(retained_draws,rhat,colour=quantity,group=quantity))+
 ggplot2::geom_hline(yintercept=1.01,linetype=2)+ggplot2::geom_line()+ggplot2::geom_point()+
 ggplot2::facet_wrap(~method_label)+theme_sblrbench()+
 ggplot2::labs(x="Retained draws after 1,000 burn-in",y="R-hat",colour="Quantity",
 caption="Dashed line is the unchanged maximum; distinct prior/global quantities are shown separately.")
```

## Least-failing settings and recurring limitations

```{r}
#| echo: false
knitr::kable(recommendations,digits=4)
```

For both methods the least-failing recorded setting uses 1,000 burn-in and 2,000 retained draws with four chains, yet the continuous-signal component-0 alpha stick fails R-hat. Longer retained windows did not supply an all-quantity pass under the frozen candidate grid. This common failure across BED and CSR points to the annotation coefficient geometry/mixing problem rather than a representation-specific numerical exception, while remaining only a diagnosis—not proof of mechanism.

## Prespecified stop decision

The decision rule required at least one supported setting for every method before beginning a five-replicate scientific benchmark. Neither method met it. Thresholds, seeds and registered quantities were not changed after seeing the result, so the correct scientific output is a documented stop rather than performance tables from inadequately mixed chains.

No annotation-model accuracy, enrichment recovery, marker selection or prediction benchmark was started. There is therefore no basis here for a BayesRC-versus-SBayesRC ranking.

## Limitations

The convergence evidence comes from one deterministic annotation scenario and one maximum-history fit per method. R-hat can identify between-chain disagreement but cannot identify its algorithmic cause. The candidate grid ends at the stored history, and alternative parameterizations, priors, initialization strategies or samplers were not tested. A future benchmark requires a new convergence design and a new prespecified gate.

## Reproducibility and downloads

Rendering reads the stop capsule only. The [analysis entry point](analysis.R) documents the current study source without running it.

- [Current stop capsule](../../results/reference/06_annotation_models/current-stop/README.md)
- [Manifest](../../results/reference/06_annotation_models/current-stop/benchmark_manifest.json)
- [Annotation design](../../results/reference/06_annotation_models/current-stop/annotation_design_summary.csv)
- [Generating alpha](../../results/reference/06_annotation_models/current-stop/true_alpha.csv)
- [Scalar chain draws](../../results/reference/06_annotation_models/current-stop/scalar_chain_draws.csv)
- [Convergence diagnostics](../../results/reference/06_annotation_models/current-stop/convergence_diagnostics.csv)
- [Candidate settings](../../results/reference/06_annotation_models/current-stop/candidate_settings.csv)
- [Recommendations](../../results/reference/06_annotation_models/current-stop/method_recommendations.csv)
- [Checksums](../../results/reference/06_annotation_models/current-stop/checksums.csv)
