---
title: "Study 06: Annotation-informed models"
description: "Qualifications failed; SNP-level annotation benefit established; final benchmark blocked."
---

```{r}
#| echo: false
source("R/benchmark-reporting.R")
capsule <- "results/reference/06_annotation_models/current-stop"
rcsv <- function(name) read.csv(file.path(capsule, name), check.names = FALSE)
design <- rcsv("annotation_design_summary.csv")
fit_status <- rcsv("fit_status.csv")
computation <- rcsv("computational_summary.csv")
```

::: {.callout-important}
## Current status

**Status: In development — formal qualifications failed; final benchmark not
authorized.**

```text
v1 sparse qualification: failed and preserved
v2 identifiable qualification: failed
paired power isolation: completed
package-side hierarchy and transition audits: completed
official SBayesRC multichain parity: blocked by the v0.2.6 seed contract
official SBayesRC single-trajectory descriptive comparison: completed
final benchmark: not authorized
larger n=5000, m≈38000 feasibility experiment: planned, not yet run
```

The official comparison was not wholly blocked: multichain convergence parity
is blocked, while the single-trajectory descriptive comparison completed.
:::

## What Study 06 tests

Study 06 asks whether correctly specified annotations improve causal-marker
prioritization and genetic-signal recovery in BayesRC/SBayesRC relative to
matched BayesR/SBayesR, and whether the learned annotation hierarchy, marker
allocation, and BED versus summary-statistic likelihood routes are reliable.

The main scientific comparison is individual-level BED BayesRC versus retained
block-eigen SBayesRC. The latter uses the prespecified block factorization and
all 100 positive modes in every block at `eigen_prop = 0.995`; this tests the
retained-factor route, block ordering, summary-model semantics, and blockwise
likelihood factorization, not substantial eigenspace truncation.

The immutable v2 identities are:

```text
Specification:      241c15afab8fefc571e38e625130de6e4ab58b958c30cff369e26998ee30fa56
Informative truth:  169d52bef390022a9106d7e61b200493869b40cbb76f1c8d911eebbc80fea1eb
```

## Designs and formal decisions

### V1 sparse design — failed and preserved

V1 used about 37,991 chromosome-1 markers, 1,400 training people, and about 50
expected non-null markers with active weights 0.60/0.30/0.10. That implies only
about 30/15/5 expected markers in the three active components. BED histories
failed convergence and approximate sparse CSR failed its residual-scale
contract. The tracked `current-stop` capsule is immutable failed-development
evidence, not a completed benchmark or a method-performance comparison.

```{r}
#| echo: false
knitr::kable(fit_status, caption = "Historical v1 fit status (historical evidence).")
knitr::kable(computation, digits = 3,
  caption = "Historical v1 recorded computation (descriptive).")
```

V1 remains useful conceptually as `v1_sparse_stress`: a late-stick, prediction,
PIP, active-count, variance, and numerical-integrity stress scenario. It is not
the qualification of precise late-stick annotation recovery.

### V2 identifiable design — formal qualification failed

V2 uses 2,000 people (1,400 training, 600 validation), 1,500 markers in 15
separated 100-marker blocks, heritability 0.50, and
`gamma = c(0, 0.01, 0.1, 1)`. The informative truth has component counts
1,329/84/50/37, hence 171 non-null markers. Informative and marginally matched
uninformative annotation scenarios use intercept, enriched binary, continuous
signal, and null columns. Deterministic truth, annotation-rank, stick-cell,
marker-order, and block audits all passed.

All four registered learned BayesRC/SBayesRC qualification entries completed,
but every entry failed the multichain convergence contract. Informative
later-stick directions and BED/block heritability agreement also failed. This
is a formal qualification failure; the final benchmark remains unauthorized.

## Evidence levels

The rest of this report uses three levels:

- **Established** — supported by converged comparisons, exact tests, or a
  clearly delimited validated descriptive comparison whose scope is stated.
- **Descriptive but useful** — pooled nonconverged summaries, one official
  deterministic trajectory, within-path drift, uncontrolled runtime, or final
  values without histories.
- **Unresolved** — competing explanations cannot yet be separated or required
  diagnostics are unavailable.

“Expected active count” means the sum of marker-specific prior non-null
probabilities. “Realized active count” or “component occupancy” means sampled
marker allocations. They are not interchangeable.

## Current conclusions

### Established

- BayesR/SBayesR baselines converge.
- Fixed-true-alpha BayesRC/SBayesRC converges for the retained quantities, with
  strong cross-chain allocation agreement where complete occupancy histories
  were unavailable.
- Audited stick eligibility, Albert–Chib latent variables, alpha conditionals,
  intercept handling, `sigmaSqAlpha` conditional, and component-probability
  transformation are mathematically and computationally correct.
- The alpha hierarchy converges rapidly when allocations are frozen.
- Fixing `sigmaSqAlpha` and replacing the teaching prior by the
  production-equivalent annotation-variance prior do not restore dynamic joint
  convergence.
- Repeating hierarchy updates improves BED alpha marginal diagnostics, but the
  full occupancy/variance posterior still fails. Repeating allocation sweeps is
  unfavorable.
- The tested complete-state coupling ladder passes exact tiny-state checks but
  has effectively zero Study 06 overlap. Retained lower-replica state is
  insufficient to evaluate exact partial exchange.
- The roughly 0.09–0.10 BED/block heritability offset persists without
  annotation learning, with true alpha fixed, and with every block mode
  retained. It is distinct from learned-alpha feedback and eigen truncation.
- Informative annotations improve causal-marker ranking. Official SBayesRC
  independently reproduces the prioritization benefit on this truth.
- Matched official SBayesRC and `sblr` strongly agree for SNP effects, PIPs,
  ranking, and validation genetic values on the delimited single trajectory.

### Descriptive but useful

The learned-alpha AUPRC and prediction summaries below are descriptive under
non-convergence. Official alpha/architecture summaries, native five-component
results, visible drift, and runtime are single-trajectory descriptive. Pooled
learned BED/block summaries and per-chain implementation comparisons are not
formal posterior validation.

| Fit | Evidence | AUPRC | AUROC | expected active | h2 | effect–truth cor. | validation-g cor. | prediction cor. |
|---|---|---:|---:|---:|---:|---:|---:|---:|
| BayesR BED baseline | formally converged | .3975 | .6943 | 167.8 | .4036 | .8922 | .8932 | .6314 |
| SBayesR block baseline | formally converged | .3270 | .6679 | 178.9 | .5050 | .8602 | .8698 | .6033 |
| learned informative BED | descriptive under non-convergence | .5948 | .8528 | 58.7 | .4162 | .9138 | .9178 | .6545 |
| learned informative block | descriptive under non-convergence | .5407 | .8300 | 131.7 | .5093 | .8847 | .8956 | .6316 |
| fixed-alpha BED | fixed-alpha, converged available contract | .6075 | .8628 | 171.0 | .4273 | .9242 | .9267 | .6616 |
| fixed-alpha block | fixed-alpha, converged available contract | .5993 | .8551 | 179.5 | .5134 | .8909 | .8994 | .6350 |
| BED H20 | nonconverged joint state | .5945 | .8520 | about 55.6 | .4164 | .9138 | .9178 | .6546 |
| block H20 | nonconverged joint state | .5419 | .8309 | about 133.6 | .5090 | .8850 | .8958 | .6320 |

### Unresolved

- Official SBayesRC multichain convergence and quantitative alpha parity.
- Quantitative parity of official `sigma_anno` and `sblr` `sigmaSqAlpha`.
- Latent architecture, component occupancy, and active/component-count
  semantics across implementations.
- Residual-variance and effect/`nbsq` scale contracts.
- BED versus block-eigen variance and heritability calibration.
- Whether substantially larger annotation/allocation information resolves joint
  mixing, or a coordinated posterior-preserving transition is required.
- Whether later-stick annotation coefficients are quantitatively interpretable
  on this dataset.

## Implementation and sampler lessons

Package-side experiments are clearly separated from benchmark-side experiments.
Low-level detail remains authoritative in the linked `sblr` documents.

| Implementation question | Finding | Scientific implication | Status | Authoritative evidence |
|---|---|---|---|---|
| Shared annotation implementation | Scalar BED, CSR, and block-eigen routes call the shared annotation-prior kernel; route adapters were audited directly. | Similar route behavior is not merely assumed from output. | correctness established | [`sblr` hierarchy audit](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_alpha_hierarchy_joint_sampling_audit.md) |
| Stick eligibility/orientation | Stick 1 uses all markers; later sticks use continuers. Truth eligible/continue/stop counts are 1500/171/1329, 171/87/84, 87/37/50. Native and independent references agree. | The observed late-stick sample sizes and continuation convention are known exactly. | correctness established | same |
| Albert–Chib and alpha conditionals | Latent truncation, residualized Gaussian means/variances, and proper intercept handling agree to numerical precision. | No audited conditional-formula defect explains mixing. | correctness established | same |
| `sigmaSqAlpha` conditional | The intercept is excluded; three non-intercepts enter the inverse-chi-square conditional. Current and production-equivalent priors were mapped and exactly checked. | Conditional correctness does not establish joint convergence. | correctness established | same |
| Component probabilities | Sequential stick transformation, flooring, normalization, and orientation passed numerical checks; ordinary states were not floor-controlled. | Numerical flooring is not the Study 06 cause. | correctness established | same |
| BED component traces | Trace markers were resolved in the full-Glist index domain instead of the fitted subset. Boundary resolution/validation was corrected without changing transitions or RNG order. | Retained traces are now reliable; the defect did not cause learned-alpha mixing. | defect corrected | same |
| Frozen allocations | Current prior, production prior, fixed variance, and fixed-alpha reference all pass when allocations are fixed to truth. | Isolated hierarchy conditionals are not the primary failure. | candidate explanation ruled out | same; decision C |
| Fixed `sigmaSqAlpha` | Dynamic fits still fail on both routes. | A centred annotation-variance funnel is not the primary bottleneck. | candidate explanation ruled out | same |
| Production-prior ablation | The production-equivalent prior does not restore dynamic convergence. | Prior-tail choice alone is insufficient. | candidate explanation ruled out | same |
| S1/H5/H20/A5/A20 | H20 improves BED alpha marginals, but joint occupancy/variance still fails; allocation repetition worsens behavior. | More conditional work alone is not the missing global transition. | candidate explanation ruled out | [`sblr` K4 audit](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_allocation_hierarchy_kernel_composition.md) |
| Coupling tempering | Tiny detailed balance/endpoints pass, but Study 06 accepts 0/2,400 exchanges with replicas hundreds of log-density units apart. | Algebraically correct tempering is an ineffective bridge for this posterior. | candidate explanation ruled out | [`sblr` T4 screen](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_bed_coupling_tempering_screen.md) |
| Partial exchange | Exact formulas exist, but lower-replica complete states were not retained; no proposal was validated or rejected. Aggregate active-count mismatch is not the sole explanation. | The candidate could not be evaluated from existing state. | unresolved contract difference | [`sblr` F6 audit](https://github.com/psoerensen/sblr/blob/master/docs/dev/study06_partial_exchange_feasibility.md) |
| BED/block variance calibration | The h2 offset persists without annotation, with fixed alpha, and with 100/100 modes. | Route calibration is distinct from alpha feedback and eigen truncation. | unresolved contract difference | paired isolation; package audits |
| Block runtime/storage | Operators are constructed before MCMC, BED is not reread during MCMC, and immutable block operators are shared. Recorded block runs were faster; analytical and observed process memory differ. | The retained block route is computationally attractive, but recorded times are not controlled benchmarks. | performance observation | [`sblr` block contract](https://github.com/psoerensen/sblr/blob/master/docs/dev/blr_block_eigen_contract.md) |
| Official/`sblr` contracts | Residual variance, effect/`nbsq` scale, p1 orientation, pi updates, active-count definitions, component counts, `sigma_anno`/`sigmaSqAlpha`, and wrapper labels differ or remain unmapped. | Strong SNP agreement does not validate latent architecture. | unresolved contract difference | [single-trajectory result](../../docs/dev/study06_gctb_single_trajectory_result.md) |

## Official SBayesRC comparison

The external implementation is **official SBayesRC R implementation v0.2.6,
source SHA `b95d3fcbad8ff358290922a58fff893439296138`**. “Official SBayesRC” below
means that implementation; it is not used as a generic synonym for GCTB.

### Strongest matched comparisons

| Comparison | PIP Pearson/Spearman | effect Pearson | validation-g correlation | top-50/top-100 | AUPRC |
|---|---:|---:|---:|---:|---:|
| Official D1 vs learned block `sblr` | .957/.964 | .998 | .998 | 46/86 | .557 vs .541 |
| Official D1 vs learned BED `sblr` | .933/.953 | .956 | .960 | 39/77 | .557 vs .595 |

Official D1 resembled every saved learned `sblr` chain, rather than one isolated
occupancy regime: block PIP correlations were .946–.965 and BED correlations
.925–.936; effect correlations were .99831–.99833 and .95483–.95664.

Official annotation benefit was large on the identical truth: D0 versus D1
AUPRC was .320 versus .557, AUROC .664 versus .842, validation genetic-value
correlation .869 versus .896, and phenotype prediction .604 versus .634.

> The `sblr` SBayesRC implementation compares well with official SBayesRC for
> SNP effects, PIPs, causal prioritization, validation prediction, and practical
> runtime. Agreement is weaker or unresolved for latent architecture,
> later-stick alpha coefficients, component occupancy, and annotation-variance
> parameters.

Official SBayesRC v0.2.6 does not connect its public seed argument to the native
RNG streams, so multichain convergence parity remains unavailable. The official
results are single-trajectory descriptive and must not be called converged.

## Alpha and annotation variance

Only unambiguously mapped values are compared. The v2 non-intercept truth for
stick 1 is enriched 1.60, continuous .30, null 0. The calibrated truth
intercept is retained in the simulation object but is not duplicated in the
compact result documents; no value is inferred here.

| Quantity, stick 1 | Truth | Official D1 single trajectory | `sblr` learned BED pooled | `sblr` learned block pooled | Fixed-alpha |
|---|---:|---:|---:|---:|---:|
| intercept | available in truth object | -2.406 | not compactly reported here | not compactly reported here | fixed to registered truth |
| enriched | 1.600 | 1.507 | 1.897 | 1.642 | 1.600 |
| continuous | .300 | .586 | .638 | .622 | .300 |
| null | 0 | -.156 | -.042 | -.371 | 0 |
| evidence label | truth | single-trajectory descriptive | pooled nonconverged | pooled nonconverged | fixed-alpha |

Official first-stick enriched and continuous directions agree with truth and
every saved learned `sblr` chain; its null coefficient is comparatively small.
Quantitative coefficient equality is not established. BED H20 put all 12 alpha
marginals within the diagnostic contract, but it remained a nonconverged joint
state; the compact package evidence reports diagnostics rather than a complete
mapped mean table.

Later-stick values are not quantitatively interpretable. Official p3 intercept
drifted about 1.85 posterior SD and p3 continuous about -1.06 SD; `sblr`
later-stick alpha and occupancy also mixed poorly.

Official final `sigma_anno` values were 1.976/1.043/4.803 for D1 and
3.480/.384/.413/.605 for D2. `sblr` informative selected-window
`sigmaSqAlpha` means were about 2.14/1.90/3.14 BED and 2.04/1.54/1.79 block;
these are pooled nonconverged posterior summaries. In production-prior dynamic
fits, BED stick-specific `sigmaSqAlpha` diagnostics passed while block sticks
1/2 failed; this did not rescue the joint posterior. Fixed-variance conditions
set `sigmaSqAlpha = 1` and are ablations, not posterior estimates.

Official retained `sigmaSqAlpha` histories are not exposed. Official
`sigma_anno` and `sblr` `sigmaSqAlpha` must not be assumed identical without a
source-contract audit, and final official values cannot establish posterior
mean, uncertainty, drift, ESS, or chain parity.

## Runtime observations

| Method | Trajectories/chains | Iterations | Wall time |
|---|---:|---:|---:|
| Official D0 SBayesR | 1 | 9,000 | 4.61 s |
| Official D1 matched SBayesRC | 1 | 9,000 | 39.28 s |
| Official D2 native SBayesRC | 1 | 9,000 plus tuning | 82.03 s |
| `sblr` informative block SBayesRC | 4 | 9,000 each | 54.74 s |
| `sblr` informative BED BayesRC | 4 | 9,000 each | 129.24 s |

Official runs used one OpenMP and one BLAS thread. `sblr` used four registered
chains and may use parallel execution; retained tracing, diagnostics, package
builds, and process overhead differ. This is descriptive, not a controlled
performance benchmark.

> The `sblr` block-eigen route is computationally competitive. Four `sblr`
> block chains required only about 1.4 times the wall time of one official
> matched trajectory in the recorded runs.

## Experiment timeline

| Phase | Question and route/repository | Result | Implication | Evidence |
|---|---|---|---|---|
| 1. v1 sparse qualification | Sparse BED/CSR annotation qualification | Failed; preserved | Too sparse for primary late-stick qualification; CSR fidelity confounded | formal qualification |
| 2. v2 identifiable design | Increase active/stick information | Deterministic gates passed | Accidental empty truth cells ruled out | design/protocol |
| 3. v2 qualification | Learned BED/block BayesRC/SBayesRC | All fits completed; all failed convergence | Final benchmark blocked | formal qualification |
| 4. paired power isolation | Baseline, learned, shuffled, fixed-alpha | Baselines/fixed pass; learned fails; ranking benefit large | Alpha/allocation feedback isolated | diagnostic experiment |
| 5. mathematical audit | Shared conditionals in `sblr` | Exact checks pass | Formula defect ruled out | correctness established |
| 6. frozen allocations | Hierarchy with truth allocations | Converges rapidly | Isolated hierarchy ruled out | exact/converged diagnostic |
| 7. variance ablations | Fixed variance and production prior | Dynamic failure persists | Variance funnel/prior not primary | diagnostic experiment |
| 8. trace correction | BED selected-marker index domain | Defect corrected; RNG/transitions unchanged | Reporting defect not mixing cause | defect corrected |
| 9. kernel schedules | S1/H5/H20/A5/A20 | H20 helps alpha; joint fails; A repeats worsen | Recomposition alone insufficient | K4 diagnostic |
| 10. tempering | Tiny exact plus Study 06 BED | Algebra passes; 0 exchanges | Tested ladder ineffective | T4 diagnostic |
| 11. partial exchange | Offline retained-state audit | Exact assessment unavailable | Candidate neither validated nor rejected | F6 unresolved |
| 12. official export/API | Exact Study 06 official input and seed test | Export valid; independent chains blocked | GCTB-P5 remains | formal decision |
| 13. official single path | D0/D1/D2, 9,000 iterations | Strong SNP agreement/benefit; architecture drift | Descriptive parity completed | single-trajectory descriptive |
| 14. documentation consolidation | This task | Evidence hierarchy normalized | One authoritative reading path | documentation decision |
| 15. larger feasibility | `n=5000`, `m≈38000` | Planned, not run | Tests information-scale hypothesis | unresolved roadmap |

## Interpretation and next phases

The accumulated evidence separates practical SNP prioritization from latent
posterior reliability. Annotation information clearly helps find causal
markers on this truth, and two implementations produce nearly identical SNP
signals. Neither fact establishes joint learned-alpha convergence or validates
component architecture.

The next scientific phase is a **planned, not yet run** one-replicate
information-scale feasibility study with `n = 5000`, `m ≈ 38000`, four
independent `sblr` chains, and alpha, `sigmaSqAlpha`, occupancy, expected and
realized active counts, genetic/residual variance, heritability, PIP ranking,
and truth-recovery checks. Replicated validation follows only if it succeeds.
An optional fixed-state official/`sblr` latent-contract audit can proceed
independently. Official multichain parity awaits a seedable official release.

## Navigation and reproducibility

Start with the [Study 06 README](README.md), then use the
[authoritative evidence synthesis](../../docs/dev/study06_annotation_inference_evidence_synthesis.md)
for the complete ledger. The [documentation inventory](../../docs/dev/study06_documentation_inventory.md)
maps every tracked Study 06 file and package-side evidence source.

The experiment-specific authoritative records are the
[v1 result](../../docs/dev/study06_annotation_qualification_result.md),
[v2 design](../../docs/dev/study06_v2_design.md),
[v2 result](../../docs/dev/study06_v2_qualification_result.md),
[paired isolation](../../docs/dev/study06_v2_power_isolation_result.md),
[GCTB-P5 result](../../docs/dev/study06_gctb_parity_result.md), and
[official single-trajectory result](../../docs/dev/study06_gctb_single_trajectory_result.md).
This report reads only the unchanged v1 capsule and never launches a sampler.
