---
title: "Study 06: Annotation-informed models"
description: "Implemented validation design; qualification failed."
---

```{r}
#| echo: false
source("R/benchmark-reporting.R")
capsule <- "results/reference/06_annotation_models/current-stop"
rcsv <- function(name) read.csv(file.path(capsule, name), check.names = FALSE)
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"),
  simplifyVector = TRUE)
design <- rcsv("annotation_design_summary.csv")
true_alpha <- rcsv("true_alpha.csv")
diagnostics <- rcsv("convergence_diagnostics.csv")
candidates <- rcsv("candidate_settings.csv")
recommendations <- rcsv("method_recommendations.csv")
fit_status <- rcsv("fit_status.csv")
computation <- rcsv("computational_summary.csv")
```

::: {.callout-important}
**Status: In development — qualification failed.** The existing capsule is a
**Prespecified convergence stop — partial evidence only**. It contains no final
BayesR/BayesRC or SBayesR/SBayesRC performance comparison.
:::

## Scientific question

Study 06 asks whether correctly specified informative annotations improve
marker prioritization and genetic-signal recovery under BayesRC/SBayesRC,
relative to marginally matched BayesR/SBayesR, without corresponding gains
under uninformative annotations.

## Study design

The implemented design uses 2,000 chromosome-1 samples and 37,991 ordered QC
markers from qgdata. A deterministic 70/30 split supplies 1,400 training and
600 test samples. Informative and uninformative annotation scenarios each use
five replicates, target heritability 0.30, and 50 expected active markers under
mixture variances 0, 0.01, 0.1, and 1. The planned final grid has four matched
methods, 40 fits, and 160 chains.

## Annotation construction

Every marker has an intercept, enriched binary annotation, standardized
continuous-signal annotation, and null annotation, in that order. Construction
is deterministic and marker aligned. The informative scenario links annotation
signal to causal allocation; the uninformative scenario preserves the same
annotation architecture without causal enrichment.

```{r}
#| echo: false
knitr::kable(design, digits = 4)
knitr::kable(true_alpha, digits = 4,
  caption = "Generating annotation-by-stick coefficients in the stop capsule.")
```

## Methods

The final comparison is BED BayesR versus BED BayesRC and CSR SBayesR versus
CSR SBayesRC (`annotation_model = "annotation_probit_stick"`). Baselines and
annotation-aware counterparts use matched marginal mixture assumptions. The
CSR route uses training-only summary statistics and sparse LD.

## Estimands

The primary annotation estimand is the same draw-wise marker component prior
for BED and CSR: retained annotation coefficients are combined with the ordered
annotation matrix using the model's probit stick transformation, and only then
summarized across draws. Posterior mean alpha, final alpha, final marker-prior
state, input probabilities, and posterior marker allocation probabilities are
kept distinct.

Marker PIP and causal rank measure prioritization. Prediction, genetic-value
and effect recovery measure signal recovery; variance components, runtime and
convergence provide supporting diagnostics. No unavailable fit quantity is
reconstructed from a final state or a posterior mean.

## Qualification gate

Four entries must qualify before the final grid can run: informative replicate
1 and uninformative replicate 1 for BED BayesRC and CSR SBayesRC. Each permits
at most 9,000 iterations and four chains. Candidate burn-ins are 1,000, 2,000,
and 3,000; retained windows are 2,000, 4,000, and 6,000. Every required scalar
must satisfy R-hat ≤ 1.01, bulk ESS ≥ 400, tail ESS ≥ 400, and relative MCSE ≤
0.05. Thresholds cannot be weakened automatically.

## Current partial evidence

The earlier maximum-history evidence covered one informative replicate for the
two annotation-aware methods. Both fits completed, but no candidate window met
all registered thresholds. The limiting quantity was
`alpha:continuous_signal:component_0_stick`.

```{r}
#| echo: false
knitr::kable(fit_status)
knitr::kable(computation, digits = 3)
knitr::kable(candidates, digits = 4)
knitr::kable(recommendations, digits = 4)
```

This is trustworthy evidence for the recorded stop decision, but it does not
cover the new four-entry qualification gate and is not evidence about method
performance.

## Qualification result

The implemented four-entry qualification was run on 2026-08-04 and failed.
Both BED BayesRC histories completed but no supported candidate window passed
all required diagnostics. Both CSR SBayesRC entries stopped in the pinned
native backend with an invalid projected residual scale. The 40-fit final grid
was therefore not launched. Full per-entry, per-window, and per-quantity
evidence is recorded in the
[qualification result](../../docs/dev/study06_annotation_qualification_result.md).

These failures are sampler qualification evidence, not method-performance
results. The ignored semantic checkpoints and diagnostic tables remain local;
this report continues to read only the tracked `current-stop` capsule.

## Why the final benchmark is not complete

A qualifying decision artifact does not exist for the implemented
specification. The recorded decision is failed, and the final runner rejects
performance execution. Completed histories can be reused only when their
semantic identity, retained traces, method, scenario, burn-in, and retained
window match exactly.

## Required next runs

The next action is a focused `sblr` investigation of BayesRC mixing and the CSR
projected-residual failure, followed by the unchanged four-entry qualification.
Only after all four entries pass may a separate explicit final command launch
the prespecified final grid.

## Validation boundaries

The stop capsule does not support conclusions about annotation accuracy,
enrichment recovery, marker selection, prediction, or superiority of BED or
CSR routes. Qualification is a gate, not a performance result. The implemented
software and deterministic tests do not substitute for scientific fits.

## Reproducibility and downloads

This page reads the unchanged `current-stop` capsule only; it does not source
the study workflow, construct annotations, load checkpoints, or call a sampler.

- [Exact analysis](analysis.R)
- [Study specification](spec.R)
- [Annotation design](annotation-design.R)
- [Reusable annotation template](../../inst/templates/annotation-analysis.R)
- [Audit](../../docs/dev/study06_annotation_audit.md)
- [Completion plan](../../docs/dev/study06_annotation_completion_plan.md)
- [Qualification result](../../docs/dev/study06_annotation_qualification_result.md)
- [Partial-evidence capsule](../../results/reference/06_annotation_models/current-stop/README.md)
- [Manifest](../../results/reference/06_annotation_models/current-stop/benchmark_manifest.json)
- [Convergence diagnostics](../../results/reference/06_annotation_models/current-stop/convergence_diagnostics.csv)
- [Checksums](../../results/reference/06_annotation_models/current-stop/checksums.csv)
- [`sblr` annotation interface documentation](https://psoerensen.github.io/sblr/reference/stblr_csr_annot.html)
