---
title: "Study 06: Single-trait LD operator benchmark"
description: "Development benchmark separating block boundaries, block-eigen reconstruction, spectral filtering, and summary-statistics versus BED effects."
---

```{r}
#| echo: false
root <- file.path("results", "reference", "06_ld_operator")
conv_dir <- file.path(root, "st-ld-operator-convergence-development-v1")
bench_dir <- file.path(root, "st-ld-operator-five-replicate-development-v1")
required <- c(file.path(conv_dir, "benchmark_manifest.json"),
  file.path(bench_dir, "benchmark_manifest.json"))
if (!all(file.exists(required)))
  stop("Study 06 frozen reference capsules are unavailable.")
conv_manifest <- jsonlite::read_json(required[1], simplifyVector = TRUE)
bench_manifest <- jsonlite::read_json(required[2], simplifyVector = TRUE)
read_table <- function(dir, name)
  utils::read.csv(file.path(dir, name), stringsAsFactors = FALSE)
block_candidates <- read_table(conv_dir, "block_design_candidates.csv")
selected_block <- read_table(conv_dir, "selected_block_design.csv")
equivalence <- read_table(conv_dir, "operator_equivalence_summary.csv")
hard_candidates <- read_table(conv_dir, "hard_filter_candidates.csv")
selected_hard <- read_table(conv_dir, "selected_hard_filter.csv")
synthetic_filter <- read_table(conv_dir, "synthetic_filter_validation.csv")
block_diagnostics <- read_table(conv_dir,
  "block_level_operator_diagnostics.csv")
lw_audit <- read_table(conv_dir, "ridge_lw_audit.csv")
operator_pilot <- read_table(conv_dir,
  "one_replicate_operator_pilot.csv")
recommendations <- read_table(conv_dir, "method_recommendations.csv")
fit_status <- read_table(bench_dir, "fit_status.csv")
prediction <- read_table(bench_dir, "prediction_metrics.csv")
recovery <- read_table(bench_dir, "parameter_recovery_summary.csv")
sbayesr <- read_table(bench_dir, "sbayesr_diagnostic_evidence.csv")
marker_agreement <- read_table(bench_dir, "marker_effect_agreement.csv")
paired <- read_table(bench_dir, "paired_comparison_summary.csv")
conv_validation <- read_table(bench_dir, "convergence_validation_summary.csv")
runtime <- read_table(bench_dir, "computational_summary.csv")
complete <- isTRUE(identical(bench_manifest$benchmark_status, "complete")) &&
  isTRUE(bench_manifest$successful_fit_count == 60L) &&
  isTRUE(bench_manifest$failed_fit_count == 0L)
```

::: {.callout-warning .callout-development}
## Development benchmark

This package-specific benchmark uses five paired simulations per architecture.
It provides descriptive evidence and operator-equivalence validation, not a
universal method ranking, universal convergence result, or resolution of the
known SBayesR heritability observation.
:::

## Scientific questions

The benchmark separates four contrasts in order:

1. runtime-matched block CSR versus unfiltered block eigen;
2. full CSR versus block CSR;
3. hard- or fixed-ridge-filtered versus unfiltered block eigen;
4. summary-statistics operators versus individual-level BED.

Block-eigen configurations are operator representations, not distinct
statistical models. `S` in SBayesC and SBayesR denotes summary statistics and
is unrelated to sampled MAF-S models. Annotation-informed and multivariate
fits are excluded.

## Implementations and data

Sparse homogeneous simulations use matched BayesC/SBayesC fits; sparse
variance-mixture simulations use matched BayesR/SBayesR fits. Each simulation
has 5,000 samples, 50 causal markers, target realized heritability 0.30, the
canonical post-QC chromosome-1 markers, and the deterministic 70/30 split.
Summary scores, frequencies, phenotype scaling, full CSR LD, and every block
operator use training individuals only.

The six configurations are individual-level BED, full sparse-LD CSR,
runtime-matched block CSR, unfiltered block eigen, hard-filtered block eigen,
and scale-aware fixed-ridge block eigen.

## Block construction and operator-equivalence gate

```{r}
#| echo: false
knitr::kable(block_candidates, digits = 4)
knitr::kable(selected_block, caption = "Frozen block design")
knitr::kable(equivalence, digits = 6,
  caption = "Runtime-matched CSR versus unfiltered block-eigen gate")
```

All block representations originate from the same installed builder's
cross-product matrix after marker alignment, standardization, symmetry
handling, and float32 runtime packing. Cross-block LD is zero. If the gate did
not pass, sampler benchmarking was not permitted.

The exact no-shrink public path is
`eigen_filter = "ridge_fixed", eigen_eta = 0`. Hard truncation with
`eigen_tau = 0` is not unfiltered because the installed implementation applies
a 0.01 floor. Negative eigenvalues are retained by the no-shrink
reconstruction and dropped by hard truncation when below its effective
threshold.

## Synthetic filtering validation

```{r}
#| echo: false
knitr::kable(synthetic_filter, digits = 6)
```

The controlled spectra validate zero-ridge reconstruction, active hard
truncation, fixed ridge, and finite positive diagonals independently of the
qgdata eigenvalue distribution.

## qgdata eigenvalue distribution and filter activity

```{r}
#| echo: false
knitr::kable(hard_candidates, digits = 5)
knitr::kable(selected_hard, caption = "Frozen hard-filter setting")
eigen_summary <- data.frame(
  block_count = nrow(block_diagnostics),
  minimum_eigenvalue = min(block_diagnostics$minimum_eigenvalue),
  maximum_eigenvalue = max(block_diagnostics$maximum_eigenvalue),
  negative_eigenvalue_count = sum(block_diagnostics$negative_eigenvalue_count),
  original_rank = sum(block_diagnostics$original_rank),
  retained_rank = sum(block_diagnostics$retained_rank))
knitr::kable(eigen_summary, digits = 5)
```

The public hard route is frozen at `eigen_tau = 0.10`. It remains valid when
all eigenvalues are retained; such blocks are labelled `effective_no_op`.
Synthetic controlled-spectrum tests separately verify that active truncation
removes the expected eigenvalues. Filtering changes or regularizes the
numerical operator but does not reduce the packed dense runtime storage.
Ledoit--Wolf is reported as a deterministic sensitivity audit. Because its
qgdata weights were near complete diagonal shrinkage, the main ridge route is
the frozen 1% `ridge_fixed` policy rather than `ridge_lw`.

```{r}
#| echo: false
knitr::kable(lw_audit, caption = "Ledoit--Wolf sensitivity audit",
  digits = 5)
```

## One-replicate operator pilot

```{r}
#| echo: false
knitr::kable(operator_pilot, digits = 4)
```

This pilot checked output dimensions, scaling, prediction, marker-effect
agreement, variance summaries, and warnings before maximum-history fitting.
It was a software gate, not a prediction-tuning exercise.

## Convergence recommendations

```{r}
#| echo: false
knitr::kable(recommendations, digits = 4)
```

BED and full-CSR settings come from the frozen Study 04 capsule. New block
configurations were assessed with four 3,000-draw maximum-history chains and
the unchanged R-hat, bulk/tail ESS, relative-MCSE, finite-domain, and chain
identity thresholds.

## Five-replicate completion

The capsule reports `r if (complete) "60/60 validated four-chain fits" else "an incomplete grid"`. A failure remains in the denominator and prevents
promotion.

## Prediction

```{r}
#| echo: false
prediction_summary <- aggregate(value ~ architecture + configuration + metric,
  prediction, function(x) c(mean = mean(x), sd = stats::sd(x),
    median = stats::median(x), minimum = min(x), maximum = max(x)))
prediction_summary <- cbind(prediction_summary[1:3],
  as.data.frame(prediction_summary$value))
knitr::kable(prediction_summary, digits = 4)
```

Held-out outcomes include phenotype correlation, RMSE and calibration slope,
plus genetic-value correlation and RMSE.

## Parameter recovery

```{r}
#| echo: false
knitr::kable(recovery, digits = 4)
```

Five-replicate interval coverage is a coarse observed proportion: each miss
changes it by 20 percentage points. Quantities without a defensible generating
truth are labelled posterior behaviour rather than recovery.

## SBayesR variance-component evidence

```{r}
#| echo: false
s_summary <- aggregate(cbind(vgs_posterior_mean, ves_posterior_mean,
  vbs_posterior_mean, pis_posterior_mean, vld_posterior_mean,
  vle_posterior_mean, benchmark_heritability) ~ configuration,
  sbayesr, mean, na.rm = TRUE)
knitr::kable(s_summary, digits = 4)
```

These fields localize the previously observed positive SBayesR heritability
bias descriptively. They do not change the installed implementation or the
benchmark heritability definition, and five simulations do not establish a
cause.

## Marker-effect agreement

```{r}
#| echo: false
knitr::kable(marker_agreement, digits = 4)
```

## Paired operator comparisons

```{r}
#| echo: false
for (level in 1:4) {
  cat("\n\n### Level ", level, "\n\n", sep = "")
  print(knitr::kable(paired[paired$interpretation_level == level, ],
    digits = 4))
}
```

Level 1 is the software/operator-equivalence comparison. Level 2 quantifies
the selected block representation relative to full sparse LD. Level 3
quantifies spectral filtering. Level 4 reflects broader likelihood, scaling,
LD, and implementation differences from BED and is not an equality test.

## Main-run convergence validation

```{r}
#| echo: false
knitr::kable(conv_validation, digits = 4)
```

Support across five simulations is not universal convergence validation. An
aggregate pass never hides an estimand-level failure.

## Runtime and storage

```{r}
#| echo: false
runtime_summary <- aggregate(cbind(model_fitting_time, total_time,
  output_size_bytes) ~ architecture + configuration,
  runtime, function(x) mean(x, na.rm = TRUE))
knitr::kable(runtime_summary, digits = 2)
```

Preprocessing, LD construction, sampling, prediction, metric generation,
output size, and available memory estimates are kept in separate columns in
the download.

## Limitations

This is a development benchmark with five paired simulations and one frozen
block/filter design. It benchmarks current installed `sblr` behaviour, does
not cover annotation models or MTBLR, does not establish universal numerical
or scientific superiority, and does not resolve the SBayesR observation.

## Downloads and provenance

- [Convergence capsule README](../../results/reference/06_ld_operator/st-ld-operator-convergence-development-v1/README.md)
- [Five-replicate capsule README](../../results/reference/06_ld_operator/st-ld-operator-five-replicate-development-v1/README.md)
- [Paired replicate differences](../../results/reference/06_ld_operator/st-ld-operator-five-replicate-development-v1/paired_replicate_differences.csv)
- [Operator diagnostics](../../results/reference/06_ld_operator/st-ld-operator-five-replicate-development-v1/block_level_operator_diagnostics.csv)
- [Reproduction script](../../results/reference/06_ld_operator/st-ld-operator-five-replicate-development-v1/reproduce.R)
- [Canonical checksums](../../results/reference/06_ld_operator/st-ld-operator-five-replicate-development-v1/checksums.csv)

The report reads only frozen capsule files. It does not load qgdata, construct
LD, decompose matrices, simulate phenotypes, or call samplers.
