---
title: "Study 06: Retained low-rank LD operators"
description: "Current five-replicate LD-operator validation and compatibility audit."
---

```{r}
#| echo: false
capsule <- "results/reference/06_ld_operator/current"
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"),
  simplifyVector = TRUE)
compatibility <- jsonlite::read_json(file.path(capsule,
  "compatibility_manifest.json"), simplifyVector = TRUE)
prediction <- read.csv(file.path(capsule, "prediction_metrics.csv"),
  check.names = FALSE)
paired <- read.csv(file.path(capsule, "paired_comparison_summary.csv"),
  check.names = FALSE)
checkpoint <- read.csv(file.path(capsule, "checkpoint_validation.csv"),
  check.names = FALSE)
rank <- read.csv(file.path(capsule, "low_rank_fit_block_diagnostics.csv"),
  check.names = FALSE)
runtime <- read.csv(file.path(capsule, "computational_summary.csv"),
  check.names = FALSE)
```

::: {.callout-warning}
This is a five-replicate development benchmark of operator implementations,
not a comparison of different BLR prior models or a universal performance
ranking.
:::

## Scope and provenance

Two architectures, five paired replicates, and six configurations produced 60
validated four-chain fits: packed BED, full CSR, block CSR, near-full retained
low rank, 0.999 retained mass, and 0.995 retained mass. Reconstructed-dense
eigen fits were excluded from the primary grid.

The numerical fits retain their truthful `sblr` source provenance at
`r manifest$numerical_fit_source_sha`. Compatibility with the current refresh
SHA `r manifest$refresh_validation_sha` was established without sampler reruns:
all resolved BayesC and BayesR priors are identical and all compiled scalar
sampler and retained-low-rank transition sources are byte-identical.

## Deterministic and numerical validation

The `block_low_rank_v1` contract stores retained factors and transformed scores. Its global
projected residual-variance identity is

$$
SSE = yy - \sum_b w_b^\top w_b + \sum_b r_b^\top r_b.
$$

Deterministic Q/w identities, rank selection, score transforms, residual
updates, and projected SSE passed. The maximum observed residual drift in the
60-fit grid was `r format(max(checkpoint$maximum_residual_drift, na.rm = TRUE),
scientific = TRUE, digits = 4)`.

The whole-marker retained ranks were 37,991/37,991 near-full,
37,839/37,991 at 0.999, and 37,231/37,991 at 0.995. Retained rank is not assumed
to guarantee a speed or memory advantage; measured timing and storage determine
the practical trade-off.

## Prediction

```{r}
#| echo: false
shown <- prediction[prediction$metric %in% c(
  "phenotype_prediction_correlation", "genetic_value_correlation",
  "prediction_rmse", "prediction_calibration_slope"), ]
cols <- intersect(c("architecture", "replicate", "configuration", "metric",
  "value"), names(shown))
knitr::kable(utils::head(shown[cols], 48), digits = 4)
```

## Paired operator contrasts

```{r}
#| echo: false
shown <- paired[paired$metric %in% c("phenotype_prediction_correlation",
  "genetic_value_correlation", "prediction_rmse", "heritability"), ]
knitr::kable(shown, digits = 4)
```

Block CSR versus full CSR describes omitted cross-block LD. Near-full retained
low rank versus block CSR isolates retained-factor implementation as closely as
the frozen operators allow. Near-full versus 0.999 and 0.995 describes spectral
retention. BED versus summary data is a broader likelihood and representation
comparison, not an equivalence requirement.

## Current decision

The scalar result is **READY FOR MTBLR LOW-RANK EXTENSION**: deterministic
identities and projected SSE are valid, chains are supported under the frozen
recommendations, retention sensitivity is not pathological, and numerical
stability, runtime, and storage are characterized. This does not itself validate
an MT implementation.

## Reproducibility

- [Current capsule](../../results/reference/06_ld_operator/current/README.md)
- [Manifest](../../results/reference/06_ld_operator/current/benchmark_manifest.json)
- [Compatibility evidence](../../results/reference/06_ld_operator/current/compatibility_manifest.json)
- [Prior compatibility](../../results/reference/06_ld_operator/current/prior_compatibility.csv)
- [Paired comparisons](../../results/reference/06_ld_operator/current/paired_replicate_differences.csv)
- [Checksums](../../results/reference/06_ld_operator/current/checksums.csv)

Earlier reconstructed-dense and versioned low-rank benchmark presentations remain
available through Git history, not as duplicate active result trees.
