---
title: "Study 06 v2: Retained low-rank LD operator validation"
description: "Five-replicate validation of the retained low-rank scalar LD operator in sblr 0.2.0."
---

```{r}
#| echo: false
repo_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
while (!file.exists(file.path(repo_root, "DESCRIPTION"))) {
  parent <- dirname(repo_root)
  if (identical(parent, repo_root))
    stop("Cannot locate the sblrbench repository root.")
  repo_root <- parent
}
root <- file.path(repo_root, "results", "reference", "06_ld_operator")
conv_dir <- file.path(root, "st-low-rank-operator-convergence-development-v2")
bench_dir <- file.path(root, "st-low-rank-operator-five-replicate-development-v2")
required <- c(file.path(conv_dir, "benchmark_manifest.json"),
  file.path(bench_dir, "benchmark_manifest.json"))
if (!all(file.exists(required)))
  stop("Study 06 v2 frozen reference capsules are unavailable.")
conv_manifest <- jsonlite::read_json(required[1L], simplifyVector = TRUE)
bench_manifest <- jsonlite::read_json(required[2L], simplifyVector = TRUE)
read_table <- function(dir, name)
  utils::read.csv(file.path(dir, name), stringsAsFactors = FALSE)
deterministic <- read_table(conv_dir, "deterministic_identity_summary.csv")
block_diagnostics <- read_table(conv_dir, "low_rank_block_diagnostics.csv")
pilot <- read_table(conv_dir, "one_replicate_operator_pilot.csv")
recommendations <- read_table(conv_dir, "method_recommendations.csv")
fit_status <- read_table(bench_dir, "fit_status.csv")
prediction <- read_table(bench_dir, "prediction_metrics.csv")
recovery <- read_table(bench_dir, "parameter_recovery_summary.csv")
parameters <- read_table(bench_dir, "parameter_estimates.csv")
markers <- read_table(bench_dir, "marker_effect_metrics.csv")
marker_agreement <- read_table(bench_dir, "marker_effect_agreement.csv")
variable <- read_table(bench_dir, "variable_selection_metrics.csv")
paired <- read_table(bench_dir, "paired_comparison_summary.csv")
convergence <- read_table(bench_dir, "convergence_validation_summary.csv")
runtime <- read_table(bench_dir, "computational_summary.csv")
checkpoint_validation <- read_table(bench_dir, "checkpoint_validation.csv")
low_rank <- read_table(bench_dir, "low_rank_fit_block_diagnostics.csv")
low_rank$block_id <- ave(low_rank$block_start,
  interaction(low_rank$architecture, low_rank$replicate,
    low_rank$configuration, drop = TRUE), FUN = seq_along)
sbayesr <- read_table(bench_dir, "sbayesr_diagnostic_evidence.csv")
readiness <- readLines(file.path(bench_dir, "readiness_decision.txt"),
  warn = FALSE)[1L]
complete <- isTRUE(bench_manifest$successful_fit_count == 60L) &&
  isTRUE(bench_manifest$failed_fit_count == 0L)
checkpoint_validation$seconds_per_chain_iteration <-
  checkpoint_validation$mcmc_wall_seconds /
  (checkpoint_validation$chain_count * checkpoint_validation$total_iterations)
summarize_metric <- function(x) {
  groups <- split(x, interaction(x$architecture, x$configuration, x$metric,
    drop = TRUE))
  do.call(rbind, lapply(groups, function(z) data.frame(
    architecture = z$architecture[1L], configuration = z$configuration[1L],
    metric = z$metric[1L], replicate_count = nrow(z), mean = mean(z$value),
    sd = if (nrow(z) > 1L) stats::sd(z$value) else NA_real_,
    median = stats::median(z$value), minimum = min(z$value),
    maximum = max(z$value), stringsAsFactors = FALSE)))
}
```

::: {.callout-warning .callout-development}
## Development evidence

This is a package-specific development benchmark based on five paired
simulations per architecture. It is descriptive evidence, not a universal
method ranking or a universal convergence result.
:::

## Executive summary

Study 06 v2 evaluates whether retained factors produce coherent scalar BLR
inference and prediction, and whether the canonical 99.5% positive-eigenvalue
mass rule is stable relative to 99.9% and near-full positive rank. The final
implementation decision is **`r readiness`**.

Across all five paired replicates, near-full retained factors closely tracked
runtime-matched block CSR. The canonical 0.995 route also remained close to
near-full and 0.999 retention for prediction and marker effects. BED and full
CSR were broadly comparable, with the larger implementation difference in the
sparse-mixture architecture. These are descriptive five-replicate findings.

## Historical Study 06 v1

The earlier capsule is retained as **Historical reconstructed-dense Study 06
v1**. It used `sblr` 0.1.2 and reconstructed packed dense block matrices. V2
uses pinned `sblr` 0.2.0 source commit
`r bench_manifest$sblr_source_sha`, retains (Q_b) factors, and stores reduced
residual state. V1 and v2 results are therefore separate evidence rather than
a homogeneous paired dataset.

## Retained low-rank contract

Every eigen-operator fit explicitly sets `representation = "low_rank"` and an
`eigen_prop`. The contract is `block_low_rank_v1`. No v2 scientific fit uses
the reconstructed-dense sampler or the historical hard/ridge filter controls.

For block (b), the sampler uses

\[
\widetilde A_b = Q_b^\top Q_b,\qquad
\widetilde s_b = Q_b^\top w_b,\qquad
r_b = w_b-Q_b\beta_b.
\]

Residual variance follows the global projected contract

\[
SSE = yy-\sum_b w_b^\top w_b+\sum_b r_b^\top r_b.
\]

This is a retained SBayesRC-style eigenspace likelihood in `sblr`
cross-product units, not an exact reproduction of every GCTB variance update.

## Simulation and configurations

The v1 scientific design is preserved: sparse homogeneous BayesC/SBayesC and
sparse-mixture BayesR/SBayesR architectures, 5,000 samples, 37,991 canonical
chromosome-1 markers, 50 causal markers, target realized heritability 0.30,
five deterministic replicates, and training-only summary statistics and LD.

The six configurations are BED, full CSR, block CSR, near-full positive-rank
low rank, 0.999 retained mass, and the canonical 0.995 retained mass. All use
the same paired simulations and held-out test data.

## Block design

The primary v1 contiguous block boundaries are reused without outcome-based
selection. Cross-block LD is omitted by block CSR and all retained-factor
operators. Consequently, full CSR versus low rank combines a block-boundary
effect with retention; block CSR and near-full factors separate these effects.

## Deterministic validation

```{r}
#| echo: false
knitr::kable(deterministic, digits = 6)
```

The dense calculation `crossprod(Q)` is used only as an algebraic oracle. The
historical reconstructed-dense sampler is never called. The validation covers
scores, diagonals, residual rebuilds and updates, corrected marker scores,
quadratic forms, and both projected-SSE expressions.

## Retained-rank diagnostics

```{r}
#| echo: false
rank_summary <- aggregate(cbind(retained_rank, positive_rank,
  retained_eigenvalue_mass, positive_eigenvalue_mass) ~ configuration,
  low_rank, sum)
rank_summary$rank_fraction <- rank_summary$retained_rank /
  rank_summary$positive_rank
rank_summary$retained_mass_fraction <- rank_summary$retained_eigenvalue_mass /
  rank_summary$positive_eigenvalue_mass
knitr::kable(rank_summary, digits = 6)
```

```{r}
#| echo: false
#| fig-cap: "Retained positive-rank fraction by block and configuration."
ggplot2::ggplot(low_rank, ggplot2::aes(x = block_id,
    y = retained_rank / positive_rank, colour = configuration)) +
  ggplot2::geom_line() + ggplot2::labs(x = "Block", y = "Rank fraction") +
  ggplot2::theme_minimal()
```

Low rank is computationally useful when retained rank is much smaller than
block size. Full-rank factors can exceed packed-dense storage, while CSR can
remain preferable for genuinely sparse LD. This study does not implement an
automatic hybrid operator selector.

## One-replicate operator pilot

```{r}
#| echo: false
knitr::kable(pilot, digits = 5)
```

The bounded pilot verified schema, finite variance components, prediction,
effect alignment, checkpoint/resume, and practical execution before
convergence selection. It was not used to tune prediction.

## Convergence selection

```{r}
#| echo: false
knitr::kable(recommendations, digits = 4)
knitr::kable(convergence, digits = 4)
```

Four identifiable chains and retained histories were assessed using the
repository's established split R-hat, bulk ESS, tail ESS, relative MCSE,
finite-domain, and chain-identity gates. Five-simulation support is not
universal convergence validation.

## Five-replicate completion

The frozen capsule contains `r if (complete) "60/60 validated four-chain fits" else "an incomplete fit grid"`.
Failures remain in denominators and prevent capsule promotion.

Finalization detected and repaired an initial control mismatch in the five
`sparse_mixture / full_csr` fits. Only those five fits were rerun. The final
capsule requires exact agreement between every non-BED checkpoint and its
frozen convergence recommendation; the corrected full-CSR fits use 250 burn-in
iterations and 2,000 retained iterations.

```{r}
#| echo: false
controls <- unique(checkpoint_validation[c("architecture", "configuration",
  "nburn", "retained_iterations", "nthin", "chain_count", "ncores")])
controls <- controls[order(controls$architecture, controls$configuration), ]
knitr::kable(controls)
```

## Prediction

```{r}
#| echo: false
prediction_summary <- summarize_metric(prediction)
knitr::kable(prediction_summary, digits = 4)
```

```{r}
#| echo: false
#| fig-cap: "Paired phenotype-prediction correlations across five simulations."
pred_cor <- prediction[prediction$metric ==
  "phenotype_prediction_correlation", ]
ggplot2::ggplot(pred_cor, ggplot2::aes(configuration, value,
    group = interaction(architecture, replicate), colour = architecture)) +
  ggplot2::geom_line(alpha = .5) + ggplot2::geom_point() +
  ggplot2::coord_flip() + ggplot2::labs(x = NULL, y = "Prediction correlation") +
  ggplot2::theme_minimal()
```

## Parameter estimation

```{r}
#| echo: false
knitr::kable(recovery, digits = 4)
```

Observed interval coverage is coarse: each miss changes a five-replicate
proportion by 20 percentage points.

## Marker effects and PIPs

```{r}
#| echo: false
knitr::kable(summarize_metric(markers), digits = 4)
knitr::kable(summarize_metric(variable[variable$available, ]), digits = 4)
```

Unavailable posterior objects remain explicit in the capsule availability
table rather than being reconstructed from unrelated quantities.

## Operator comparison hierarchy

```{r}
#| echo: false
knitr::kable(paired, digits = 5)
```

The comparisons are interpreted in order: BED versus full CSR; full CSR
versus block CSR; near-full retained factors versus their direct (Q^\top Q)
oracle; 0.995 versus near-full; 0.995 versus 0.999; and full CSR versus the
practical 0.995 backend. Full CSR versus low rank is not attributed solely to
rank truncation because block boundaries also differ.

## Retention sensitivity

```{r}
#| echo: false
sensitivity <- marker_agreement[marker_agreement$comparison_id ==
  "low_rank_0995_minus_low_rank_0999", ]
knitr::kable(sensitivity, digits = 6)
```

A no-op or near-full-rank outcome is scientifically valid. Readiness depends
on deterministic identities, projected SSE, convergence, scientific
coherence, and prespecified 0.995-versus-0.999 sensitivity gates—not on a
required speed improvement.

The optimized sampler records the residual rebuild interval, rebuild count,
and maximum absolute residual drift for every retained-low-rank fit. These
diagnostics are validated against direct rebuilds before results are accepted.

```{r}
#| echo: false
drift <- checkpoint_validation[
  checkpoint_validation$operator_representation == "low_rank",
  c("architecture", "replicate", "configuration",
    "residual_rebuild_interval", "residual_rebuild_count",
    "maximum_residual_drift")]
drift_summary <- aggregate(cbind(residual_rebuild_count,
  maximum_residual_drift) ~ architecture + configuration, drift,
  function(x) c(mean = mean(x), maximum = max(x)))
knitr::kable(drift_summary, digits = 6)
```

## MTBLR readiness decision

The capsule decision is reported verbatim as either **READY FOR MTBLR
LOW-RANK EXTENSION** or **NOT READY FOR MTBLR LOW-RANK EXTENSION**. For this
completed capsule the decision is **`r readiness`**; it is not inferred from
runtime alone.

## SBayesR variance-component evidence

```{r}
#| echo: false
knitr::kable(sbayesr, digits = 5)
```

The current installed implementation is benchmarked as-is. Study 06 v2 does
not diagnose or alter earlier SBayesR heritability behaviour.

## Runtime and storage

```{r}
#| echo: false
runtime_summary <- aggregate(cbind(mcmc_runtime_seconds,
  seconds_per_retained_draw, fit_object_bytes, checkpoint_bytes,
  immutable_operator_bytes, per_chain_state_bytes) ~ architecture +
  configuration, runtime, function(x) mean(x, na.rm = TRUE))
knitr::kable(runtime_summary, digits = 3)
```

```{r}
#| echo: false
timing <- aggregate(cbind(mcmc_wall_seconds, construction_seconds,
  eigendecomposition_seconds, transformation_seconds,
  seconds_per_chain_iteration) ~ architecture + configuration,
  checkpoint_validation, function(x) mean(x, na.rm = TRUE), na.action = na.pass)
knitr::kable(timing, digits = 5)
```

Operator construction, eigendecomposition, transformation, and sampler wall
time are reported separately where the backend exposes them. Missing
preprocessing components for BED or CSR are not folded into fabricated values.
Seconds per chain-iteration use the recorded wall time, chain count, and total
iterations and should be interpreted as a normalized wall-time measure rather
than summed CPU time.

Measured rank, runtime, and storage—not theoretical complexity alone—support
the computational conclusions. The capsule excludes full (Q) matrices and
native fits.

## Limitations

This reduced development benchmark covers current scalar BayesC/SBayesC and
BayesR/SBayesR behaviour on one genotype resource and two architectures. It
does not establish universal performance, automatic block selection, or an
exact reproduction of all external SBayesRC implementations. Historical v1
posterior values are descriptive context only because source and residual
representations differ.

## Downloads and provenance

- [Convergence capsule](../../results/reference/06_ld_operator/st-low-rank-operator-convergence-development-v2/README.md)
- [Five-replicate capsule](../../results/reference/06_ld_operator/st-low-rank-operator-five-replicate-development-v2/README.md)
- [Historical reconstructed-dense Study 06 v1](ld-operator-development-pilot.qmd)
- [Development run record](../../docs/dev/study06_low_rank_operator_v2_run.md)

All scientific fits record the pinned `sblr` SHA, version, operator contract,
representation, eigen proportion, block definitions, chain seeds, and input
hash. No network access or package installation is part of reproduction.
