---
title: "Study 06 Supplemental: SBayesR Sensitivity to LD-Operator Approximation"
subtitle: "Exact, hard-sparse and block-eigen validation"
description: "Additional validation evidence for the sblr package."
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
root <- normalizePath(getwd(), winslash = "/")
while (!file.exists(file.path(root, "sblrbench.Rproj"))) root <- dirname(root)
capsule <- file.path(root, "results", "reference", "06_ld_operator",
  "sbayesr_ld_robustness", "current")
rcsv <- function(name) utils::read.csv(file.path(capsule, name), check.names = FALSE)
design <- rcsv("design.csv")
provenance <- rcsv("provenance.csv")
fit_status <- rcsv("fit_status.csv")
scheduler <- rcsv("scheduler_comparison.csv")
operators <- rcsv("operator_summary.csv")
pairwise <- rcsv("operator_pairwise_comparison.csv")
eigenvalues <- rcsv("eigenvalue_summary.csv")
blocks <- rcsv("block_diagnostics.csv")
variance <- rcsv("variance_summary.csv")
pis <- rcsv("pi_summary.csv")
components <- rcsv("component_summary.csv")
convergence <- rcsv("convergence_summary.csv")
effects <- rcsv("effect_recovery.csv")
prediction <- rcsv("prediction_metrics.csv")
conditionals <- rcsv("conditional_marker_summary.csv")
quadratics <- rcsv("operator_quadratic_checks.csv")
scores <- rcsv("operator_score_checks.csv")
posterior_operator <- rcsv("posterior_operator_comparison.csv")
decomposition <- rcsv("quadratic_error_decomposition.csv")
low_direction <- rcsv("low_eigen_direction_summary.csv")
prior <- rcsv("prior_resolution.csv")
inputs <- rcsv("input_equivalence.csv")
kable <- function(x, digits = 5) knitr::kable(x, digits = digits)
method_order <- c("BED exact data", "CSR exact LD", "CSR hard sparse",
  "Block eigen full rank", "Block eigen retained 0.995")
```

## Validation question

How do exact, element-wise hard-sparse, full-rank block-eigen, and
retained-low-rank LD operators affect SBayesR marker conditionals, posterior
variance estimates, component allocation, and recovery? This is a deliberately
focused source-linked validation of the `sblr` package. It does not compare
every `sblr` method or select a genome-wide LD policy.

::: {.callout-note}
## Additional validation evidence

This page supplements the established Study 06 retained-low-rank benchmark. It
tests a specific SBayesR discrepancy and does not change current `sblr`
defaults or make a general methodological claim.
:::

## Relationship to the broader `sblrbench` validation suite

The [main Study 06 retained-low-rank benchmark](../report.qmd)
remains the broader LD-operator validation. Study 03 remains the authoritative
parameter-estimation benchmark. This supplemental Study 06 analysis uses one
deterministic stress-test window and a new reduced phenotype so omitted genetic
signal outside the window cannot confound operator identities. The full Study
03 discrepancy is retained as motivation; these fits are not promoted back
into Study 03.

## Provenance

All numerical fits use the installed `sblr` source SHA shown below. The three
original reduced-window fits and all scheduler fits were reused; only the two
block-eigen variants were added.

```{r}
kable(provenance)
kable(fit_status[, c("label", "elapsed_seconds", "fit_origin", "warnings")], 3)
```

## Full Study 03 discrepancy

Frozen Study 03 evidence showed BED BayesR behaving reasonably while CSR
SBayesR had inflated genetic variance and heritability, reduced residual
variance, and poorer recovery. The scheduler diagnostic isolates whether the
BED marker-traversal policy explains that contrast.

```{r}
study03_frozen <- scheduler[scheduler$quantity %in% c("heritability", "vgs", "ves"), ]
study03_frozen <- reshape(study03_frozen[, c("label", "quantity", "mean")],
  idvar = "label", timevar = "quantity", direction = "wide")
names(study03_frozen) <- sub("mean\\.", "", names(study03_frozen))
kable(study03_frozen, 5)
```

## BED scheduler diagnostic

Scheduled and full-sweep BED BayesR were nearly identical on Study 03
sparse-mixture replicate 1, while current CSR SBayesR remained different. This
rules out the adaptive BED scheduler as the primary explanation.

```{r}
wide_scheduler <- reshape(scheduler[, c("label", "quantity", "mean")],
  idvar = "label", timevar = "quantity", direction = "wide")
names(wide_scheduler) <- sub("mean\\.", "", names(wide_scheduler))
kable(wide_scheduler, 5)
```

## Reduced LD-rich window design

The selected 1,500-marker window maximized retained absolute Study-03 sparse-LD
magnitude among all contiguous windows. It contains 5,000 samples, 15 causal
markers, and a phenotype simulated at realized heritability 0.30 with seed
17,002.

```{r}
kable(design[, c("qc_start", "qc_end", "first_marker", "last_marker",
  "samples", "markers", "causal_markers", "realized_h2")])
```

## Exact CSR operator validation

The full-pair CSR correlation representation was rescaled by the marker-specific
BED cross-product diagonals. It reproduced direct $X'X$ with maximum absolute
error (2.84\times10^{-5}), relative Frobenius error
(1.95\times10^{-8}), negligible diagonal error, and exact symmetry at the
reported precision.

## Exact CSR versus BED BayesR

Exact CSR closely reproduced full-sweep BED in posterior variance, component
allocation, marker effects and genetic-value recovery. The deterministic audit
also reproduced the current BayesR score, component-probability and conditional
effect formulas. No defect in those published/current conditional formulas was
identified.

```{r}
exact_bed <- variance[variance$variant %in% c("bed_exact_data", "csr_exact_ld") &
  variance$quantity %in% c("heritability", "vgs", "ves", "vbs"), ]
exact_bed <- reshape(exact_bed[, c("label", "quantity", "mean")],
  idvar = "label", timevar = "quantity", direction = "wide")
names(exact_bed) <- sub("mean\\.", "", names(exact_bed))
kable(exact_bed, 5)
```

## Hard-sparse CSR result

Exact CSR closely reproduced full-sweep BED. The hard-sparse operator moved
genetic variance and heritability upward, residual variance downward, and
assigned more markers moderate posterior inclusion probability.

```{r}
central <- variance[variance$quantity %in% c("heritability", "vgs", "ves", "vbs"), ]
central_wide <- reshape(central[, c("label", "quantity", "mean")],
  idvar = "label", timevar = "quantity", direction = "wide")
names(central_wide) <- sub("mean\\.", "", names(central_wide))
central_wide$label <- factor(central_wide$label, levels = method_order)
central_wide <- central_wide[order(central_wide$label), ]
kable(central_wide, 5)
```

## Deterministic conditional audit

Direct/BED and exact-CSR corrected scores, component probabilities, and
conditional effect means agree to numerical precision. For the fixed marker
panel, hard-sparse score differences reached 82.99 and component-probability
differences reached 0.1715.

```{r}
kable(conditionals, 7)
```

## Residual and quadratic-form audit

For fixed effect states, direct genotype and exact-CSR quadratic and residual
identities agree. Hard-sparse and block-diagonal operators understate the
quadratic form and corresponding residual expression because omitted LD terms
are absent.

```{r}
kable(quadratics[quadratics$effect_state %in% c("truth", "csr_sparse_ld"), ], 4)
```

## Spectral properties

Neither exact nor hard-sparse LD has eigenvalues below (-10^{-8}) or within
(pm10^{-8}). Hard sparsification is therefore not producing an indefinite
operator in this window. It instead compresses the spectrum and accumulates a
large matrix error from omitted entries.

```{r}
kable(eigenvalues[, c("operator", "minimum", "maximum",
  "below_negative_tolerance", "near_zero", "effective_rank",
  "positive_condition_number")], 4)
```

The hard-sparse relative Frobenius error is 0.4422 despite an exactly preserved
diagonal. The absence of negative and near-zero directions makes accumulated
omitted weak LD the better-supported mechanism here.

## Full-rank block-eigen validation

Blocks use deterministic 250-marker starts and split at the chromosome
boundary, yielding starts `r design$block_starts`. The native
`dense_reconstructed` representation with `ridge_fixed` and `eta=0` reproduces
its configured block-diagonal target to numerical tolerance. It does **not**
reproduce global exact LD: cross-block omission gives relative Frobenius error
0.4427. This is Case B under the prespecified decision rule.

```{r}
kable(pairwise, 5)
```

Thus the full-rank block-eigen implementation is internally faithful to its
blocks, but its configured representation is not equivalent to exact CSR in
this diagnostic.

## Retained-low-rank result

The production-like retained representation uses cumulative positive mass
0.995 and rebuilds reduced residuals every 100 iterations. It retained 1,490 of
1,500 block directions. Five 250-marker blocks retained 248 directions each;
the two chromosome-boundary fragments retained full rank.

```{r}
kable(blocks[, c("block_id", "start", "end", "positive_rank",
  "retained_rank", "retained_mass_fraction")], 5)
```

Retention modestly moves heritability, genetic variance, residual variance and
recovery toward exact CSR relative to full-rank blocking, but block-boundary
error dominates both operators. It therefore cannot yet be called a validated
alternative to exact CSR.

## Component allocation and PIP behavior

```{r}
kable(central_wide, 5)
active <- pis[pis$quantity == "active_probability", c("label", "mean", "sd")]
active$label <- factor(active$label, levels = method_order)
kable(active[order(active$label), ], 6)
kable(components, 3)
```

All active-probability traces remain near 0.01 because the common Dirichlet
prior is highly concentrated. Expected non-null counts are about 12.8–12.9 for
BED/exact CSR and about 15.5–15.7 for hard-sparse and block operators.

## Effect and genetic-value recovery

```{r}
kable(effects, 6)
kable(prediction, 6)
```

Exact CSR and BED provide the strongest recovery. Retained low rank improves
slightly over full-rank blocking and hard sparsification, but the differences
are small in one replicate.

## Posterior-mean operator comparison

Direct genotype and exact-operator quadratics agree. For the sparse-CSR
posterior-mean effects, the exact quadratic is 2083.57, versus 2020.88 under
hard sparsification. Full-rank blocking and retained low rank show comparable
errors because all omit cross-block LD.

```{r}
kable(quadratics[quadratics$effect_state %in%
  c("truth", "csr_sparse_ld", "block_eigen_full_rank", "block_eigen_retained"), ], 4)
kable(posterior_operator, 5)
```

These are identities for fixed or posterior-mean effect vectors, not posterior
means of variance traces.

### Spectral projection of fitted effects

Projection was evaluated in both exact and sparse eigenbases. Because the bases
differ, these are diagnostic representations rather than unique causal
decompositions. No contribution comes from negative sparse-eigenvalue
directions because none exist.

```{r}
kable(decomposition, 4)
kable(low_direction, 6)
```

Only a minority of the exact-minus-sparse quadratic discrepancy is localized
to the lowest sparse eigenvalue directions. The error is distributed broadly,
again favoring accumulated omitted LD over a small set of unstable modes.

## Integrated interpretation

Established findings are:

1. BED scheduling does not explain the original BED-versus-CSR discrepancy.
2. Exact CSR reproduces BED BayesR and direct deterministic conditionals.
3. Hard sparsification changes marker scores, component probabilities,
   quadratic forms, residual expressions, posterior variance and recovery.
4. The sparse operator remains positive definite; indefiniteness is not the
   mechanism in this window.
5. Unfiltered full-rank block eigen reproduces its block target, but not global
   exact LD because cross-block entries are omitted.
6. Retaining 0.995 positive mass adds only ten discarded directions and gives
   modest improvement, while the block-boundary approximation remains dominant.

Within one deterministic LD-rich 1,500-marker window, exact CSR reproduced
full-sweep BED BayesR. Hard-sparse and block-diagonal LD operators omitted weak
or cross-block LD, altering corrected marker scores, component probabilities,
residual expressions and posterior recovery. The hard-sparse operator was
positive definite, and retaining 99.5% of within-block spectral mass gave only
a modest improvement.

## Validation boundary

- One selected window and one phenotype are diagnostic rather than definitive.
- The block policy deliberately follows a deterministic repository fallback;
  another prespecified block definition could behave differently.
- Shared chain seeds do not imply identical random streams across backends.
- The concentrated mixture prior limits movement in posterior global `pi`.
- Posterior intervals describe this fitted model, not repeated-window uncertainty.
- This window does not prove the complete cause of every whole-genome result.
- Optimal LD construction is data dependent; no universal threshold is selected.
- The study does not assert that GCTB or another implementation has the same
  specific behavior, and it does not imply that `sblr` itself is invalid.

## Practical implications

SBayesR validation should include operator-level corrected-score and quadratic
checks, not only diagonal agreement or matrix density. A block-eigen route must
first demonstrate equivalence to its intended global or block target before
rank retention is judged by posterior performance. Current `sblr` defaults are
not changed by this evidence.

## Deferred work

The following are explicitly deferred: LD window/threshold sensitivity,
whole-genome reruns, application-specific LD tuning, shrinkage LD, PSD
correction, prior-strength sensitivity, `adjE` sensitivity, revised defaults,
rerunning Studies 01–06, and modifying `sblr`.

## Reproducibility and downloads

The report reads only the compact supplemental Study 06 capsule. Large local checkpoints are
not tracked.

- [Study README](README.md)
- [Analysis entry point](analysis.R)
- [Authoritative Study 06 specification](../spec.R)
- [Main Study 06 operator-validation report](../report.qmd)
- [Scheduler diagnostic script](scripts/scheduler-diagnostic.R)
- [Exact-versus-sparse diagnostic script](scripts/exact-sparse-diagnostic.R)
- [Eigen extension helpers](scripts/eigen-extension.R)
- [Capsule README](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/README.md)
- [Manifest](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/benchmark_manifest.json)
- [Checksums](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/checksums.csv)
- [Operator summary](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/operator_summary.csv)
- [Variance summary](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/variance_summary.csv)
- [Component summary](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/component_summary.csv)
- [Effect recovery](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/effect_recovery.csv)
- [Prediction metrics](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/prediction_metrics.csv)
- [Corrected-score checks](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/operator_score_checks.csv)
- [Quadratic checks](../../../results/reference/06_ld_operator/sbayesr_ld_robustness/current/operator_quadratic_checks.csv)

```powershell
Rscript studies/06_ld_operator/sbayesr_ld_robustness/analysis.R
quarto render studies/06_ld_operator/sbayesr_ld_robustness/report.qmd
```
