---
title: "Study 02: Single-trait prediction"
description: "A five-replicate held-out prediction benchmark across BayesC, BayesR, BED and CSR workflows."
---

::: {.callout-warning .callout-development}
## Development benchmark—not a final scientific comparison.
Fits use Study 04's method-specific retained-draw recommendations, 250 warm-up
iterations and four identifiable chains. Test phenotypes are held out, and scaling,
LD and summary statistics use training individuals only.
:::

```{r}
#| echo: false
historical_capsule <- file.path("results", "reference", "02_prediction",
  "st-bayesc-bayesr-one-replicate-development-v1")
five_capsule <- file.path("results", "reference", "02_prediction",
  "st-bayesc-bayesr-five-replicate-development-v1")
source("studies/reporting_helpers.R")
if (dir.exists(five_capsule)) {
  source(file.path("studies", "five_replicate_promotion.R"), local = TRUE)
  .five_validate_study02_capsule(five_capsule); capsule <- five_capsule
} else {
  source(file.path("studies", "02_prediction", "promotion.R"), local = TRUE)
  .study02_validate_capsule(historical_capsule); capsule <- historical_capsule
}
metrics <- read.csv(file.path(capsule, "prediction_metrics.csv"))
paired <- read.csv(file.path(capsule, "paired_method_differences.csv"))
computation <- read.csv(file.path(capsule, "computational_summary.csv"))
status <- read.csv(file.path(capsule, "replicate_status.csv"))
simulations <- read.csv(file.path(capsule, "simulation_summary.csv"))
benchmark_summary <- read.csv(file.path(capsule, "benchmark_summary.csv"))
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
example_data <- read.csv(file.path(capsule,"example_data_manifest.csv"))
metrics$method_label <- sblrbench_method_factor(metrics$method)
metrics$architecture_label <- sblrbench_architecture_factor(metrics$architecture)
computation$method_label <- sblrbench_method_factor(computation$method)
computation$architecture_label <- sblrbench_architecture_factor(computation$architecture)
available <- TRUE
```

## Benchmark status

> **Complete five-replicate development benchmark.** Aggregate uncertainty is
> descriptive across paired simulations and is not a general method-ranking claim.

The primary capsule records 2 architectures, `r manifest$replicate_count` replicates per architecture, 4
methods, `r manifest$expected_fit_count` expected fits, `r manifest$successful_fit_count` successful fits and `r manifest$failed_fit_count` failed fits. Quarto reads only
the tracked capsule and never runs simulations or samplers.

## Scientific question

How do BayesC and BayesR prediction models compare when fitted from
individual-level BED data versus summary statistics and sparse-LD CSR data
under different single-trait genetic architectures?

## Methods compared

- ST-BED BayesC
- ST-BED BayesR
- ST-CSR SBayesC
- ST-CSR SBayesR

The summary-statistics `s` prefix denotes a corresponding model class, not an
identical likelihood or representation.

## Genetic architectures

`sparse_homogeneous` samples 50 causal effects from one normal distribution.
`sparse_mixture` assigns the same number of causals to variance classes 0.01,
0.1 and 1 with probabilities 0.60, 0.30 and 0.10. Each complete effect vector
is globally rescaled to realized h2 = 0.30. These are approximately BayesC-like
and BayesR-like architectures, not exact draws from either fitted sampler prior.

## Data and simulation design

The pinned public qgdata PLINK resource contributes 5,000 samples and 37,991
post-QC chromosome-1 markers. Every architecture and replicate uses the same
marker set, causal count, target heritability and split.

## Train/test and leakage controls

The deterministic split contains 3,500 training and 1,500 test samples.
Training rows alone determine allele frequencies, imputation, genotype scaling,
sparse LD and summary statistics. Test genotypes are transformed on the frozen
training scale. Test phenotypes enter evaluation metrics only.

## Configuration

The primary five-replicate capsule records the Study 04 operational settings:
250 burn-in iterations, thinning 1, four identifiable chains and four cores.
BayesC and SBayesC retain 250 post-burn-in draws per chain; BayesR and SBayesR
retain 1,000. The historical one-replicate capsule used 500 retained iterations,
250 burn-in iterations, one chain and one core; those controls do not describe
the primary five-replicate results. BayesR uses explicit mixture multipliers
`c(0, .01, .1, 1)`; BayesC uses an explicit initial inclusion probability.

```{r}
#| echo: false
knitr::kable(manifest$mcmc_controls[, c(
  "method", "recommended_nburn", "recommended_post_burnin_draws",
  "recommended_nthin", "recommended_nchains", "recommended_ncores"
)], caption = "MCMC controls recorded in the primary five-replicate capsule.")
```

## Validation checks

Each branch requires 50 unique causals, target realized heritability, a passing
`G = ZB` oracle, training `reference_n = 3500`, summary-statistics `n = 3500`,
canonical marker order, aligned effects and finite held-out predictions.

```{r}
#| echo: false
status_display <- status
status_display$method <- sblrbench_method_factor(status_display$method)
names(status_display) <- c("Architecture","Replicate","Method","Status","Reason")
knitr::kable(status_display)
```

## Prediction results

```{r}
#| echo: false
display_metrics <- c("prediction_correlation", "prediction_nmse",
  "phenotype_prediction_correlation", "prediction_calibration_intercept",
  "prediction_calibration_slope", "effect_rmse")
main <- reshape(metrics[metrics$metric %in% display_metrics,
  c("architecture", "method", "metric", "value")],
  idvar = c("architecture", "method"), timevar = "metric", direction = "wide")
main <- merge(main, computation[, c("architecture", "method", "runtime")],
  by = c("architecture", "method"), sort = FALSE)
names(main) <- sub("^value\\.", "", names(main))
main$method <- sblrbench_method_factor(main$method)
main$architecture <- sblrbench_architecture_factor(main$architecture)
names(main) <- c("Architecture","Method","Prediction correlation","Prediction NMSE",
  "Phenotype correlation","Calibration intercept","Calibration slope","Effect RMSE","Runtime (s)")
knitr::kable(main,digits=4)
```

### Prediction correlation by architecture and method

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "prediction_correlation", ],
  ggplot2::aes(method, value, colour = method,shape=method)) + ggplot2::geom_point(size=2) +
  ggplot2::facet_wrap(~architecture_label) + sblrbench_method_scales()+theme_sblrbench() +
  ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())+
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 25, hjust = 1), legend.position = "none"))
```

### Prediction NMSE

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "prediction_nmse", ],
  ggplot2::aes(method, value, colour = method,shape=method)) + ggplot2::geom_point(size=2) +
  ggplot2::facet_wrap(~architecture_label) + sblrbench_method_scales()+theme_sblrbench() +
  ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())+
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

### Phenotype prediction correlation

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "phenotype_prediction_correlation", ],
  ggplot2::aes(method, value, colour = method,shape=method)) + ggplot2::geom_point(size=2) +
  ggplot2::facet_wrap(~architecture_label) + sblrbench_method_scales()+theme_sblrbench() +
  ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())+
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

### Calibration slope

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "prediction_calibration_slope", ],
  ggplot2::aes(method, value, colour = method,shape=method)) + ggplot2::geom_hline(yintercept = 1, linetype = 2) +
  ggplot2::geom_point(size=2) + ggplot2::facet_wrap(~architecture_label) + sblrbench_method_scales()+theme_sblrbench() +
  ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())+
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

## Calibration

Calibration intercept and slope are reported in the main table. The slope plot
uses 1 as the descriptive reference. Variation across five paired simulations
is descriptive development evidence, not a calibrated uncertainty estimate.

## BayesR-versus-BayesC comparisons

Positive advantages favor BayesR/SBayesR. Differences are paired by replicate;
five pairs remain descriptive rather than definitive inferential evidence.

```{r}
#| echo: false
#| fig-width: 10
if (available) {
  z <- paired[grepl("bayesr_vs_bayesc", paired$comparison_id), ]
  print(ggplot2::ggplot(z, ggplot2::aes(paired_metric, advantage)) +
    ggplot2::geom_hline(yintercept = 0, linetype = 2) + ggplot2::geom_point(colour="#4B5563") +
    ggplot2::facet_wrap(~architecture, scales = "free_x") + ggplot2::coord_flip() + theme_sblrbench())
}
```

## CSR versus BED comparisons

Positive advantages favor CSR for the corresponding model class.

```{r}
#| echo: false
#| fig-width: 10
if (available) {
  z <- paired[grepl("csr_vs_bed", paired$comparison_id), ]
  print(ggplot2::ggplot(z, ggplot2::aes(paired_metric, advantage)) +
    ggplot2::geom_hline(yintercept = 0, linetype = 2) + ggplot2::geom_point(colour="#4B5563") +
    ggplot2::facet_wrap(~architecture, scales = "free_x") + ggplot2::coord_flip() + theme_sblrbench())
}
```

## Computational performance

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(computation, ggplot2::aes(method, runtime, colour = method,shape=method)) +
  ggplot2::geom_point(size=2) + ggplot2::facet_wrap(~architecture_label) + sblrbench_method_scales()+theme_sblrbench() +
  ggplot2::scale_x_discrete(limits=sblrbench_method_levels(),labels=sblrbench_method_labels())+
  ggplot2::scale_y_log10() + ggplot2::labs(y = "Runtime (seconds, log scale)") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

## Interpretation

The development run validates paired BED/CSR prediction and prior-class
comparisons. It does not establish general performance rankings.

## Limitations

MCMC uses the Study 04 operational settings and four chains, but five simulations
do not establish universal convergence or performance. The public simulated
resource has limited LD, and only one heritability and two stylized sparse
architectures are represented.

## Code and reproduction

## Worked example

This frozen script fits all four active models to one held-out single-trait prediction problem. It is displayed but not executed during rendering.

```{r, echo=FALSE, results='asis'}
display_capsule_script(file.path(historical_capsule,"worked_prediction_example.R"))
```

[Download the worked example](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/worked_prediction_example.R).

## Public example data

The simulated PLINK files are checksum-validated from the pinned qgdata revision.

```{r, echo=FALSE}
knitr::kable(example_data[c("filename","role","size_bytes","md5")],
  col.names=c("File","Role","Bytes","MD5"))
```

[Download the example-data manifest](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/example_data_manifest.csv).

## Developer contract smoke test

This sampler-free test validates metrics, alignment and paired comparison orientation.

```{r, echo=FALSE, results='asis'}
display_capsule_script_collapsed(file.path(historical_capsule,"prediction_contract_smoke_test.R"))
```

[Download the smoke test](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/prediction_contract_smoke_test.R).

## Complete benchmark reproduction

The frozen targets entry point reproduces the benchmark but is never executed while rendering.

```{r, echo=FALSE, results='asis'}
display_capsule_script(file.path(historical_capsule,"run_prediction_benchmark.R"))
```

[Download the reproduction script](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/run_prediction_benchmark.R).

## Downloads and provenance

| Resource | Download |
|---|---|
| Primary capsule | [README](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/README.md); [manifest](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/benchmark_manifest.json) |
| Primary results | [Benchmark summary](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/benchmark_summary.csv); [prediction metrics](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/prediction_metrics.csv); [paired summary](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/paired_comparison_summary.csv) |
| Computational evidence | [Runtime](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/computational_summary.csv); [fit status](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/replicate_status.csv) |
| Historical reproduction material | [Configuration](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/config.R); [worked example](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/worked_prediction_example.R); [smoke test](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/prediction_contract_smoke_test.R); [run script](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/run_prediction_benchmark.R) |
| Primary provenance | [Example data](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/example_data_manifest.csv); [source inventory](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/source_files.csv); [session information](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/session_info.txt); [checksums](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/checksums.csv) |

The [five-replicate capsule](../../results/reference/02_prediction/st-bayesc-bayesr-five-replicate-development-v1/README.md)
is the primary result source. The [original one-replicate capsule](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/README.md)
remains an immutable historical reference.

## Software versions and commits

```{r, echo=FALSE}
provenance <- data.frame(Item=c("sblrbench commit","sblr commit","qgdata commit","Source tree","Profile"),
  Value=c(manifest$sblrbench_source_commit,manifest$versions$sblr_source_commit,manifest$qgdata$commit,
    if(manifest$source_tree_clean) "Clean" else "Dirty","Development"))
knitr::kable(provenance)
```
