---
title: "Study 02: Single-trait prediction"
description: "Current five-replicate held-out prediction benchmark."
---

```{r}
#| echo: false
capsule <- "results/reference/02_prediction/current"
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"),
  simplifyVector = TRUE)
summary <- read.csv(file.path(capsule, "benchmark_summary.csv"),
  check.names = FALSE)
paired <- read.csv(file.path(capsule, "paired_comparison_summary.csv"),
  check.names = FALSE)
status <- read.csv(file.path(capsule, "replicate_status.csv"),
  check.names = FALSE)
```

::: {.callout-warning}
This complete five-replicate development benchmark provides descriptive paired
evidence. It does not establish a universal prediction ranking.
:::

## Design and execution contract

Two architectures—sparse homogeneous and sparse variance mixture—were evaluated
in five paired replicates with ST-BED BayesC/BayesR and ST-CSR SBayesC/SBayesR.
All `r nrow(status)` fits completed with four identifiable chains. The current
run uses the validated Study 04 recommendations: 250 burn-in plus 250 retained
draws for BayesC/SBayesC, and 250 burn-in plus 2,000 retained draws for
BayesR/SBayesR; no thinning was used.

The deterministic 70/30 split is shared within replicate. Allele frequencies,
genotype scaling, summary statistics, and sparse LD use training individuals
only. Test phenotypes enter evaluation only. The exact package source is
`r manifest$versions$sblr_source_commit`.

## Held-out results

```{r}
#| echo: false
keep <- c("phenotype_prediction_correlation", "prediction_correlation",
  "prediction_nmse", "prediction_calibration_slope", "effect_rmse")
knitr::kable(summary[summary$metric %in% keep,
  c("architecture", "method_label", "metric", "successful_replicates",
    "mean", "sd", "minimum", "maximum")], digits = 4)
```

## Paired method contrasts

```{r}
#| echo: false
shown <- paired[paired$metric %in%
  c("phenotype_prediction_correlation", "prediction_correlation",
    "prediction_nmse"), ]
knitr::kable(shown, digits = 4)
```

## Reproducibility

- [Current capsule](../../results/reference/02_prediction/current/README.md)
- [Manifest](../../results/reference/02_prediction/current/benchmark_manifest.json)
- [Prediction metrics](../../results/reference/02_prediction/current/prediction_metrics.csv)
- [Paired replicate differences](../../results/reference/02_prediction/current/paired_method_differences.csv)
- [Checksums](../../results/reference/02_prediction/current/checksums.csv)

The report reads one canonical capsule; earlier one-replicate and versioned
development snapshots remain accessible through Git history.
