---
title: "Single-trait prediction development benchmark"
---

> **Development benchmark—not a final scientific comparison.** Fits use 500
> iterations, 250 burn-in iterations and one chain with simulated public
> genotypes that have limited LD. Test phenotypes are held out, and scaling,
> LD and summary statistics use training individuals only. No convergence or
> general superiority claim is supported.

```{r}
#| echo: false
capsule <- file.path("results", "reference", "02_prediction",
  "st-bayesc-bayesr-one-replicate-development-v1")
source(file.path("studies", "02_prediction", "promotion.R"), local = TRUE)
.study02_validate_capsule(capsule)
metrics <- read.csv(file.path(capsule, "prediction_metrics.csv"))
paired <- read.csv(file.path(capsule, "paired_method_differences.csv"))
computation <- read.csv(file.path(capsule, "computational_summary.csv"))
status <- read.csv(file.path(capsule, "replicate_status.csv"))
simulations <- read.csv(file.path(capsule, "simulation_summary.csv"))
benchmark_summary <- read.csv(file.path(capsule, "benchmark_summary.csv"))
manifest <- jsonlite::read_json(file.path(capsule, "benchmark_manifest.json"), simplifyVector = TRUE)
available <- TRUE
```

## Benchmark status

> This is a complete full-size one-replicate development benchmark. It validates
> the complete prediction workflow but does not estimate replicate-to-replicate
> uncertainty or support general method-ranking claims.

The frozen capsule records 2 architectures, 1 replicate per architecture, 4
methods, 8 expected fits, 8 successful fits and 0 failed fits. Quarto reads only
the tracked capsule and never runs simulations or samplers.

## Scientific question

How do BayesC and BayesR prediction models compare when fitted from
individual-level BED data versus summary statistics and sparse-LD CSR data
under different single-trait genetic architectures?

## Methods compared

- ST-BED BayesC
- ST-BED BayesR
- ST-CSR SBayesC
- ST-CSR SBayesR

The summary-statistics `s` prefix denotes a corresponding model class, not an
identical likelihood or representation.

## Genetic architectures

`sparse_homogeneous` samples 50 causal effects from one normal distribution.
`sparse_mixture` assigns the same number of causals to variance classes 0.01,
0.1 and 1 with probabilities 0.60, 0.30 and 0.10. Each complete effect vector
is globally rescaled to realized h2 = 0.30. These are approximately BayesC-like
and BayesR-like architectures, not exact draws from either fitted sampler prior.

## Data and simulation design

The pinned public qgdata PLINK resource contributes 5,000 samples and 37,991
post-QC chromosome-1 markers. Every architecture and replicate uses the same
marker set, causal count, target heritability and split.

## Train/test and leakage controls

The deterministic split contains 3,500 training and 1,500 test samples.
Training rows alone determine allele frequencies, imputation, genotype scaling,
sparse LD and summary statistics. Test genotypes are transformed on the frozen
training scale. Test phenotypes enter evaluation metrics only.

## Configuration

Development controls are 500 iterations, 250 burn-in, thinning 1, one chain and
one core. BayesR uses explicit mixture multipliers `c(0, .01, .1, 1)`; BayesC
uses an explicit initial inclusion probability. Prior parameterizations differ
where required by their public model contracts.

## Validation checks

Each branch requires 50 unique causals, target realized heritability, a passing
`G = ZB` oracle, training `reference_n = 3500`, summary-statistics `n = 3500`,
canonical marker order, aligned effects and finite held-out predictions.

```{r}
#| echo: false
print(status)
```

## Prediction results

```{r}
#| echo: false
display_metrics <- c("prediction_correlation", "prediction_nmse",
  "phenotype_prediction_correlation", "prediction_calibration_intercept",
  "prediction_calibration_slope", "effect_rmse")
main <- reshape(metrics[metrics$metric %in% display_metrics,
  c("architecture", "method", "metric", "value")],
  idvar = c("architecture", "method"), timevar = "metric", direction = "wide")
main <- merge(main, computation[, c("architecture", "method", "runtime")],
  by = c("architecture", "method"), sort = FALSE)
names(main) <- sub("^value\\.", "", names(main))
print(main, row.names = FALSE)
```

### Prediction correlation by architecture and method

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "prediction_correlation", ],
  ggplot2::aes(method, value, colour = method)) + ggplot2::geom_point() +
  ggplot2::facet_wrap(~architecture) + ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

### Prediction NMSE

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "prediction_nmse", ],
  ggplot2::aes(method, value, colour = method)) + ggplot2::geom_point() +
  ggplot2::facet_wrap(~architecture) + ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

### Phenotype prediction correlation

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "phenotype_prediction_correlation", ],
  ggplot2::aes(method, value, colour = method)) + ggplot2::geom_point() +
  ggplot2::facet_wrap(~architecture) + ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

### Calibration slope

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(metrics[metrics$metric == "prediction_calibration_slope", ],
  ggplot2::aes(method, value, colour = method)) + ggplot2::geom_hline(yintercept = 1, linetype = 2) +
  ggplot2::geom_point() + ggplot2::facet_wrap(~architecture) + ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

## Calibration

Calibration intercept and slope are reported in the main table. The slope plot
uses 1 as the descriptive reference; one replicate provides no calibration
uncertainty estimate.

## BayesR-versus-BayesC comparisons

Positive advantages favor BayesR/SBayesR. With one replicate, points are paired
descriptions rather than uncertainty estimates.

```{r}
#| echo: false
#| fig-width: 10
if (available) {
  z <- paired[grepl("bayesr_vs_bayesc", paired$comparison_id), ]
  print(ggplot2::ggplot(z, ggplot2::aes(paired_metric, advantage, colour = comparison_id)) +
    ggplot2::geom_hline(yintercept = 0, linetype = 2) + ggplot2::geom_point() +
    ggplot2::facet_wrap(~architecture, scales = "free_x") + ggplot2::coord_flip() + ggplot2::theme_minimal())
}
```

## CSR-versus-BED comparisons

Positive advantages favor CSR for the corresponding model class.

```{r}
#| echo: false
#| fig-width: 10
if (available) {
  z <- paired[grepl("csr_vs_bed", paired$comparison_id), ]
  print(ggplot2::ggplot(z, ggplot2::aes(paired_metric, advantage, colour = comparison_id)) +
    ggplot2::geom_hline(yintercept = 0, linetype = 2) + ggplot2::geom_point() +
    ggplot2::facet_wrap(~architecture, scales = "free_x") + ggplot2::coord_flip() + ggplot2::theme_minimal())
}
```

## Runtime

```{r}
#| echo: false
#| fig-width: 9
if (available) print(ggplot2::ggplot(computation, ggplot2::aes(method, runtime, colour = method)) +
  ggplot2::geom_point() + ggplot2::facet_wrap(~architecture) + ggplot2::theme_minimal() +
  ggplot2::scale_y_log10() + ggplot2::labs(y = "Runtime (seconds, log scale)") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 35, hjust = 1), legend.position = "none"))
```

## Interpretation

The development run validates paired BED/CSR prediction and prior-class
comparisons. It does not establish general performance rankings.

## Limitations

MCMC is short and single-chain; replicate counts are preliminary; the public
simulated resource has limited LD; only one heritability and two stylized sparse
architectures are represented; convergence is not assessed.

## Worked example

The [worked example](../../scripts/worked_prediction_example.R) fits all four
active models to one held-out single-trait prediction problem.

## Public example data

Data are downloaded and checksum-validated from pinned qgdata commit
`6cca5819e711d326cfb2614d7e9d9f34942612cd`.

## Developer smoke test

The [prediction contract smoke test](../../scripts/prediction_contract_smoke_test.R)
validates metrics, alignment and paired comparison orientation without MCMC.

## Complete benchmark reproduction

Use the staged [targets entry point](../../scripts/run_prediction_benchmark.R).
Rendering this report never executes it.

## Downloads

The complete [benchmark reproducibility capsule](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/README.md)
contains direct downloads for the [manifest](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/benchmark_manifest.json),
[benchmark summary](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/benchmark_summary.csv),
[paired summary](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/paired_comparison_summary.csv),
[prediction metrics](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/prediction_metrics.csv),
[paired rows](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/paired_method_differences.csv),
[runtime](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/computational_summary.csv),
[fit status](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/replicate_status.csv),
[simulation summary](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/simulation_summary.csv),
[configuration](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/config.R),
[run script](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/run_prediction_benchmark.R),
[worked example](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/worked_prediction_example.R),
[smoke test](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/prediction_contract_smoke_test.R),
[data manifest](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/example_data_manifest.csv),
[source inventory](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/source_files.csv),
[session information](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/session_info.txt), and
[checksums](../../results/reference/02_prediction/st-bayesc-bayesr-one-replicate-development-v1/checksums.csv).

## Software versions and commits

Compact outputs record package versions, source commit/dirty state, split,
simulation and MCMC seeds, controls, runtime and branch status.
