---
title: "Study 05: LD-operator validation"
description: "Integrated retained-low-rank and SBayesR LD-sensitivity validation."
---

```{r}
#| echo: false
source("R/benchmark-reporting.R")
capsule <- "results/reference/05_ld_operator/current"
rcsv <- function(x) read.csv(file.path(capsule,x),check.names=FALSE)
manifest <- jsonlite::read_json(file.path(capsule,"benchmark_manifest.json"),simplifyVector=TRUE)
main_manifest <- manifest$components$operator_validation
compatibility <- jsonlite::read_json(file.path(capsule,"compatibility_manifest.json"),simplifyVector=TRUE)
identity <- rcsv("deterministic_identity_summary.csv")
blocks <- rcsv("block_definitions.csv")
rank_design <- rcsv("low_rank_block_diagnostics.csv")
rank_fit <- rcsv("low_rank_fit_block_diagnostics.csv")
checkpoint <- rcsv("checkpoint_validation.csv")
convergence <- rcsv("convergence_validation_summary.csv")
recommendations <- rcsv("method_recommendations.csv")
prediction <- rcsv("prediction_metrics.csv")
parameters <- rcsv("parameter_estimates.csv")
recovery <- rcsv("parameter_recovery_summary.csv")
marker <- rcsv("marker_effect_metrics.csv")
agreement <- rcsv("marker_effect_agreement.csv")
selection <- rcsv("variable_selection_metrics.csv")
paired_rep <- rcsv("paired_replicate_differences.csv")
paired <- rcsv("paired_comparison_summary.csv")
runtime <- rcsv("computational_summary.csv")
sbayesr <- rcsv("sbayesr_diagnostic_evidence.csv")
priors <- rcsv("prior_compatibility.csv")
sources <- rcsv("source_compatibility.csv")
fit_status <- rcsv("fit_status.csv")
sbayesr_design <- rcsv("sbayesr_design.csv")
sbayesr_provenance <- rcsv("sbayesr_provenance.csv")
sbayesr_fit_status <- rcsv("sbayesr_fit_status.csv")
scheduler <- rcsv("sbayesr_scheduler_comparison.csv")
sensitivity_operators <- rcsv("sbayesr_operator_summary.csv")
sensitivity_pairwise <- rcsv("sbayesr_operator_pairwise_comparison.csv")
sensitivity_eigenvalues <- rcsv("sbayesr_eigenvalue_summary.csv")
sensitivity_blocks <- rcsv("sbayesr_block_diagnostics.csv")
sensitivity_variance <- rcsv("sbayesr_variance_summary.csv")
sensitivity_pis <- rcsv("sbayesr_pi_summary.csv")
sensitivity_components <- rcsv("sbayesr_component_summary.csv")
sensitivity_effects <- rcsv("sbayesr_effect_recovery.csv")
sensitivity_prediction <- rcsv("sbayesr_prediction_metrics.csv")
sensitivity_conditionals <- rcsv("sbayesr_conditional_marker_summary.csv")
sensitivity_quadratics <- rcsv("sbayesr_operator_quadratic_checks.csv")
sensitivity_scores <- rcsv("sbayesr_operator_score_checks.csv")
sensitivity_posterior_operator <- rcsv("sbayesr_posterior_operator_comparison.csv")
sensitivity_decomposition <- rcsv("sbayesr_quadratic_error_decomposition.csv")
sensitivity_low_direction <- rcsv("sbayesr_low_eigen_direction_summary.csv")
config_levels <- c("bed","full_csr","block_csr","low_rank_full","low_rank_0999","low_rank_0995")
config_labels <- c(bed="BED",full_csr="Full CSR",block_csr="Block CSR",low_rank_full="Near-full low rank",
 low_rank_0999="Low rank 0.999",low_rank_0995="Low rank 0.995")
config_colours <- c(bed="#0072B2",full_csr="#009E73",block_csr="#56B4E9",low_rank_full="#E69F00",
 low_rank_0999="#D55E00",low_rank_0995="#CC79A7")
config_factor <- function(x) factor(x,levels=config_levels,labels=config_labels[config_levels])
plot_config_metric <- function(data,metric_id,ylab,lower=FALSE,reference=NULL) {
 keep <- data$metric == metric_id
 if ("available" %in% names(data)) keep <- keep & data$available
 z<-data[keep,]
 z$configuration_label<-config_factor(z$configuration); z$architecture_label<-sblrbench_architecture_factor(z$architecture)
 p<-ggplot2::ggplot(z,ggplot2::aes(configuration_label,value,colour=configuration))
 if(!is.null(reference)) p<-p+ggplot2::geom_hline(yintercept=reference,linetype=2,colour="#6B7280")
 p+ggplot2::geom_point(position=ggplot2::position_jitter(width=.08),alpha=.7)+
  ggplot2::stat_summary(fun=mean,geom="point",shape=18,size=3,colour="black")+
  ggplot2::facet_wrap(~architecture_label,scales="free_y")+
  ggplot2::scale_colour_manual(values=config_colours)+theme_sblrbench()+
  ggplot2::labs(x=NULL,y=ylab,caption=paste("Points are five paired replicates; diamonds are means.",if(lower)"Lower is better." else "Higher is better."))+
  ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1),legend.position="none")
}
```

::: {.callout-warning}
This is a five-replicate development benchmark of LD-operator implementations. It does not compare unrelated prior models and does not establish a universal speed, accuracy or storage ranking.
:::

## Benchmark status and scope

All `r main_manifest$successful_fit_count` of `r main_manifest$expected_fit_count` four-chain fits completed across two architectures, five paired replicates and six configurations. The benchmark combines deterministic operator identities, checkpoint/residual validation, convergence support, prediction and recovery metrics, marker-level agreement, runtime/storage measurement, and a current-source compatibility audit.

The question is whether retained low-rank block operators implement their stated projected model correctly and behave stably enough, relative to explicit CSR and BED references, to support a later multivariate extension.

The same study and capsule also contain a focused SBayesR LD-sensitivity
component. It tests exact, hard-sparse, full-rank block-eigen, and retained
0.995 block-eigen operators in one deterministic LD-rich window.

## Configurations

1. **BED**: packed individual-level genotype reference.
2. **Full CSR**: summary-data model with the full sparse LD operator.
3. **Block CSR**: explicit block-diagonal sparse LD; comparison with full CSR measures omitted cross-block LD.
4. **Near-full retained low rank**: retains all numerically positive block eigenmodes.
5. **0.999 retained mass**: smallest per-block rank reaching 0.999 positive eigenvalue mass.
6. **0.995 retained mass**: analogous canonical compression at 0.995.

Reconstructed-dense eigen fits are not active configurations and are not restored as a report.

## Data and simulation design

The two frozen sparse architectures each have `r main_manifest$replicate_count` paired simulations. Every configuration receives the same marker order, phenotype truth and replicate seed. Configuration-specific chain seeds are registered. The capsule records `r nrow(unique(blocks[c("block_id","start","end")]))` genomic blocks and the exact qgdata revision `r main_manifest$qgdata_commit`.

```{r}
#| echo: false
knitr::kable(rcsv("simulation_summary.csv"),digits=4)
```

## Blocks and retained-rank policy

Blocks are contiguous marker intervals frozen in `block_definitions.csv`. Each low-rank block eigendecomposes the matching block-CSR matrix, discards non-positive numerical modes, and chooses the minimum rank meeting the requested positive-eigenvalue mass. Near-full means all positive modes, not an arbitrary dense reconstruction.

```{r}
#| echo: false
rank_summary <- aggregate(cbind(positive_rank,retained_rank,positive_eigenvalue_mass,retained_eigenvalue_mass)~configuration,
 data=rank_fit,FUN=sum)
rank_summary$retained_rank_fraction <- rank_summary$retained_rank/rank_summary$positive_rank
rank_summary$retained_mass_fraction <- rank_summary$retained_eigenvalue_mass/rank_summary$positive_eigenvalue_mass
knitr::kable(rank_summary,digits=6)
```

## Deterministic operator validation

The `block_low_rank_v1` representation (confirmed against `r main_manifest$operator_contract`) stores retained factors $Q_b$, transformed scores $w_b$, and projected residual coordinates $r_b$. Deterministic tests validate construction, diagonal recovery, score transformation, residual updates, corrected marker scores and quadratic forms against explicit source matrices.

```{r}
#| echo: false
identity_cols <- c("configuration","representation","positive_rank","retained_rank","retained_mass_fraction",
 "maximum_score_error","maximum_diagonal_error","maximum_residual_update_error","quadratic_form_error",
 "projected_sse_identity_error","maximum_source_matrix_error","pass")
knitr::kable(identity[,identity_cols],digits=6)
```

## Projected residual/SSE identity

For block scores and residual coordinates, the frozen projected variance identity is

$$
SSE = yy - \sum_b w_b^\top w_b + \sum_b r_b^\top r_b.
$$

The deterministic table reports both residual- and quadratic-form routes and their discrepancy. Passing this identity is required because a low-rank sampler can update marker scores correctly yet still report a wrong residual variance if discarded spectral mass is mishandled.

## Residual-drift validation

Checkpoint validation independently rebuilds residual state and records the maximum drift from incremental updates.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
checkpoint$configuration_label<-config_factor(checkpoint$configuration)
checkpoint$architecture_label<-sblrbench_architecture_factor(checkpoint$architecture)
ggplot2::ggplot(checkpoint,ggplot2::aes(configuration_label,maximum_residual_drift,colour=configuration))+
 ggplot2::geom_point(position=ggplot2::position_jitter(width=.08),alpha=.7)+
 ggplot2::facet_wrap(~architecture_label)+ggplot2::scale_colour_manual(values=config_colours)+theme_sblrbench()+
 ggplot2::scale_y_log10()+ggplot2::labs(x=NULL,y="Maximum residual drift (log scale)",
 caption=paste("Largest observed drift:",format(max(checkpoint$maximum_residual_drift,na.rm=TRUE),scientific=TRUE,digits=4)))+
 ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1),legend.position="none")
```

## Convergence settings and support

Each fit uses four chains with Study 04-informed, method-specific settings. Recommendations and five-replicate support remain visible; convergence is not inferred from successful completion alone.

```{r}
#| echo: false
knitr::kable(recommendations,digits=4)
knitr::kable(convergence,digits=4)
```

## Prediction results

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
plot_config_metric(prediction,"genetic_value_correlation","Genetic-value prediction correlation")
```

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
plot_config_metric(prediction,"phenotype_prediction_correlation","Phenotype prediction correlation")
```

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
plot_config_metric(prediction,"genetic_value_rmse","Genetic-value prediction RMSE",TRUE)
```

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
plot_config_metric(prediction,"prediction_regression_slope","Prediction regression slope",FALSE,1)
```

## Parameter recovery

```{r}
#| echo: false
knitr::kable(recovery,digits=4)
```

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
z<-parameters[parameters$available & parameters$estimand %in% c("heritability","genetic_variance","residual_variance"),]
z$configuration_label<-config_factor(z$configuration); z$architecture_label<-sblrbench_architecture_factor(z$architecture)
ggplot2::ggplot(z,ggplot2::aes(truth,posterior_mean,colour=configuration))+
 ggplot2::geom_abline(slope=1,intercept=0,linetype=2,colour="#6B7280")+ggplot2::geom_point(alpha=.7)+
 ggplot2::facet_grid(estimand~architecture_label,scales="free")+ggplot2::scale_colour_manual(values=config_colours,labels=config_labels)+
 theme_sblrbench()+ggplot2::labs(x="Truth",y="Posterior mean",colour="Configuration",
 caption="Points are replicate-level estimates; the dashed line is equality.")
```

## Marker-effect agreement

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
plot_config_metric(marker,"marker_effect_rmse","Marker-effect RMSE",TRUE)
```

```{r}
#| echo: false
agreement_summary<-aggregate(cbind(posterior_mean_effect_correlation,posterior_mean_effect_rmse,
 nonnull_probability_correlation,high_pip_overlap)~architecture+comparison_id+orientation,data=agreement,FUN=mean,na.rm=TRUE)
knitr::kable(agreement_summary,digits=4)
```

Agreement is reported for explicit oriented comparisons. High correlation can coexist with scale bias; RMSE and high-PIP overlap therefore remain alongside correlation.

## Variable-selection metrics

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
plot_config_metric(selection,"pip_auprc","PIP area under precision-recall curve")
```

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
plot_config_metric(selection,"pip_causal_mean","Mean causal-marker PIP")
```

Availability flags are respected; unsupported outputs remain recorded in the capsule rather than silently removed.

## Paired operator contrasts

The contrasts have different interpretation levels: block CSR minus full CSR measures block approximation; near-full low rank minus block CSR isolates retained-factor implementation most closely; 0.999/0.995 versus near-full measures spectral truncation; full CSR minus BED is the broader summary-data versus individual-data workflow difference.

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
z<-paired[paired$metric %in% c("genetic_value_correlation","genetic_value_rmse","heritability",
 "marker_effect_rmse","pip_auprc"),]
ggplot2::ggplot(z,ggplot2::aes(metric,mean_difference,colour=comparison_id))+
 ggplot2::geom_hline(yintercept=0,linetype=2)+ggplot2::geom_pointrange(
  ggplot2::aes(ymin=minimum_difference,ymax=maximum_difference),position=ggplot2::position_dodge(width=.5))+
 ggplot2::facet_wrap(~architecture,scales="free_y")+ggplot2::coord_flip()+theme_sblrbench()+
 ggplot2::labs(x=NULL,y="Mean focal-minus-reference difference (observed range)",colour="Comparison")
```

## Rank and spectral-retention diagnostics

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
rs<-rank_summary[rank_summary$configuration %in% c("low_rank_full","low_rank_0999","low_rank_0995"),]
rs$configuration_label<-config_factor(rs$configuration)
rl<-rbind(transform(rs,diagnostic="Retained rank fraction",value=retained_rank_fraction),
 transform(rs,diagnostic="Retained positive spectral mass",value=retained_mass_fraction))
ggplot2::ggplot(rl,ggplot2::aes(configuration_label,value,fill=configuration))+
 ggplot2::geom_col()+ggplot2::facet_wrap(~diagnostic,scales="free_y")+
 ggplot2::scale_fill_manual(values=config_colours)+theme_sblrbench()+
 ggplot2::labs(x=NULL,y="Fraction",caption="Fractions are summed over frozen blocks; spectral mass and rank are not interchangeable.")+
 ggplot2::theme(axis.text.x=ggplot2::element_text(angle=25,hjust=1),legend.position="none")
```

## Runtime and storage

Retained rank alone does not guarantee speed or memory savings. Immutable operator bytes, per-chain state, fit/checkpoint size, construction time and MCMC time are measured separately.

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
runtime$configuration_label<-config_factor(runtime$configuration);runtime$architecture_label<-sblrbench_architecture_factor(runtime$architecture)
rl<-rbind(transform(runtime,resource="MCMC runtime (s)",value=mcmc_runtime_seconds),
 transform(runtime,resource="Immutable operator (bytes)",value=immutable_operator_bytes),
 transform(runtime,resource="Per-chain state (bytes)",value=per_chain_state_bytes),
 transform(runtime,resource="Fit object (bytes)",value=fit_object_bytes))
ggplot2::ggplot(rl,ggplot2::aes(configuration_label,value,colour=configuration))+
 ggplot2::geom_point(position=ggplot2::position_jitter(width=.08),alpha=.65)+
 ggplot2::stat_summary(fun=mean,geom="point",shape=18,size=2.8,colour="black")+
 ggplot2::facet_grid(resource~architecture_label,scales="free_y")+ggplot2::scale_y_log10()+
 ggplot2::scale_colour_manual(values=config_colours)+theme_sblrbench()+
 ggplot2::labs(x=NULL,y="Measured value (log scale)",caption="Points are fits; diamonds are means.")+
 ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1),legend.position="none")
```

## SBayesR diagnostic evidence

The capsule preserves package-reported variance components, benchmark heritability, marker-effect variance, operator diagonals and log-CPO summaries for SBayesR configurations. This guards against an apparently stable operator comparison masking a model-output pathology.

```{r}
#| echo: false
sbayes_vars <- c("vgs_posterior_mean","ves_posterior_mean","benchmark_heritability",
 "package_reported_heritability","marker_effect_variance","operator_diagonal_mean","mean_log_cpo")
sbayes_summary<-aggregate(sbayesr[sbayes_vars],sbayesr[c("architecture","configuration")],
 FUN=function(x) if(all(is.na(x))) NA_real_ else mean(x,na.rm=TRUE))
knitr::kable(sbayes_summary,digits=5)
```

## SBayesR sensitivity to LD approximation

The integrated SBayesR component asks how exact, element-wise hard-sparse,
full-rank block-eigen, and retained-low-rank LD operators affect marker
conditionals, posterior variance estimates, component allocation, and
recovery. It is a focused source-linked validation, not a genome-wide LD-policy
selection exercise.

All scheduler, BED, exact-CSR, hard-sparse CSR, full-rank block-eigen, and
retained-low-rank fits were reused from the completed diagnostic checkpoints
when the compact evidence was originally frozen. This integration did not call
a sampler.

```{r}
#| echo: false
knitr::kable(sbayesr_provenance)
knitr::kable(sbayesr_fit_status[, c("label", "elapsed_seconds", "fit_origin",
  "warnings")], digits = 3)
```

### BED scheduler comparison

Scheduled and full-sweep BED BayesR were nearly identical on the matched
sparse-mixture coordinate, while current CSR SBayesR remained different. The
adaptive BED scheduler is therefore not the primary explanation.

```{r}
#| echo: false
scheduler_wide <- reshape(scheduler[, c("label", "quantity", "mean")],
  idvar = "label", timevar = "quantity", direction = "wide")
names(scheduler_wide) <- sub("mean\\.", "", names(scheduler_wide))
knitr::kable(scheduler_wide, digits = 5)
```

### Reduced LD-rich window

The deterministic 1,500-marker window contains 5,000 samples and 15 causal
markers. Its reduced phenotype was simulated at realized heritability 0.30
with seed 17,002, so omitted signal outside the window cannot confound the
operator identities.

```{r}
#| echo: false
knitr::kable(sbayesr_design[, c("qc_start", "qc_end", "first_marker",
  "last_marker", "samples", "markers", "causal_markers", "realized_h2")])
```

### Exact CSR versus BED BayesR

The full-pair CSR correlation representation, rescaled by marker-specific BED
cross-product diagonals, reproduced direct \(X'X\) to numerical tolerance.
Exact CSR closely reproduced full-sweep BED in posterior variance, component
allocation, marker effects, and genetic-value recovery. The deterministic
conditional audit identified no defect in the published/current BayesR
conditional formulas.

```{r}
#| echo: false
central <- sensitivity_variance[sensitivity_variance$quantity %in%
  c("heritability", "vgs", "ves", "vbs"), ]
central_wide <- reshape(central[, c("label", "quantity", "mean")],
  idvar = "label", timevar = "quantity", direction = "wide")
names(central_wide) <- sub("mean\\.", "", names(central_wide))
knitr::kable(central_wide, digits = 5)
```

### Hard-sparse CSR behavior

Hard sparsification moved genetic variance and heritability upward, residual
variance downward, and assigned more markers moderate posterior inclusion
probability. Its relative Frobenius error was 0.4422 despite an exactly
preserved diagonal.

### Block-eigen and retained-low-rank SBayesR

The full-rank representation reproduces its configured block-diagonal target,
but not global exact LD because cross-block entries are omitted. The retained
representation uses cumulative positive mass 0.995 and retained 1,490 of 1,500
directions. Five 250-marker blocks retained 248 directions each; the two
chromosome-boundary fragments retained full rank.

```{r}
#| echo: false
knitr::kable(sensitivity_pairwise, digits = 5)
knitr::kable(sensitivity_blocks[, c("block_id", "start", "end",
  "positive_rank", "retained_rank", "retained_mass_fraction")], digits = 5)
```

### Corrected-score audit

Direct/BED and exact-CSR corrected scores, component probabilities, and
conditional effect means agree to numerical precision. Hard-sparse score and
component-probability differences are material for the fixed marker panel.

```{r}
#| echo: false
knitr::kable(sensitivity_conditionals, digits = 7)
```

### Quadratic and residual audit

For fixed effect states, direct genotype and exact-CSR quadratic and residual
identities agree. Hard-sparse and block-diagonal operators understate the
quadratic form and corresponding residual expression because omitted LD terms
are absent.

```{r}
#| echo: false
knitr::kable(sensitivity_quadratics[
  sensitivity_quadratics$effect_state %in% c("truth", "csr_sparse_ld",
    "block_eigen_full_rank", "block_eigen_retained"), ], digits = 4)
knitr::kable(sensitivity_posterior_operator, digits = 5)
```

### Spectral findings

Neither exact nor hard-sparse LD has eigenvalues below \(-10^{-8}\) or within
\(\pm10^{-8}\). Hard sparsification is positive definite in this window. The
evidence favors accumulated omitted weak LD rather than a small set of
indefinite or near-zero directions.

```{r}
#| echo: false
knitr::kable(sensitivity_eigenvalues[, c("operator", "minimum", "maximum",
  "below_negative_tolerance", "near_zero", "effective_rank",
  "positive_condition_number")], digits = 4)
knitr::kable(sensitivity_decomposition, digits = 4)
```

### SBayesR recovery

Exact CSR and BED provide the strongest recovery. Retained low rank improves
modestly over full-rank blocking and hard sparsification, but the difference is
small in this single replicate.

```{r}
#| echo: false
knitr::kable(sensitivity_effects, digits = 6)
knitr::kable(sensitivity_prediction, digits = 6)
```

### Integrated interpretation and validation boundaries

LD-operator representation affects both computational efficiency and
statistical behavior. Block and retained-low-rank representations provide
compact approximations, but hard sparsification and block boundaries can omit
weak or cross-block LD. In the tested SBayesR window, exact CSR reproduced
full-sweep BED BayesR, while hard-sparse and block-diagonal operators altered
corrected marker scores, component probabilities, residual expressions, and
posterior recovery. The hard-sparse matrix was positive definite, and retaining
99.5% of within-block spectral mass gave modest improvement but did not fully
recover omitted LD.

This one LD-rich reduced window does not prove the complete cause of every
whole-genome result. It establishes no universal LD window or threshold,
does not prove retained low rank universally superior, and does not by itself
recommend changing `sblr` defaults. Operator choice remains data dependent.

## Current-source compatibility audit

The 60 numerical fits truthfully retain source SHA `r main_manifest$numerical_fit_source_sha`. The refresh validation used `r main_manifest$refresh_validation_sha` without sampler reruns. The audit established identical resolved BayesC/BayesR priors and byte-identical scalar sampler-transition and retained-low-rank sources, yielding status `r main_manifest$compatibility_status`.

```{r}
#| echo: false
compat_table<-data.frame(Check=c("Numerical provenance preserved","Sampler transition sources identical",
 "Retained-low-rank sources identical","Resolved priors identical","Rerun required"),
 Value=c(compatibility$numerical_fit_provenance_preserved,compatibility$sampler_transition_sources_identical,
 compatibility$retained_low_rank_sources_identical,compatibility$resolved_priors_identical,compatibility$rerun_required))
knitr::kable(compat_table)
knitr::kable(priors[,c("model","configuration","identical_resolved_prior")])
knitr::kable(sources)
```

## Readiness decision for MTBLR

The frozen decision is **READY FOR MTBLR LOW-RANK EXTENSION** (confirmed against `r main_manifest$readiness_decision`). Deterministic identities and projected SSE pass, residual drift is characterized, all fits and convergence checks meet the frozen support contract, retention sensitivity is not pathological in the tested grid, and runtime/storage costs are measured. This is permission to begin a separate multivariate implementation effort—not evidence that an MT sampler already exists or is scientifically validated.

## Interpretation

Block-CSR differences from full CSR quantify the cost of omitting cross-block LD. Near-full low rank versus block CSR is the closest implementation comparison; 0.999 and 0.995 contrasts add spectral approximation. BED versus CSR is intentionally broader. Scientific metrics, numerical identities and resource measurements must agree before compression can be judged useful.

## Limitations

Only two sparse architectures, five replicates, one genotype/LD panel, one block scheme and two compressed mass targets were tested. Observed ranges are not inferential uncertainty. Near-full positive rank can approach the original dimension, and decomposition cost may dominate. Scalar compatibility does not validate annotation or multivariate transitions; those are explicitly outside scope.

## Reproducibility and downloads

Rendering reads the current capsule only and performs no sampler or checkpoint work. The [analysis entry point](analysis.R) documents the deterministic/current workflow separately.

- [Study specification](spec.R)
- [Operator design](operator-design.R)
- [Reusable operator-analysis template](../../inst/templates/operator-analysis.R)
- [Integrated capsule README](../../results/reference/05_ld_operator/current/README.md)
- [Integrated manifest](../../results/reference/05_ld_operator/current/benchmark_manifest.json)
- [Checksums](../../results/reference/05_ld_operator/current/checksums.csv)
- [Compatibility manifest](../../results/reference/05_ld_operator/current/compatibility_manifest.json)
- [Deterministic identities](../../results/reference/05_ld_operator/current/deterministic_identity_summary.csv)
- [Prediction metrics](../../results/reference/05_ld_operator/current/prediction_metrics.csv)
- [Computational summary](../../results/reference/05_ld_operator/current/computational_summary.csv)
- [SBayesR scheduler comparison](../../results/reference/05_ld_operator/current/sbayesr_scheduler_comparison.csv)
- [SBayesR operator summary](../../results/reference/05_ld_operator/current/sbayesr_operator_summary.csv)
- [SBayesR variance summary](../../results/reference/05_ld_operator/current/sbayesr_variance_summary.csv)
- [SBayesR corrected-score audit](../../results/reference/05_ld_operator/current/sbayesr_operator_score_checks.csv)
- [SBayesR quadratic audit](../../results/reference/05_ld_operator/current/sbayesr_operator_quadratic_checks.csv)
- [`sblr` CSR documentation](https://github.com/psoerensen/sblr/blob/02e8c74baa906e83c4a08d42a9cc6339b4e81072/man/stblr_csr.Rd)
